---
title: "Comparing common and rare single variant and gene aggregate instrumentation strategies for molecular MR"
author: "Aimee Hanson"
date: "`r format(Sys.time(), '%d %B %Y')`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 4
    toc-location: left
    page-layout: full
    code-fold: true
---

```{r}
#| label: setup
#| message: FALSE
#| echo: FALSE

library(here)
library(dplyr)
library(ggplot2)
library(dotenv)
library(UpSetR)
library(grid)
library(gridExtra)
library(Gviz)
library(purrr)
library(TwoSampleMR)

here::i_am(path = "scripts/04_analysis_proteins/rarevcommon_molecularMR_v2.qmd")

dotenv::load_dot_env(file = here("config.env"))
data_dir <- Sys.getenv("data_dir")
res_dir <- Sys.getenv("results_dir")

knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center")
                      #dev = "pdf",
                      #fig.path = "figures_pdf/")

source(here("scripts/04_analysis_proteins/functions.R"))

#install.packages("TwoSampleMR", repos = c("https://mrcieu.r-universe.dev", "https://cloud.r-project.org"))
```

## Introduction

Classically, Mendelian Randomisation (MR) methods utilising trait-associated genetic variants from GWAS studies have employed common polymorphisms (population MAF \> 1%) targeted by genotyping chips, or reliably imputed from reference populations, to instrument modifiable exposures. However, common variants tested in GWAS typically explain a very small fraction of the variability in a measured trait, potentially exhibit pleiotropic effects acted upon by balancing selection or as a consequent of genetic linkage, and are rarely causal. Rare variants, which typically show large biological effects (e.g. through abolishing protein expression) provide a means of more unambiguously instrumenting relevant molecular processes. Comparison of causal estimates derived using differing methods of genetically instrumenting modifiable exposures may enhance the interpretation of the biological mechanisms underlying exposure-outcome relationships. This includes using variants from across the allele frequency spectrum, but also leveraging rare variant aggregate approaches to instrument gene-level perturbations in expression and function.

Here, caual estimates for pairwises combinations of **2937** molecular exposures (O-link plasma proteins) and **11** health related compelex traits (below) have been derived using nine instrumentation strategies, with further instrument splitting into cis (variants within 1MB of the encoding gene start and end site), trans and full variant sets (a total of **27** instrument sets per exposure).

**Outcomes:** Low density lipoprotein levels (LDL), Body Mass Index (BMI), Vitamin D (VitD), Triglycerides (Trig), Glycated Haemoglobin (HbA1c), Mean Platelet Volume (PlatVol), IGF-1 (IGF1), Waist-to-Hip Ratio (WHRadjBMI), Red Blood Cell Erythrocyte Count (RBCCount), Mean Corpuscular Volume (CorpVol) and Systolic Blood Pressure (SBP)

### Instrumental variable (IV) sets

Nine sets of IVs for each protein exposure have been extracted from across two studies:\
GWAS common variants: "Plasma proteomic associations with genetics and health in the UK Biobank" (Sun, B. 2023, Nature).\
WES variants/gene aggregate tests: "Rare variant associations with plasma protein levels in the UK Biobank" (Dhindsa, R. 2023, Nature).\

**Common GWAS**\
Common genome-wide fine-mapped variants from the GPMAP (\>1% MAF) (cis, trans and combined)\
Common genome-wide clumped and pruned variants (\>1% MAF) (cis, trans and combined)

**WES (variants)**\
Common exome-wide (\>1% MAF) (cis, trans and combined)\
Rare exome-wide (0-1% MAF) (cis, trans and combined)\
Ultra-rare exome-wide (0-0.1% MAF) (cis, trans and combined)\
*LD clumping was performed for common variants (for which LD matrices have been derived) only. For rare variants, instrument sets were stored both pre and post pruning to variants independent at r2\<0.1.*

**WES (masks)**\
pLOF burden mask (cis, trans and combined)\
missense/low-confidence burden mask (cis, trans and combined)\
pLOF/missense/low-confidence burden mask (cis, trans and combined)\
synonymous burden mask (cis, trans and combined)

## Data import

For direct comparison of the performance of instrument sets, the theshold for pQTL significance has been set at 5x10\^-8 for both rare and common variants. MR results for all exposure-outcome pairs, derived using differing IV sets, and the harmonised summary statistics of those variant sets, are read in below.

Read in MR results from `scripts/04_analysis_proteins/001_performmolecularMR.R` and harmonised summary statistics:

```{r}
#| label: import-mr-results

# ============== Import MR results (final set) ============= #
# # Origininal results (rare IVs P<5e-8, common IVs P<5e-8)
# files <- list.files(res_dir, pattern = "results_molecular_MR_GW5e-08_EW5e-08_[A-Z]", full.names = T)
# files_snps <- list.files(file.path(data_dir, "harmonised"), pattern = "harmonised_studies_GW5e-08_EW5e-08_[A-Z]", full.names = T)
# 
# # Repeat with rare variant LD pruning (rare IVs P<5e-8, common IVs P<5e-8)
# files_pruned <- list.files(res_dir, pattern = "results_molecular_MR_GW5e-08_EW5e-08_rarepruned", full.names = T)
# files_pruned <- files_pruned[-grep("COMBINED", files_pruned)]
# files_snps_pruned <- list.files(file.path(data_dir, "harmonised"), pattern = "harmonised_studies_GW5e-08_EW5e-08_rarepruned", full.names = T)
# files_snps_pruned <- files_snps_pruned[-grep("COMBINED", files_snps_pruned)]
# 
# res_MR <- list()
# for(i in 1:length(files)){
#   load(files[i])
#   res_MR[[i]] <- results_MR
# }
# names(res_MR) <- sub(".*08_(.*)\\.rda","\\1", files)
# 
# ### Append results from pruned rare IVs
# res_MR_pruned <- list()
# for(i in 1:length(files_pruned)){
#   load(files_pruned[i])
#   res_MR_pruned[[i]] <- lapply(results_MR, function(x){x$IV_set = paste0(x$IV_set, "_pruned"); return(x)})
# }
# names(res_MR_pruned) <- sub(".*rarepruned_(.*)\\.rda","\\1", files_pruned)
# 
# res_MR_joined <- Map(function(x, y) {
#   Map(rbind, x, y)
# }, res_MR, res_MR_pruned)
# 
# rm(res_MR_pruned)

# ============== Import MR results (relaxed set) ============= #
# Less strict thresholding for rare variants w LD pruning (rare IVs P<1e-4, common IVs P<5e-8)
# files_relaxed_pruned <- list.files(res_dir, pattern = "results_molecular_MR_rarepruned", full.names = T)
# files_snps_relaxed_pruned <- list.files(file.path(data_dir, "harmonised"), pattern = "harmonised_studies_rarepruned", full.names = T)
# 
# ### Append results from pruned rare IVs
# res_MR_relaxed_pruned <- list()
# for(i in 1:length(files_relaxed_pruned)){
#   load(files_relaxed_pruned[i])
#   res_MR_relaxed_pruned[[i]] <- lapply(results_MR, function(x){x$IV_set = paste0(x$IV_set, "_pruned"); return(x)})
# }
# names(res_MR_relaxed_pruned) <- sub(".*_(.*)\\.rda","\\1", files_relaxed_pruned)
# 
# res_MR_relaxed <- Map(function(x, y) {
#   Map(rbind, x, y)
# }, res_MR, res_MR_relaxed_pruned)
# 
# rm(results_MR)
# rm(res_MR_relaxed_pruned)
# res_MR <- res_MR_joined
# rm(res_MR_joined)

# ========== Import harmonised SNP data ========= #
# harmonised_snps <- list()
# for(i in 1:length(files_snps)){
#   load(files_snps[i])
#   harmonised_snps[[i]] <- harmonised_studies
# }
# names(harmonised_snps) <- sub(".*08_(.*)\\.rda","\\1", files_snps)
# 
# ### Append pruned rare IVs
# harmonised_snps_pruned <- list()
# for(i in 1:length(files_snps_pruned)){
#   load(files_snps_pruned[i])
#   harmonised_snps_pruned[[i]] <- lapply(harmonised_studies_pruned, function(x){names(x) <- paste0(names(x), "_pruned"); return(x)})
# }
# names(harmonised_snps_pruned) <- sub(".*rarepruned_(.*)\\.rda","\\1", files_snps_pruned)
# 
# # Less strict thresholding for rare variants w LD pruning (rare IVs P<1e-4, common IVs P<5e-8)
# harmonised_snps_relaxed <- list()
# for(i in 1:length(files_snps_relaxed_pruned)){
#   load(files_snps_relaxed_pruned[i])
#   harmonised_snps_relaxed[[i]] <- lapply(harmonised_studies_pruned, function(x){names(x) <- paste0(names(x), "_pruned"); return(x)})
# }
# names(harmonised_snps_relaxed) <- sub(".*pruned_(.*)\\.rda","\\1", files_snps_relaxed_pruned)
# 
# harmonised_snps_relaxed <- Map(function(a1, b1) {
#   Map(function(a2, b2) {
#     c(a2, b2)   # append entire gene-level lists
#   }, a1, b1)
# }, harmonised_snps, harmonised_snps_relaxed) # Rare variants pruned with relaxed threshold
# 
# harmonised_snps <- Map(function(a1, b1) {
#   Map(function(a2, b2) {
#     c(a2, b2)   # append entire gene-level lists
#   }, a1, b1)
# }, harmonised_snps, harmonised_snps_pruned) # Rare variants pruned
# 
# rm(harmonised_studies)
# rm(harmonised_snps_pruned)
# rm(harmonised_studies_pruned)

# --------------------------------------------------------------------------------------- #
# List of LD pruned rare variants:
# P<5e-8
prunein_files <- grep("prune.in", list.files(file.path(data_dir,"sumstats/proteomics/protein_pQTL_pruning"), pattern = "p5e-8_.*", full = TRUE), value = T)
prunein <- do.call("rbind", lapply(prunein_files, function(x){data.table::fread(x, header = F, data.table = F)}))[,1]
# Correct variant name formatting
prunein <- unlist(lapply(strsplit(prunein, ":"), function(x){paste0(x[1],":",x[2],"_",tolower(x[3]),"_",tolower(x[4]))}))

# P<1e-4
prunein_files_relaxed <- grep("prune.in", list.files(file.path(data_dir,"sumstats/proteomics/protein_pQTL_pruning"), pattern = "wes_pQTLs_chr",, full = TRUE), value = T)
prunein_relaxed <- do.call("rbind", lapply(prunein_files_relaxed, function(x){data.table::fread(x, header = F, data.table = F)}))[,1]
# Correct variant name formatting
prunein_relaxed <- unlist(lapply(strsplit(prunein_relaxed, ":"), function(x){paste0(x[1],":",x[2],"_",tolower(x[3]),"_",tolower(x[4]))}))

# --------------------------------------------------------------------------------------- #
# List of LD pruned common variants (filtered from all IVs combined; P<5e-8)
prunein_exwas_files <- grep("prune.in", list.files(file.path(data_dir,"sumstats/proteomics/protein_pQTL_pruning"), pattern = "common", full = TRUE), value = T)
prunein_exwas <- do.call("rbind", lapply(prunein_exwas_files, function(x){data.table::fread(x, header = F, data.table = F)}))[,1]

prunein_gwas_files <- grep("prune.in", list.files(file.path(data_dir,"sumstats/proteomics/protein_pQTL_pruning"), pattern = "gwas", full = TRUE), value = T)
prunein_gwas <- do.call("rbind", lapply(prunein_gwas_files, function(x){data.table::fread(x, header = F, data.table = F)}))[,1]

# --------------------------------------------------------------------------------------- #
# Data filtering
## Retain the following instrument sets for final analysis:
IVsets <- c("dhindsa_exwas_mask_ptv" = "mask:pLOF",
            "dhindsa_exwas_mask_raredmg" = "mask:miss",
            "dhindsa_exwas_mask_ptvraredmg" = "mask:pLOF|miss",
            "dhindsa_exwas_mask_syn" = "mask:synonymous",
            "dhindsa_exwas_variant_ultrarare_pruned" = "exome:ultrarare",
            "dhindsa_exwas_variant_rare_pruned" = "exome:rare",
            "dhindsa_exwas_variant_common" = "exome:common",
            "sun_gwas_variant_common_clumped" = "genome:common")

variant_IVsets <- c("dhindsa_exwas_variant_ultrarare_pruned" = "exome:ultrarare",
                    "dhindsa_exwas_variant_rare_pruned" = "exome:rare",
                    "dhindsa_exwas_variant_common" = "exome:common",
                    "sun_gwas_variant_common_clumped" = "genome:common")

unpruned_IVsets <- c("dhindsa_exwas_mask_ptv" = "mask:pLOF",
                    "dhindsa_exwas_mask_raredmg" = "mask:miss",
                    "dhindsa_exwas_mask_ptvraredmg" = "mask:pLOF|miss",
                    "dhindsa_exwas_mask_syn" = "mask:synonymous",
                    "dhindsa_exwas_variant_ultrarare" = "exome:ultrarare",
                    "dhindsa_exwas_variant_rare" = "exome:rare",
                    "dhindsa_exwas_variant_common" = "exome:common",
                    "sun_gwas_variant_common_clumped" = "genome:common")

# Varaint and gene mask comparison for manuscript
paper_IVsets <- c(#"dhindsa_exwas_mask_ptv" = "mask:pLOF",
                  #"dhindsa_exwas_mask_raredmg" = "mask:miss",
                  "dhindsa_exwas_mask_ptvraredmg" = "mask:pLOF|miss",
                  "dhindsa_exwas_variant_ultrarare_pruned" = "exome:ultrarare",
                  "dhindsa_exwas_variant_rare_pruned" = "exome:rare",
                  "dhindsa_exwas_variant_common" = "exome:common",
                  "sun_gwas_variant_common_clumped" = "genome:common")

IV_sets <- names(IVsets)

# res_MR <- lapply(res_MR, function(x){
#   lapply(x, function(y){
#     y |> dplyr::filter(IV_set %in% names(IVsets))
#     })
#   })
# 
# res_MR_relaxed <- lapply(res_MR_relaxed, function(x){
#   lapply(x, function(y){
#     y |> dplyr::filter(IV_set %in% names(IVsets))
#     })
#   })
# 
# harmonised_snps <- lapply(harmonised_snps, function(x){lapply(x, function(y){y[names(IVsets)]})})
# harmonised_snps_relaxed <- lapply(harmonised_snps_relaxed, function(x){lapply(x, function(y){y[names(IVsets)]})})

# Save
#save(res_MR, file = file.path(res_dir, "results_molecular_MR_GW5e-08_EW5e-08_rarepruned_COMBINED.rda"))
#save(res_MR_relaxed, file = file.path(res_dir, "results_molecular_MR_GW5e-08_EW1e-04_rarepruned_COMBINED.rda"))
#save(harmonised_snps, file = file.path(data_dir, "harmonised/harmonised_studies_GW5e-08_EW5e-08_rarepruned_COMBINED.rda"))
#save(harmonised_snps_relaxed, file = file.path(data_dir, "harmonised/harmonised_studies_GW5e-08_EW1e-04_rarepruned_COMBINED.rda"))

load(file = file.path(res_dir, "results_molecular_MR_GW5e-08_EW5e-08_rarepruned_COMBINED.rda")) # res_MR
load(file = file.path(res_dir, "results_molecular_MR_GW5e-08_EW1e-04_rarepruned_COMBINED.rda")) # res_MR_relaxed
load(file = file.path(data_dir, "harmonised/harmonised_studies_GW5e-08_EW5e-08_rarepruned_COMBINED.rda")) # harmonised_snps
load(file = file.path(data_dir, "harmonised/harmonised_studies_GW5e-08_EW1e-04_rarepruned_COMBINED.rda")) # harmonised_snps_relaxed

# --------------------------------------------------------------------------------------- #

# List of the cis-IVs used for MR (with palendromic SNPs removed)
# all_cisIVs <- harmonised_snps |>
#   # Iterate at the innermost level so the name is exactly the IV set name
#   map(~ map(.x, ~ imap(.x, \(z, ivset) {
#     if ("variant" %in% colnames(z)) {
#       z$SNP_ID <- toupper(z$variant)
#     } else {
#       z$SNP_ID <- toupper(z$SNP)
#     }
#     z[, c("SNP_ID", "exposure", "effect_allele.exposure", "other_allele.exposure", "cis_trans", "mr_keep")] |>
#       mutate(IVset = ivset)
#   }))) |>
#   list_flatten() |>
#   list_flatten() |>
#   list_rbind() |>
#   distinct() |>                              # deduplicate identical rows across traits before filtering
#   na.exclude() |>
#   filter(cis_trans == "cis", mr_keep == TRUE) |>
#   group_by(IVset, SNP_ID, effect_allele.exposure, other_allele.exposure) |>
#   summarise(n_cis_assocs = n(), .groups = "drop") |>
#   mutate(
#     IV_set   = IVsets[IVset],
#     IV_group = ifelse(grepl("rare",   IV_set), "variant:rare",
#                ifelse(grepl("common", IV_set), "variant:common", IV_set)),
#     effect_allele.exposure = ifelse(grepl("mask", IV_set), "P", effect_allele.exposure), # Present/absent for masks
#     other_allele.exposure  = ifelse(grepl("mask", IV_set), "A", other_allele.exposure)
#   )
# 
# all_cisIVs_relaxed <- harmonised_snps_relaxed |>
#   # Iterate at the innermost level so the name is exactly the IV set name
#   map(~ map(.x, ~ imap(.x, \(z, ivset) {
#     if ("variant" %in% colnames(z)) {
#       z$SNP_ID <- toupper(z$variant)
#     } else {
#       z$SNP_ID <- toupper(z$SNP)
#     }
#     z[, c("SNP_ID", "exposure", "effect_allele.exposure", "other_allele.exposure", "cis_trans", "mr_keep")] |>
#       mutate(IVset = ivset)
#   }))) |>
#   list_flatten() |>
#   list_flatten() |>
#   list_rbind() |>
#   distinct() |>                              # deduplicate identical rows across traits before filtering
#   na.exclude() |>
#   filter(cis_trans == "cis", mr_keep == TRUE) |>
#   group_by(IVset, SNP_ID, effect_allele.exposure, other_allele.exposure) |>
#   summarise(n_cis_assocs = n(), .groups = "drop") |>
#   mutate(
#     IV_set   = IVsets[IVset],
#     IV_group = ifelse(grepl("rare",   IV_set), "variant:rare",
#                ifelse(grepl("common", IV_set), "variant:common", IV_set)),
#     effect_allele.exposure = ifelse(grepl("mask", IV_set), "P", effect_allele.exposure),
#     other_allele.exposure  = ifelse(grepl("mask", IV_set), "A", other_allele.exposure)
#   )

# --------------------------------------------------------------------------------------- #
# Full SNP details for exome varaints
wes_deets <- data.table::fread(file.path(data_dir, "sumstats/proteomics/proteins_dhindsa_pQTL_annotations.tsv"))
# Vep annotations for common GWAS variants
gwas_deets <- data.table::fread(file.path(data_dir, "variant_annotation/vep_variantannotations_hg38_formatted.txt"))
# Remove gene designation for any variant that is not situated in a gene
gwas_deets$GENE = ifelse(gwas_deets$ANNOTATION %in% c("intergenic_variant", "downstream_gene_variant", "upstream_gene_variant"), "", gwas_deets$GENE)

# Format variant annotations and append to cis_IV details
# cis_IVdeets <- wes_deets |> dplyr::filter(Variant %in% gsub("_","-",sub(":","-",all_cisIVs_relaxed |> dplyr::filter(grepl("exome", IV_set)) |> pull(SNP_ID)))) |>
#   dplyr::mutate(GENE = gsub("\\'","",GENE)) |>
#   rbind(gwas_deets |> dplyr::filter(Variant %in% gsub("_","-",sub(":","-",all_cisIVs_relaxed |> dplyr::filter(IV_set == "genome:common") |> pull(SNP_ID)))), fill = T)

IV_consequence = c(
  "3_prime_UTR_variant" = "UTR_variant",
  "5_prime_UTR_variant" = "UTR_variant",
  "conservative_inframe_deletion" = "inframe_ins_del",
  "conservative_inframe_insertion" = "inframe_ins_del",
  "disruptive_inframe_deletion" = "inframe_ins_del",
  "disruptive_inframe_insertion" = "inframe_ins_del",
  "downstream_gene_variant" = "intergenic",
  "intergenic_variant" = "intergenic",
  "upstream_gene_variant" = "intergenic",
  "frameshift_variant" = "frameshift",
  "initiator_codon_variant" = "start_codon_variant",
  "start_lost" = "start_codon_variant",
  "stop_gained" = "truncating",
  "stop_lost" = "stop_lost",
  "intron_variant" = "intronic",
  "missense_variant" = "missense",
  "non_coding_transcript_exon_variant" = "noncoding",
  "regulatory_region_variant" = "regulatory",
  "splice_acceptor_variant" = "splice_variant",
  "splice_donor_5th_base_variant" = "splice_variant",
  "splice_donor_region_variant" = "splice_variant",
  "splice_donor_variant" = "splice_variant",
  "splice_polypyrimidine_tract_variant" = "splice_variant",
  "splice_region_variant" = "splice_variant",
  "missense_variant&splice_region_variant" = "splice_variant",
  "synonymous_variant&splice_region_variant" = "splice_variant",
  "stop_retained_variant" = "synonymous",
  "synonymous_variant" = "synonymous")

cols_consequences <- colorRampPalette(RColorBrewer::brewer.pal(n= 12,"Spectral"))(length(unique(IV_consequence)))
names(cols_consequences) <- unique(IV_consequence)[order(unique(IV_consequence))]

# cis_IVdeets$Variant_group = IV_consequence[cis_IVdeets$ANNOTATION]

# all_cisIVs <- all_cisIVs |> 
#   dplyr::mutate(Variant = gsub("_","-",sub(":","-",SNP_ID))) |>
#   left_join(cis_IVdeets, by = "Variant") |>
#   dplyr::rename("effect_allele.vep" = EA,
#                 "other_allele.vep" = OA) |>
#   dplyr::filter(duplicated(paste(IV_set, Variant)) == FALSE) |>
#   dplyr::mutate(ea_matches_annot = effect_allele.exposure == effect_allele.vep)
# 
# all_cisIVs_relaxed <- all_cisIVs_relaxed |> 
#   dplyr::mutate(Variant = gsub("_","-",sub(":","-",SNP_ID))) |>
#   left_join(cis_IVdeets, by = "Variant") |>
#   dplyr::rename("effect_allele.vep" = EA,
#                 "other_allele.vep" = OA) |>
#   dplyr::filter(duplicated(paste(IV_set, Variant)) == FALSE) |>
#   dplyr::mutate(ea_matches_annot = effect_allele.exposure == effect_allele.vep) 
# 
#write.table(all_cisIVs, file = file.path(res_dir, "allcisIVs_VEPannots_GW5e-08_EW5e-08_rarepruned.txt"), row.names = F, col.names = T, quote = F, sep = "\t")
#write.table(all_cisIVs_relaxed, file = file.path(res_dir, "allcisIVs_VEPannots_GW5e-08_EW1e-04_rarepruned.txt"), row.names = F, col.names = T, quote = F, sep = "\t")

# Read in list of all cis IVs with vep annotations
all_cisIVs <- data.table::fread(file = file.path(res_dir, "allcisIVs_VEPannots_GW5e-08_EW5e-08_rarepruned.txt"), sep = "\t")
all_cisIVs_relaxed <- data.table::fread(file = file.path(res_dir, "allcisIVs_VEPannots_GW5e-08_EW1e-04_rarepruned.txt"), sep = "\t")

# --------------------------------------------------------------------------------------- #

# Variables and aesthetics
n_genes <- length(harmonised_snps[[1]]) # 2937
names_genes <- sub("_.*", "", names(harmonised_snps[[1]]))
n_IVsets <- length(harmonised_snps[[1]][[1]]) #11 IV sets
names_IVsets <- names(harmonised_snps[[1]][[1]])

# 0.05/2937 proteins tested per outcome
n_proteins <- length(res_MR[[1]])
alpha <- signif(0.05/n_proteins,2)

# IVsets
cols_IVsets <- c("magenta4","magenta3","mediumorchid","pink",
                 "steelblue4","steelblue1","darkorange2","orange")
names(cols_IVsets) <- IVsets

# Read in GRCh38 gene ranges:
geneinfo <- data.table::fread(file.path(data_dir, "variant_annotation/gene_info_GRCh38.tsv"), data.table = F) |>
  dplyr::arrange(as.numeric(chr),as.numeric(start))
# Read in Olink gene ranges:
geneinfo_olink <- data.table::fread(file.path(data_dir, "sumstats/proteomics/olink_generanges_Sunetal.csv"), data.table = F) |>
  dplyr::rename(
    "chr" = "Gene CHROM",
    "start" = "Gene start",
    "stop" = "Gene end") |>
  dplyr::mutate(
    gene_biotype = "protein_coding",
    gene = `Gene symbol`)

# Split mutigene rows
geneinfo_olink <- geneinfo_olink %>%
  tidyr::separate_longer_delim(3, delim = "_") %>%
  tidyr::separate_longer_delim(c(5:9, 11), delim = ";") |>
  dplyr::filter(`Assay Target` == `Gene symbol`)

# High LD regions
# Sourced from: https://genome.sph.umich.edu/wiki/Regions_of_high_linkage_disequilibrium_%28LD%29
highld_regions <- data.table::fread(file.path(data_dir, "highld_hg38.txt"))

```

## Instrument characteristics

### IV availability

**The average number of genetic instruments (pQTLs) per protein across instrument sets**

Rare and common variant instruments (pQTLs) have been partitioned according to their proximity to the instrumented protein-coding gene into cis (+- 1MB from gene start site) and trans (\>1MB distant) sets. A similar number of cis-IVs per protein can be derived from each of the IV sets (typically \<5). Proteins that are instrumented with a large number of rare cis variants fall typically gene dense genomic regions. More trans-IVs are availble in the common-varaint instrument sets (both GWAS and ExWAS derived), suggesting there are fewer rare and ultra-rare variants that associate with protein levels in trans.

```{r}
#| label: fig-numivs
#| cache: TRUE
#| fig.width: 11
#| fig.height: 6.5

# Number of instruments per protein for each IV set
# num_IVs <- lapply(res_MR, function(x){
#   df_out <- data.frame(expand.grid(IV_set = IV_sets, cis_trans = c("cis","trans")))
# 
#   for(i in 1:length(x)){
#     df <- x[[i]] |> dplyr::summarise(unique(nsnp), .by = c(IV_set, cis_trans))
#     colnames(df)[3] <- names(x[i])
#     df_out <- left_join(df_out, df, by = c("IV_set","cis_trans"))
#   }
# 
#   df_out <- df_out |> tidyr::pivot_longer(cols = 3:length(df_out)) |> na.exclude()
#   return(df_out)
# })
# 
# num_IVs <- do.call("rbind", num_IVs) |> dplyr::mutate(outcome = sub(".*_","",name))

#write.table(num_IVs, file = file.path(res_dir, "numIVs_GW5e-08_EW5e-08_rarepruned.txt"), row.names = F, col.names = T, quote = F)
num_IVs <- data.table::fread(file.path(res_dir, "numIVs_GW5e-08_EW5e-08_rarepruned.txt"))

# Plot
p1 <- mapply(FUN = function(x,y){
  x <- x |> dplyr::filter(cis_trans == y) |>
    dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets)) |>
    dplyr::filter(label %in% variant_IVsets)
  
  ggplot(x, aes(x = label, y = value, fill = label)) +
    facet_grid(.~outcome) +
    stat_boxplot(geom ='errorbar', width = 0.2) + 
    geom_boxplot(width = 0.7) +
    scale_y_log10() +
    scale_fill_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          legend.position = "none",
          plot.title = element_text(hjust = 0.5)) +
    labs(x = "IV set",
         y = "Num. instruments", 
         title = paste0("Number of ",y ," instruments per plasma protein (p<5e-8)"))
  
}, x = list(num_IVs,num_IVs), y = c("cis","trans"), SIMPLIFY = FALSE)

gridExtra::grid.arrange(grobs = p1)

# Manuscript values:
num_IVs |> dplyr::filter(outcome == "BMI") |> group_by(IV_set, cis_trans) |> summarise(mean_IVs = mean(value)) |> knitr::kable(digits = 3)
```

A GWS theshold for significance is strict for rare variants, which must exhibit very strong effects in order to be detected despite their low minor allele frequency in the population. If this theshold is relaxed (for rare and ultrarare variants only), are more cis and trans IVs available, or common variants continue to show more protein associations in trans?

```{r}
#| label: fig-numivs-relaxed
#| cache: TRUE
#| fig.width: 11
#| fig.height: 7

# Number of instruments per protein for each IV set
# num_IVs_relaxed <- lapply(res_MR_relaxed, function(x){
#   df_out <- data.frame(expand.grid(IV_set = IV_sets, cis_trans = c("cis","trans")))
#   
#   for(i in 1:length(x)){
#     df <- x[[i]] |> dplyr::summarise(unique(nsnp), .by = c(IV_set, cis_trans))
#     colnames(df)[3] <- names(x[i])
#     df_out <- left_join(df_out, df, by = c("IV_set","cis_trans"))
#   }
#   
#   df_out <- df_out |> tidyr::pivot_longer(cols = 3:length(df_out)) |> na.exclude()
#   return(df_out)
# })
# 
# num_IVs_relaxed <- do.call("rbind", num_IVs_relaxed) |> dplyr::mutate(outcome = sub(".*_","",name))

#write.table(num_IVs_relaxed, file = file.path(res_dir, "numIVs_GW5e-08_EW1e-04_rarepruned.txt"), row.names = F, col.names = T, quote = F)
num_IVs_relaxed <- data.table::fread(file.path(res_dir, "numIVs_GW5e-08_EW1e-04_rarepruned.txt"))

# Plot
p1.2 <- mapply(FUN = function(x,y){
  x <- x |> dplyr::filter(cis_trans == y) |>
    dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets)) |>
    dplyr::filter(label %in% variant_IVsets)
  
  ggplot(x, aes(x = label, y = value, fill = label)) +
    facet_grid(.~outcome) +
    stat_boxplot(geom ='errorbar', width = 0.2) + 
    geom_boxplot(width = 0.7) +
    scale_y_log10() +
    scale_fill_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          legend.position = "none",
          plot.title = element_text(hjust = 0.5)) +
    labs(x = "IV set",
         y = "Num. instruments", 
         title = paste0("Number of ",y ," instruments per plasma protein \n(common p<5e-8; rare p<1e-4)"))
  
}, x = list(num_IVs_relaxed,num_IVs_relaxed), y = c("cis","trans"), SIMPLIFY = FALSE)

gridExtra::grid.arrange(grobs = p1.2)

# Manuscript values:
num_IVs_relaxed |> dplyr::filter(outcome == "BMI") |> group_by(IV_set, cis_trans) |> summarise(mean_IVs = mean(value), sum_IVs = sum(value)) |> knitr::kable(digits = 3)
```

**The number of proteins able to be instrumented with pQTLs across instrument sets**

Note, results below are for proteins with \>=3 independent IVs avialble, those considered instrumentable.

Although the power to detect highly significant rare-variant associations is expected to be lower than that for common variants due to to low alternate allele counts, an approximately equivalent percentage (\~25%) of the plasma proteome can be instrumented using GWAS common or ExWAS rare variants in cis (and \~6% with ultra-rare variants). ExWAS common variants provide less coverage, likely because fewer independent common cis-variants will typically be found in the coding region of a gene than within the 1MB surrounding putative regulatory region captured by GWAS. Alternatively, over 50% of the proteome can be instrumented with ExWAS and GWAS derived trans common variants relative to \<12% with rare and ultra-rare variants. This again suggests there are fewer (or less readily detected) rare trans-variant associations with plasma protein levels. This is slightly counterintuitive, as it would be anticipated that stronger rare varaint cis-effects on protein levels would also be detected as strong trans associations with proteins in interacting pathways.\

What are some potential explanations for this?:\
- The common variants that exist at intermediate frequencies in a population (i.e. have not reached fixation) are acted upon by competing balancing selective forces and show a greater number of trans assocaitions across diverse biological processes (i.e. are more plieotropic). Converse to this is the idea that rare variants can only be observed in genomic regions where the DNA sequence has been evolutionarially optimised (conserved) for a highly specific biological mechanism.\
- The nature of the genes instrumented by common and rare variants is fundamentally different. For example the proteins instrumented by common cis-IVs are more often involved in the regulation of diverse cellular pathways (e.g. variants in enhancers impacting transcriptional or cell signalling regulation), where strong-effect rare variants modify protein levels or function directly, with consequences for a smaller numebr of interacting/downstream proteins? - It's a power thing (explored below)

Questions:\
- What is the relative strength (betas) of cis and trans, rare and common pQTL effects?\
- What is the intersection of the plasma proteins that have rare and common IVs available (significant pQTL associations)?

```{r}
#| label: fig-summary
#| warning: FALSE
#| fig.width: 11
#| fig.height: 6.5

# Number of proteins with at least n cis/trans IV
avail_IVs <- num_IVs |> group_by(IV_set, cis_trans, outcome) |> 
  dplyr::summarise(n_prots_1IV = sum(value >= 1),
                   prop_prots_1IV = sum(value >= 1)/n_proteins,
                   n_prots_3IV = sum(value >= 3),
                   prop_prots_3IV = sum(value >= 3)/n_proteins)

# Plot
p2 <- sapply(c("cis","trans"), function(x){
  avail_IVs |>
    dplyr::filter(IV_set %in% names(paper_IVsets) & cis_trans == x) |>
    dplyr::mutate(labels = factor(paper_IVsets[IV_set], levels = paper_IVsets)) |>
    ggplot(aes(x = labels)) +
    facet_grid(.~outcome) +
    geom_col(aes(y = prop_prots_1IV*100, colour = labels), fill = "white", width = 0.8, lwd = 0.8) +
    geom_col(aes(y = prop_prots_3IV*100, fill = labels), width = 0.8) +
    scale_y_continuous(limits = c(0,100)) +
    scale_fill_manual(values = cols_IVsets) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          plot.title = element_text(hjust = 0.5),
          legend.position = "none") +
    labs(x = "IV set",
         y = "%", 
         title = paste0("Percentage of plasma proteins with ",x, "-IV sets available (with n>=1 or >=3 instruments; p<5e-8)"))}, simplify = FALSE)
  
gridExtra::grid.arrange(grobs = p2)

# Manuscript values:
avail_IVs |> group_by(IV_set, cis_trans) |> summarise(max(n_prots_1IV), max(prop_prots_1IV), max(n_prots_3IV), max(prop_prots_3IV)) |> knitr::kable(digits = 3)

# Total proteins instrumentable with variant/gene IVs (n=2017)
sub("_.*", "", num_IVs |> dplyr::filter(IV_set %in% names(paper_IVsets), cis_trans == "cis", value >= 1) |> dplyr::pull(name)) |> unique() |> length()
```

Repeat with relaxed thresholding for rare variants:

```{r}
#| label: fig-summary-relaxed
#| warning: FALSE
#| fig.width: 11
#| fig.height: 7

# Number of proteins with at least n cis/trans IV
avail_IVs_relaxed <- num_IVs_relaxed |> group_by(IV_set, cis_trans, outcome) |> 
  dplyr::summarise(n_prots_1IV = sum(value >= 1),
                   prop_prots_1IV = sum(value >= 1)/n_proteins,
                   n_prots_3IV = sum(value >= 3),
                   prop_prots_3IV = sum(value >= 3)/n_proteins)

# Plot
p2.2 <- sapply(c("cis","trans"), function(x){
  avail_IVs_relaxed |>
    dplyr::filter(IV_set %in% names(paper_IVsets) & cis_trans == x) |>
    dplyr::mutate(labels = factor(paper_IVsets[IV_set], levels = paper_IVsets)) |>
    ggplot(aes(x = labels)) +
    facet_grid(.~outcome) +
    geom_col(aes(y = prop_prots_1IV*100, colour = labels), fill = "white", width = 0.8, lwd = 0.8) +
    geom_col(aes(y = prop_prots_3IV*100, fill = labels), width = 0.8) +
    scale_y_continuous(limits = c(0,100)) +
    scale_fill_manual(values = cols_IVsets) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          plot.title = element_text(hjust = 0.5),
          legend.position = "none") +
    labs(x = "IV set",
         y = "%", 
         title = paste0("Percentage of plasma proteins with ",x, "-IV sets available (with n>=1 or >=3 instruments) \n(common p<5e-8; rare p<1e-4)"))}, simplify = FALSE)
  
gridExtra::grid.arrange(grobs = p2.2)

# Manuscript values 
#Number of proteins instrumentable:
avail_IVs_relaxed |> group_by(IV_set, cis_trans) |> summarise(max(n_prots_1IV), max(prop_prots_1IV), max(n_prots_3IV), max(prop_prots_3IV)) |> knitr::kable(digits = 3)

# Number of cis-IVs and cis-IV associations (common variant sets combined):
all_cisIVs |> dplyr::group_by(IV_group) |> dplyr::summarise(num_cis_IVs = length(unique(SNP_ID)), num_cis_assocs = sum(n_cis_assocs))
all_cisIVs |> dplyr::group_by(IV_set) |> dplyr::summarise(num_cis_IVs = length(unique(SNP_ID)), num_cis_assocs = sum(n_cis_assocs))

all_cisIVs_relaxed |> dplyr::group_by(IV_group) |> dplyr::summarise(num_cis_IVs = length(unique(SNP_ID)), num_cis_assocs = sum(n_cis_assocs))
all_cisIVs_relaxed |> dplyr::group_by(IV_set) |> dplyr::summarise(num_cis_IVs = length(unique(SNP_ID)), num_cis_assocs = sum(n_cis_assocs))

# Total proteins instrumentable with variant IVs (n=2051)
sub("_.*", "", num_IVs_relaxed |> dplyr::filter(IV_set %in% names(variant_IVsets), cis_trans == "cis", value >= 1) |> dplyr::pull(name)) |> unique() |> length()

```

**Instrument overlap**

What is the overlap in the proteins that have at least 1 IV across instrument sets?

```{r}
#| label: instrument overlap
#| fig.width: 6.5
#| fig.height: 3

num_IVs$protein <- sub("_.*","",num_IVs$name)

upset_cis <- list(
  "exome:rare" = num_IVs |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_rare_pruned","dhindsa_exwas_variant_ultrarare_pruned")  & cis_trans == "cis") |> 
    dplyr::pull(protein) |> unique(),
  "ex/genome:common" = num_IVs |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "cis") |>
    dplyr::pull(protein) |> unique())

upset_cis_mask <- list(
  "exome:rare" = num_IVs |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_rare_pruned","dhindsa_exwas_variant_ultrarare_pruned") & cis_trans == "cis") |> 
    dplyr::pull(protein) |> unique(),
  "exome:mask" = num_IVs |> 
    dplyr::filter(IV_set == "dhindsa_exwas_mask_ptvraredmg" & cis_trans == "cis") |> 
    dplyr::pull(protein) |> unique(),
  "ex/genome:common" = num_IVs |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "cis") |>
    dplyr::pull(protein) |> unique())

upset_trans <- list(
  "exome:rare" = num_IVs |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_rare_pruned","dhindsa_exwas_variant_ultrarare_pruned") & cis_trans == "trans") |> 
    dplyr::pull(protein) |> unique(),
  "ex/genome:common" = num_IVs |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "trans") |>
    dplyr::pull(protein) |> unique())

# Upset plots
upset(fromList(upset_cis), order.by = "freq", text.scale = 1.5, mainbar.y.label = "cis-instrumented \nproteins (P<5e-8)")
upset(fromList(upset_cis_mask), order.by = "freq", text.scale = 1.5, mainbar.y.label = "cis-instrumented \nproteins (P<5e-8)")
upset(fromList(upset_trans), order.by = "freq", text.scale = 1.5, mainbar.y.label = "trans-instrumented \n proteins (P<5e-8)")

num_IVs_relaxed$protein <- sub("_.*","",num_IVs_relaxed$name)

upset_cis_relaxed <- list(
  "exome:rare" = num_IVs_relaxed |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_rare_pruned","dhindsa_exwas_variant_ultrarare_pruned") & cis_trans == "cis") |> 
    dplyr::pull(protein) |> unique(),
  "ex/genome:common" = num_IVs_relaxed |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "cis") |>
    dplyr::pull(protein) |> unique())

upset_cis_mask_relaxed <- list(
  "exome:rare" = num_IVs_relaxed |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_rare_pruned","dhindsa_exwas_variant_ultrarare_pruned") & cis_trans == "cis") |> 
    dplyr::pull(protein) |> unique(),
  "exome:mask" = num_IVs_relaxed |> 
    dplyr::filter(IV_set == "dhindsa_exwas_mask_ptvraredmg" & cis_trans == "cis") |> 
    dplyr::pull(protein) |> unique(),
  "ex/genome:common" = num_IVs_relaxed |> 
    dplyr::filter(outcome == "BMI" & IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "cis") |>
    dplyr::pull(protein) |> unique())

upset_trans_relaxed <- list(
  "exome:rare" = num_IVs_relaxed |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_rare_pruned","dhindsa_exwas_variant_ultrarare_pruned") & cis_trans == "trans") |> 
    dplyr::pull(protein) |> unique(),
  "ex/genome:common" = num_IVs_relaxed |> 
    dplyr::filter(IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "trans") |>
    dplyr::pull(protein) |> unique())

# Upset plots
upset(fromList(upset_cis_relaxed), order.by = "freq", text.scale = 1.5, mainbar.y.label = "cis-instrumented \n proteins \n P<5e-8 common \n P<1e-4 rare")
upset(fromList(upset_cis_mask_relaxed), order.by = "freq", text.scale = 1.5, mainbar.y.label = "cis-instrumented \n proteins \n P<5e-8 common \n P<1e-4 rare")
upset(fromList(upset_trans_relaxed), order.by = "freq", text.scale = 1.5, mainbar.y.label = "trans-instrumented \n proteins \n P<5e-8 common \n P<1e-4 rare")

```

### IV effect estiamtes
**The strength of IV associations**

What are the relative effect sizes of instruments (pQTLs) in the different IV sets?

First pull the pQTL associations for each IV set into a consistant format:

```{r}
#| label: pulling pQTLs

# Pull exposure associations from harmonised SNP sets

# pull_pqtls <- function(snpset, IVsetnames) {
#   lapply(seq_len(n_genes), function(gene){
#   out <- lapply(IVsetnames, function(set){
# 
#     dfs <- lapply(snpset, function(x){
#       df <- x[[gene]][[set]]
#       if("gene.outcome" %in% colnames(df)){
#         df$gene <- df$gene.outcome
#       }else{df$gene <- toupper(df$SNP)}
# 
#       if(!("chr.exposure" %in% colnames(df))){
#         df$chr.exposure = sub(":.*","",df$variant)
#         df$pos.exposure = sub(".*:(.*)_.*_.*","\\1",df$variant)
#       }
# 
#       df <- df |> dplyr::select(SNP,
#                                 CHR = chr.exposure,
#                                 POS = pos.exposure,
#                                 EA = effect_allele.exposure,
#                                 OA = other_allele.exposure,
#                                 EAF = eaf.exposure,
#                                 beta = beta.exposure,
#                                 se = se.exposure,
#                                 P = pval.exposure,
#                                 exposure,
#                                 cis_trans,
#                                 proximal_gene = gene)
#       return(df)
#     })
# 
#     dfs_join <- do.call("rbind", dfs) |> unique()
#     rownames(dfs_join) <- NULL
#     dfs_join
#   })
#   names(out) <- IVsetnames
#   out
# })}
  
#pqtls <- pull_pqtls(snpset = harmonised_snps, IVsetnames = IV_sets)
#names(pqtls) <- names_genes
#save(pqtls, file = file.path(data_dir,"sumstats/proteomics/protein_pQTLs_5e-08.rda"))

# pqtls_relaxed <- pull_pqtls(snpset = harmonised_snps_relaxed, IVsetnames = IV_sets)
# names(pqtls_relaxed) <- names_genes
# save(pqtls_relaxed, file = file.path(data_dir,"sumstats/proteomics/protein_pQTLs_5e-08GW_1e-04EW.rda"))

load(file = file.path(data_dir,"sumstats/proteomics/protein_pQTLs_5e-08.rda")) #pqtls
load(file = file.path(data_dir,"sumstats/proteomics/protein_pQTLs_5e-08GW_1e-04EW.rda")) #pqtls_relaxed

### Fix the list of relaxed threshold SNPs written out for lookup....

# # Significant pQTLs, combined across all exposures, for each instrument sets
# pqtls_joined <- lapply(seq_len(n_IVsets), function(set){
#   #dfs <- lapply(pqtls, function(x){
#   dfs <- lapply(pqtls_relaxed, function(x){
#     out <- x[[set]]
#     out$IVset <- names_IVsets[set]
#     return(out)
#   })
#   dfs_join <- do.call("rbind", dfs) |> dplyr::arrange(SNP)
#   rownames(dfs_join) <- NULL
#   return(na.exclude(dfs_join))
# })
# 
# names(pqtls_joined) <- names_IVsets
# 
# # Derive a set of variants that are pQTLs for at least one protein to look up in the raw summary statistics of all others (split by IV sets)
# snp_lookup <- lapply(pqtls_joined, function(x){
#   unique(x$SNP)
# })
# # Format as in summary stats
# snp_lookup[grepl("mask", names(snp_lookup))] <- lapply(snp_lookup[grepl("mask", names(snp_lookup))], function(x){toupper(x)})
# snp_lookup[grepl("exwas_variant", names(snp_lookup))] <- lapply(snp_lookup[grepl("exwas_variant", names(snp_lookup))],
#                                                                 function(x){gsub(":","-",gsub("_","-",toupper(x)))})
#save(snp_lookup, file = file.path(data_dir,"sumstats/proteomics/protein_pQTLstolookup_5e-08.rda"))
#save(snp_lookup, file = file.path(data_dir,"sumstats/proteomics/protein_pQTLstolookup_5e-08GW_1e-04EW.rda"))
# Variant lookup performed as in 002_lookuppQTLs.R (on P1 server)
```

```{r}
#| label: instrument strength
#| fig.height: 4
#| fig.width: 12

pqtls_joined <- sapply(names(IVsets), function(x){do.call(rbind, lapply(pqtls, `[[`, x))} |> na.exclude(), simplify = F)
# Correct variant names and add proxinal gene info to GWAS common IVs
pqtls_joined$sun_gwas_variant_common_clumped <- pqtls_joined$sun_gwas_variant_common_clumped |> 
  dplyr::mutate(SNP = paste0(CHR,"-",POS,"-",EA,"-",OA)) |>
  dplyr::mutate(proximal_gene = gwas_deets[match(SNP, gwas_deets$Variant)]$GENE)

rm(gwas_deets) # Clean workspace

pqtls_joined_variants <- do.call("rbind", pqtls_joined[grep("variant", names(pqtls_joined))]) |>
  tibble::rownames_to_column("IV_set") |>
  dplyr::mutate(IV_set = sub("\\..*","",IV_set),
                cis_trans_true = ifelse(proximal_gene == exposure, "true_cis", cis_trans),
                SNP = toupper(sub(":","-",gsub("_","-",SNP))))

p3 <- pqtls_joined_variants |> 
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |> 
  ggplot(aes(x = IV_set, y = abs(beta))) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Instrument |exposure beta|")

p4 <- pqtls_joined_variants |> dplyr::select(-IV_set) |> unique() |>
  dplyr::mutate(MAF = ifelse(EAF>0.5, 1-EAF,EAF),
                beta_MA = ifelse(EAF>0.5, beta*-1, beta)) |>
  ggplot(aes(x = MAF , y = beta_MA)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_vline(xintercept = c(0.01,0.001), col = "red", lty = 2, lwd = 0.5) +
  theme_minimal() +
  labs(x = "MAF", y = "Instrument exposure beta")

gridExtra::grid.arrange(p3,p4,nrow = 1, widths = c(0.18,0.75), top = "Observed effect size of IVs for protein expression across instrument sets")
```

```{r}
#| label: instrument-strength-split
#| fig.height: 3.5
#| fig.width: 5

## Gene masks
pqtls_joined_masks <- do.call("rbind", pqtls_joined[grep("dhindsa_exwas_mask_ptvraredmg", names(pqtls_joined))]) |>
  tibble::rownames_to_column("IV_set") |>
  dplyr::mutate(IV_set = sub("\\..*","",IV_set),
                cis_trans_true = ifelse(proximal_gene == toupper(exposure), "true_cis", cis_trans))

# Split variant cis and trans effects
pqtls_joined_variants |> 
  dplyr::filter(!(cis_trans_true == "")) |>
  dplyr::mutate(cis_trans_true = factor(cis_trans_true, levels = c("trans","cis","true_cis"))) |>
  rbind(pqtls_joined_masks) |>
  dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets)) |> 
  ggplot(aes(x = cis_trans_true, y = abs(beta), fill = label)) +
  facet_grid(.~label, scale = "free", space = "free") +
  geom_boxplot() +
  scale_fill_manual(values = cols_IVsets) +
  #scale_x_discrete(labels = c("trans","proximal \ncis", "true \ncis")) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "", y = "Instrument |exposure beta|", title = "Effect size of IVs (P<5e-8)")

# Cis only
pqtls_joined_variants |> 
  dplyr::filter(!(cis_trans_true == "")) |>
  dplyr::mutate(cis_trans_true = factor(cis_trans_true, levels = c("trans","cis","true_cis"))) |>
  rbind(pqtls_joined_masks) |>
  dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets)) |> 
  dplyr::filter(cis_trans_true != "trans") |>
  ggplot(aes(x = cis_trans_true, y = abs(beta), fill = label)) +
  facet_grid(.~label, scale = "free", space = "free") +
  geom_boxplot() +
  scale_fill_manual(values = cols_IVsets) +
  scale_x_discrete(labels = c("cis" = "proximal \ncis", "true_cis" = "true \ncis")) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "", y = "Instrument |exposure beta|", title = "Effect size of IVs (P<5e-8)")

```

### Genomic location

```{r}
#| label: genomic-location
#| fig.height: 4
#| fig.width: 9

# Valid cis-IVs for MR (filter to those in harmonised_snps)
pqtls_joined_variants_MRset <- pqtls_joined_variants |> 
  dplyr::filter(cis_trans == "cis",
                SNP %in% all_cisIVs$Variant) |>
  dplyr::left_join(unique(all_cisIVs[,c("Variant","ANNOTATION","GENE","Amino acid change","Variant_group")]), by = c("SNP" = "Variant"))
  
# Plot 1: Proportion of true_cis vs cis by IV_set
p.truecis <- pqtls_joined_variants_MRset |>
  dplyr::group_by(IV_set, cis_trans_true) |> dplyr::summarise(num_cis_IVs = length(unique(SNP))) |>
  dplyr::mutate(prop = num_cis_IVs/sum(num_cis_IVs)) |>
  ggplot(aes(x = factor(IVsets[IV_set], levels = IVsets), y = prop, fill = cis_trans_true)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(name = "", values = c("grey90","grey30")) +
  labs(y = "Proportion", x = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "right")

p.IVgroup <- pqtls_joined_variants_MRset |>
  dplyr::group_by(IV_set, cis_trans_true, Variant_group) |> dplyr::summarise(num_cis_IVs = length(unique(SNP))) |>
  na.exclude() |> 
  dplyr::mutate(prop = num_cis_IVs/sum(num_cis_IVs)) |>
  ggplot(aes(x = cis_trans_true, y = prop, fill = Variant_group)) +
  facet_grid(.~factor(IVsets[IV_set], levels = IVsets)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cols_consequences) +
  labs(y = "Proportion", x = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "right")

gridExtra::grid.arrange(grobs = list(p.truecis, p.IVgroup), nrow = 1, widths = c(0.35, 0.65))

# Manuscript values:
p.truecis@data |> knitr::kable()
```

### Pleiotropy profiles

On average, how many molecular traits (protein levels) does a variant significantly associate with across instrument sets (e.g. ultra-rare, rare, common)?

To assess the pleiotropy profiles of variants in each IV set, all significant pQTL associations reported with a given variant (p\<=5e-8; or p\<=1e-4 for rare varaints at a relaxed threshod) were extracted from the O-link protein GWAS/ExWAS summary statistics. LD pruning was then performed across all varaints - such that the variants used here may differ from the exct variants used in subsequent MR (which were pruned within (not between) each IV set)

```{r}
#| label: pleiotropy-matrix

lookup_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_*", full.names = TRUE)

#lookup_masks_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_dhindsa_exwas_mask_*", full.names = TRUE)
#lookup_exome_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_dhindsa_exwas_variant_*", full.names = TRUE)
#lookup_genome_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_sun_gwas_variant_*", full.names = TRUE)

# Read in exposure IVs
lookup_pQTLs <- lapply(lookup_files, function(x){data.table::fread(x)})
names(lookup_pQTLs) <- sub(".*pQTL_lookup_(.*)\\.tsv","\\1",lookup_files)

# Format
lookup_pQTLs <- lapply(lookup_pQTLs, function(x){
  x <- as.data.frame(x)
  if(!("GENE" %in% colnames(x))){x$GENE <- NA}
  
  out <- data.frame(variant = sub(".*\\.tsv:","",x[,1]),
                    protein_target = x$protein_target,
                    proximal_gene = x$GENE,
                    pvalue = x[,which(colnames(x) %in% c("P","pValue"))],
                    cis_trans = x$cis_trans) |>
    dplyr::arrange(variant, pvalue) |>
    dplyr::filter(duplicated(paste(variant,protein_target)) == FALSE) |>
  unique()
  
  return(out)})

# Pre-cross-IV set pruning (retain all variants used in MR analyses)
lookup_pQTLs_MRset <- lookup_pQTLs
# Threshold common exome variants
lookup_pQTLs_MRset$dhindsa_exwas_variant_common <- lookup_pQTLs_MRset$dhindsa_exwas_variant_common |> dplyr::filter(pvalue <= 5e-8)

# Retain LD pruned common variants (such that all SNPs in pleiotropy analysis are independent)
lookup_pQTLs$sun_gwas_variant_common_clumped <- lookup_pQTLs$sun_gwas_variant_common_clumped |>
 dplyr::filter(variant %in% prunein_gwas)

lookup_pQTLs$dhindsa_exwas_variant_common <- lookup_pQTLs$dhindsa_exwas_variant_common |>
dplyr::filter(variant %in% gsub(":","-",prunein_exwas))

# Relaxed theshold for rare variants (P<1e-4)
lookup_pQTLs_relaxed <- lapply(lookup_pQTLs, function(x){
  x |> dplyr::filter(!(cis_trans == "")) # Exclude composite protein targets with no cis_trans assignment
})

# Retain LD pruned rare variants
lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein_relaxed))))
lookup_pQTLs_relaxed$dhindsa_exwas_variant_ultrarare_pruned <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_ultrarare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein_relaxed))))
lookup_pQTLs_relaxed <- lookup_pQTLs_relaxed[names(IVsets)]
#lapply(lookup_pQTLs_relaxed, function(x){max(x$pvalue)})

lookup_pQTLs_relaxed_both <- lookup_pQTLs_relaxed # pre thresholding common exome varaints
# Threshold common exome variants
lookup_pQTLs_relaxed$dhindsa_exwas_variant_common <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |> dplyr::filter(pvalue <= 5e-8)

# Retain associations at 5x10^-8 (note gene masks are already thresholded at 1x10^-8)
lookup_pQTLs <- lapply(lookup_pQTLs, function(x){
  x |> dplyr::filter(pvalue <= 5e-8 & !(cis_trans == "")) # Exclude composite protein targets with no cis_trans assignment
})
# Prune rare variants
lookup_pQTLs$dhindsa_exwas_variant_rare_pruned <- lookup_pQTLs$dhindsa_exwas_variant_rare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein))))
lookup_pQTLs$dhindsa_exwas_variant_ultrarare_pruned <- lookup_pQTLs$dhindsa_exwas_variant_ultrarare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein))))
lookup_pQTLs <- lookup_pQTLs[names(IVsets)]
#lapply(lookup_pQTLs, function(x){max(x$pvalue)})

# Annotate proximal gene and protein target with the chromosome they are on
lookup_pQTLs <- lapply(lookup_pQTLs, function(x){
  x <- x |> dplyr::mutate(protein_target_chr = sub(";.*","",geneinfo_olink[match(x$protein_target, geneinfo_olink$`Assay Target`), "chr"]),
                          proximal_gene_chr = geneinfo[match(gsub("'", "", x$proximal_gene), geneinfo$gene), "chr"])
  return(x)
})

lookup_pQTLs_relaxed <- lapply(lookup_pQTLs_relaxed, function(x){
  x <- x |> dplyr::mutate(protein_target_chr = sub(";.*","",geneinfo_olink[match(x$protein_target, geneinfo_olink$`Assay Target`), "chr"]),
                          proximal_gene_chr = geneinfo[match(gsub("'", "", x$proximal_gene), geneinfo$gene), "chr"])
  return(x)
})

# For gene masks - reassign cis/trans calls to match the method used for varaints (+-1Mb from gene start site)
# Currently only masks for the encoded protein are considered cis (considered these instead as "true_cis" below)
lookup_pQTLs_relaxed[grepl("mask", names(lookup_pQTLs_relaxed))] <- lapply(lookup_pQTLs_relaxed[grepl("mask", names(lookup_pQTLs_relaxed))], function(x){
  x$proximal_gene = x$variant
  
  for(i in 1:nrow(x)){
    if(x$proximal_gene[i] %in% geneinfo$gene){
      prox_gene_chr = as.character(geneinfo[geneinfo$gene == x$proximal_gene[i],"chr"])
      prox_gene_range = geneinfo[geneinfo$gene == x$proximal_gene[i],c("start","stop")]
      prot_gene_chr = as.character(geneinfo_olink[geneinfo_olink$gene == x$protein_target[i],"chr"])
      prot_gene_start = as.numeric(geneinfo_olink[geneinfo_olink$gene == x$protein_target[i],"start"])
      
      prot_gene_cisrange = c(start = prot_gene_start - 1000000, stop = prot_gene_start + 1000000)
      
      in_cis <- prox_gene_chr == prot_gene_chr & 
        ((prox_gene_range["start"] > prot_gene_cisrange["start"] & prox_gene_range["start"] < prot_gene_cisrange["stop"]) |
           (prox_gene_range["stop"] > prot_gene_cisrange["start"] & prox_gene_range["stop"] < prot_gene_cisrange["stop"])) |> as.logical()
      
      x$cis_trans[i] = ifelse(in_cis, "cis","trans")
      x$proximal_gene_chr[i] = prox_gene_chr
    } else {x$cis_trans[i] = NA}}
  return(x)
})

# Summarise associations per variant (pQTL)
pQTLs_assocs <- mapply(function(x, y){
  x <- x |> dplyr::group_by(variant, proximal_gene) |> 
    dplyr::summarise(n_assocs = n(), 
                     n_cis = length(which(cis_trans == "cis")),
                     n_truecis = sum(proximal_gene == paste0("'",protein_target,"'")),
                     n_trans = length(which(cis_trans == "trans")),
                     assoc_genes = paste(unique(protein_target), collapse = ";"),
                     assoc_cis_genes = paste(unique(protein_target[cis_trans == "cis"]), collapse = ";"),
                     assoc_trans_genes = paste(unique(protein_target[cis_trans == "trans"]), collapse = ";"))
  x$IV_set <- y
  return(x)
}, x = lookup_pQTLs, y = as.list(names(lookup_pQTLs)), SIMPLIFY = F)

# Summarise the number of cis and trans genes each variant associates with
pQTLs_assocs_relaxed <- mapply(function(x, y){
  if(grepl("variant",y)){
    x <- x |> dplyr::group_by(variant, proximal_gene) |> 
      dplyr::summarise(n_assocs = n(), 
                       n_cis = length(which(cis_trans == "cis")),
                       n_truecis = sum(proximal_gene == paste0("'",protein_target,"'")),
                       n_trans = length(which(cis_trans == "trans")),
                       assoc_genes = paste(unique(protein_target), collapse = ";"),
                       assoc_cis_genes = paste(unique(protein_target[cis_trans == "cis"]), collapse = ";"),
                       assoc_trans_genes = paste(unique(protein_target[cis_trans == "trans"]), collapse = ";"))
  }
  if(grepl("mask",y)){
    x <- x |> dplyr::group_by(variant, proximal_gene) |> 
      dplyr::summarise(n_assocs = n(), 
                       n_cis = length(which(cis_trans == "cis")),
                       n_truecis = sum(proximal_gene == protein_target),
                       n_trans = length(which(cis_trans == "trans")),
                       assoc_genes = paste(unique(protein_target), collapse = ";"),
                       assoc_cis_genes = paste(unique(protein_target[cis_trans == "cis"]), collapse = ";"),
                       assoc_trans_genes = paste(unique(protein_target[cis_trans == "trans"]), collapse = ";"))
  }
  x$IV_set <- y
  return(x)
}, x = lookup_pQTLs_relaxed, y = as.list(names(lookup_pQTLs_relaxed)), SIMPLIFY = F)

```

**Cis and trans plieotropy profiles**

```{r}
#| label: cis-v-trans-pleiotropy
#| fig.height: 4
#| fig.width: 7
#| include: FALSE

pqtls_assocs_joined <- do.call("rbind", pQTLs_assocs) |>
  dplyr::filter(grepl("variant",IV_set)) |>
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |>
  dplyr::group_by(IV_set) |> 
  dplyr::mutate(label = ifelse(n_trans == max(n_trans) & grepl("exome",IV_set), paste(variant,proximal_gene), NA)) |> dplyr::ungroup()

# Filter to varaints that will be used as cis-IVs (at least 1 cis-pQTL association)
p5 <- pqtls_assocs_joined  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = IV_set, y = n_assocs)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations")

## Number of trans associations of each true-cis or proximal-cis pQTL
# pqtls_assocs_joined |> 
#   dplyr::filter(n_cis > 0 & !(is.na(n_truecis))) |>
#   dplyr::mutate(x = paste(IV_set,n_truecis, sep = "_")) |>
#   ggplot(aes(x = x, y = n_assocs)) +
#   geom_boxplot(aes(fill = IV_set)) +
#   scale_fill_manual(values = cols_IVsets) +
#   scale_y_log10() +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 55, hjust = 1),
#         legend.position = "none") +
#   labs(x = "", y = "Number of pQTL associations")

p6 <- pqtls_assocs_joined  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-pQTL associations", y = "Number of trans-pQTL associations")

gridExtra::grid.arrange(p5,p6,nrow = 1, widths = c(0.2,0.7), top = "Pleiotropy profile of IV sets (P<5e-8)")
```

```{r}
#| label: cis-v-trans-pleiotropy-relaxed
#| fig.height: 4
#| fig.width: 15

# Variant pleiotropy profiles
pqtls_assocs_joined_relaxed <- do.call("rbind", pQTLs_assocs_relaxed) |>
  dplyr::filter(IV_set %in% names(paper_IVsets)) |>
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |>
  dplyr::group_by(IV_set) |> 
  dplyr::mutate(label = ifelse(n_trans == max(n_trans) & grepl("exome",IV_set), paste(variant,proximal_gene), NA)) |> dplyr::ungroup()

# Filter to varaints that will be used as cis-IVs (at least 1 cis-pQTL association)
p5.2.1 <- pqtls_assocs_joined_relaxed  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = IV_set, y = n_assocs)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations\n per cis-IV")

p5.2.2 <- pqtls_assocs_joined_relaxed  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = IV_set, y = n_cis)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of cis-window pQTL\n associations per cis-IV")

p5.2.3 <- pqtls_assocs_joined_relaxed  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = IV_set, y = n_trans)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of trans-window pQTL\n associations per cis-IV")

p6.2 <- pqtls_assocs_joined_relaxed  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_jitter(width = 0.2, height = 0) +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-window pQTL associations per cis-IV", y = "Number of trans-window pQTL\n associations per cis-IV")

p6.2.zoom <- pqtls_assocs_joined_relaxed  |> dplyr::filter(n_cis >= 1) |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_jitter(width = 0.2, height = 0) +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  scale_y_continuous(limits = c(0,200)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-pQTL associations per cis-IV", y = "Number of trans-pQTL associations\n per cis-IV")

gridExtra::grid.arrange(p5.2.2, p5.2.3, p6.2.zoom,nrow = 1, widths = c(0.15,0.18,0.7), top = "Pleiotropy profile of IV sets (common: P<5e-8; rare: P<1e-4) [Y-ZOOMED]")

# Proportional of cis-IVs with x other QTL associations
p.propncis.plotdat <- pqtls_assocs_joined_relaxed |>
  dplyr::filter(IV_set %in% paper_IVsets, n_cis >= 1) |>
  dplyr::mutate(n_cis_plot = factor(ifelse(n_cis >= 5, ">=5", as.character(n_cis)), levels = c(">=5","5","4","3","2","1"))) |>
  dplyr::count(IV_set, n_cis_plot) |>
  group_by(IV_set) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

p.propncis <- p.propncis.plotdat |>
  ggplot(aes(x = IV_set,y = prop, fill = factor(n_cis_plot))) +
  geom_col(position = "fill", col = "black") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Greys", name = "n_cis", direction = -1) +
  labs(x = "", y = "Proportion of cis-pQTLs") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1))

gridExtra::grid.arrange(p.propncis, p5.2.2, p5.2.3, p6.2,nrow = 1, widths = c(0.18,0.14,0.14,0.54), top = "Pleiotropy profile of IV sets (common: P<5e-8; rare: P<1e-4)")

gridExtra::grid.arrange(p.propncis, p5.2.2, p5.2.3, p6.2.zoom,nrow = 1, widths = c(0.18,0.14,0.14,0.54), top = "Pleiotropy profile of IV sets (common: P<5e-8; rare: P<1e-4) [Y-ZOOMED]")
  
# Average number of associations
do.call("rbind", pqtls_assocs_joined_relaxed |> dplyr::filter(n_cis >= 1) |> group_by(IV_set) |> group_split() |> lapply(function(x){
  data.frame(IVset = unique(x$IV_set), n_cis_vars = nrow(x), prop_cis_gt1 = sum(x$n_cis > 1)/nrow(x))}))

do.call("rbind", pqtls_assocs_joined_relaxed |> dplyr::filter(n_truecis >= 1) |> group_by(IV_set) |> group_split() |> lapply(function(x){
  data.frame(IVset = unique(x$IV_set), n_cis_vars = nrow(x), n_cis_gt1 = sum(x$n_cis > 1), prop_cis_gt1 = sum(x$n_cis > 1)/nrow(x))}))


```

Restricting to genes that have both rare and common cis-IVs

```{r}
#| label: cis-v-trans-pleiotropy-relaxed-sharegenes
#| fig.height: 4
#| fig.width: 9
#| include: FALSE

# Filter to genes that have both common (genome) and rare cis-IVs (n=1421 genes)
shared_cis_genome <- intersect(unique(lookup_pQTLs_relaxed$sun_gwas_variant_common_clumped |> dplyr::filter(cis_trans == "cis") |> dplyr::pull(protein_target)),
                               unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned |> dplyr::filter(cis_trans == "cis") |> dplyr::pull(protein_target)))
# Filter to genes that have both common (exome) and rare cis-IVs (n=941 genes)
shared_cis_exome <- intersect(unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |> dplyr::filter(cis_trans == "cis") |> dplyr::pull(protein_target)),
                              unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned |> dplyr::filter(cis_trans == "cis") |> dplyr::pull(protein_target)))

pqtls_assocs_joined_relaxed_sharedcis <- do.call("rbind", pQTLs_assocs_relaxed) |>
  dplyr::filter(grepl("variant",IV_set),
                sapply(strsplit(assoc_cis_genes, ";"), function(x){any(x %in% unique(c(shared_cis_genome,shared_cis_exome)))})) |> # Keep varaints that are cis-QTLs for shared genes
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |>
  dplyr::group_by(IV_set) |> 
  dplyr::mutate(label = ifelse(n_trans == max(n_trans) & grepl("exome",IV_set), paste(variant,proximal_gene), NA)) |> dplyr::ungroup()

p5.3.1 <- pqtls_assocs_joined_relaxed_sharedcis |>
  ggplot(aes(x = IV_set, y = n_assocs)) +
  facet_grid(.~n_truecis) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations")

p5.3.2 <- pqtls_assocs_joined_relaxed_sharedcis |>
  ggplot(aes(x = IV_set, y = n_cis)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of cis-pQTL associations\n per cis-IV")

p5.3.3 <- pqtls_assocs_joined_relaxed_sharedcis |>
  ggplot(aes(x = IV_set, y = n_trans)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of trans-pQTL associations\n per cis-IV")

p6.3 <- pqtls_assocs_joined_relaxed_sharedcis |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  geom_jitter(width = 0.2, height = 0) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-pQTL associations per cis-IV", y = "Number of trans-pQTL associations\n per cis-IV")

gridExtra::grid.arrange(p5.3.2, p5.3.3, p6.3, nrow = 1, widths = c(0.15,0.15,0.7), top = "Pleiotropy profile of IV sets (common: P<5e-8; rare: P<1e-4) - cis-QTL shared genes")

p6.3.zoom <- pqtls_assocs_joined_relaxed_sharedcis |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_jitter(width = 0.2, height = 0) +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  scale_y_continuous(limits = c(0,200)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-pQTL associations per cis-IV", y = "Number of trans-pQTL associations\n per cis-IV [Y-ZOOMED]")

gridExtra::grid.arrange(p5.3.2, p5.3.3, p6.3.zoom, nrow = 1, widths = c(0.18,0.18,0.64), top = "Pleiotropy profile of IV sets (common: P<5e-8; rare: P<1e-4) - cis-QTL shared genes")

#pqtls_assocs_joined_relaxed_sharedcis |> group_by(IV_set) |> group_split() |> lapply(summary)

```

Can we take genes that have both rare and common cis-IVs or trans-IVs available (using the relaxed p-value theshold for rare variants) and look at how many other genes, on average, IVs in these genes associate with? We might hypothesise that perturbations to a gene (either via a common or rare variant) will exhibit similar vertical pleiotropic effects on genes downstream. Although there is differences in the power to detect rare vs common variant effects, at relaxed p-value thresholds many more rare trans QTLs can be detected than common (pontentially due to the larger effect sizes of rare trans IVs), and almost all O'link proteins have rare-trans associations (so we can see them). But *how many* other proteins do these variants associate with (i.e. how specific are they?)? Lower specificity of common over rare IVs may suggest trans effects that are mediated through LD (genetic confounding) when instrumented with common variants, which is less problematic for rare variants:

```{r}
#| label: rare-v-common-pleiotropy
#| fig.height: 4
#| fig.width: 13.5

# Using the most pleiotropic cis-IVs for shared genes
# Repeated below restricting to those IVs that have true cis effects (exonic variants in instrumented gene)

## Filter to top pleiotropic vartaint per gene (of exome:rare and genome:commmon shared sets, n = 1249)
pqtls_assocs_joined_relaxed_toppleiotropic <- pqtls_assocs_joined_relaxed_sharedcis |>
  dplyr::mutate(all_cis_genes = assoc_cis_genes) |>
  tidyr::separate_rows(assoc_cis_genes, sep = ";") |>
  dplyr::filter(IV_set %in% c("exome:rare","genome:common"),
                n_cis >= 1,
                assoc_cis_genes %in% shared_cis_genome) |>
  dplyr::arrange(assoc_cis_genes, desc(n_assocs)) |>
  dplyr::group_by(IV_set) |>
  dplyr::filter(duplicated(assoc_cis_genes) == FALSE) |> dplyr::ungroup()

## Variant pleiotropy
p7.cis <- pqtls_assocs_joined_relaxed_toppleiotropic  |>
  ggplot(aes(x = IV_set, y = n_cis)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of cis-pQTL associations\n per cis-IV")

# w1 <- wilcox.test(pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(IV_set == "exome:rare") |> dplyr::pull(n_cis),
#             pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(IV_set == "genome:common") |> dplyr::pull(n_cis))
# P=1.3x10^-16

p7.trans <- pqtls_assocs_joined_relaxed_toppleiotropic  |>
  ggplot(aes(x = IV_set, y = n_assocs)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations\n per cis-IV")

# w2 <- wilcox.test(pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(IV_set == "exome:rare") |> dplyr::pull(n_assocs),
#              pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(IV_set == "genome:common") |> dplyr::pull(n_assocs))
# P=0.0007

## Gene wise comparison
p7.plotdat <- pqtls_assocs_joined_relaxed_toppleiotropic[,c("assoc_cis_genes", "IV_set", "n_cis", "all_cis_genes")] |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(n_cis, all_cis_genes),
    names_glue = "{IV_set}_{.value}"
  ) |>
  dplyr::filter(!(is.na(`exome:rare_n_cis`)) & !(is.na(`genome:common_n_cis`))) |> 
  rowwise() |>
  dplyr::mutate( # Calc prpoprtion of shared cis genes
    prop_shared = {
      s1 <- strsplit(`exome:rare_all_cis_genes`, ";")[[1]]
      s2 <- strsplit(`genome:common_all_cis_genes`, ";")[[1]]
      length(intersect(s1, s2)) / length(union(s1, s2))
    }) |> ungroup()

p7 <- p7.plotdat |>
  ggplot(aes(x = `exome:rare_n_cis`, y = `genome:common_n_cis`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_jitter(size = 0.5) +
  scale_x_continuous(limits = c(0,14), breaks = 1:14) +
  scale_y_continuous(limits = c(0,14), breaks = 1:14) +
  scale_colour_gradient(low = "grey90", high = "red") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. cis-pQTL associations \n (rare cis-IV for gene)",
       y = "No. cis-pQTL associations \n (common cis-IV for gene)")

p8.plotdat <- pqtls_assocs_joined_relaxed_toppleiotropic[,c("assoc_cis_genes", "IV_set", "n_trans", "assoc_trans_genes")] |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(n_trans, assoc_trans_genes),
    names_glue = "{IV_set}_{.value}"
  ) |>
  dplyr::filter(!(is.na(`exome:rare_n_trans`)) & !(is.na(`genome:common_n_trans`))) |> 
  rowwise() |>
  dplyr::mutate( # Calc prpoprtion of shared trans genes
    prop_shared = {
      s1 <- strsplit(`exome:rare_assoc_trans_genes`, ";")[[1]]
      s2 <- strsplit(`genome:common_assoc_trans_genes`, ";")[[1]]
      length(intersect(s1, s2)) / length(union(s1, s2))
    }) |> ungroup()
  
p8 <- p8.plotdat |> dplyr::arrange(prop_shared) |>
  ggplot(aes(x = `exome:rare_n_trans`, y = `genome:common_n_trans`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_point(size = 0.5) +
  scale_x_continuous(limits = c(0,650)) +
  scale_y_continuous(limits = c(0,650)) +
  scale_colour_gradient(low = "grey", high = "red") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. trans-pQTL associations \n (rare cis-IV for gene)",
       y = "No. trans-pQTL associations \n (common cis-IV for gene)")
  
p8_inset <- p8.plotdat |> dplyr::arrange(prop_shared) |>
  ggplot(aes(x = `exome:rare_n_trans`, y = `genome:common_n_trans`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_jitter() +
  scale_x_continuous(limits = c(0,30)) +
  scale_y_continuous(limits = c(0,30)) +
  scale_colour_gradient(low = "grey", high = "red") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.background = element_rect(fill = "white", colour = "grey")) +
  labs(x = "No. trans-pQTL associations \n (rare cis-IV for gene)",
       y = "No. trans-pQTL associations \n (common cis-IV for gene)")

p8_join <- cowplot::ggdraw() + cowplot::draw_plot(p8) + cowplot::draw_plot(p8_inset, x = 0.5, y = 0.6, width = 0.95 - 0.5, height = 0.95 - 0.6)

## Average proportion of shared cis genes
summary(p7.plotdat |> dplyr::filter(`exome:rare_n_cis` > 1 | `genome:common_n_cis` > 1) |> dplyr::pull(prop_shared))
## Average proportion of shared tranas genes
summary(p8.plotdat$prop_shared)
  
# Correlation between cis and trans associations 
p8.2 <-  pqtls_assocs_joined_relaxed_toppleiotropic |>
  ggplot(aes(x = n_cis, y = n_trans, group = IV_set, colour = IV_set)) +
  geom_jitter(width = 0.2, height = 0, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_log10() +
  #scale_x_log10() +
  scale_colour_manual(name = "", values = cols_IVsets) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. cis-pQTL associations",
       y = "No. trans-pQTL associations")

# Associations
dat_c <- pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(IV_set == "exome:rare") |> dplyr::mutate(n_trans = n_trans + 1)
sum_c <- summary(lm(log10(n_trans) ~ n_cis, data = dat_c))
10^sum_c$coefficients["n_cis",1] # beta = 1.13 
sum_c$coefficients["n_cis",4] #P = 1.4x10^-08

dat_r <- pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(IV_set == "genome:common") |> dplyr::mutate(n_trans = n_trans + 1)
sum_r <- summary(lm(log10(n_trans) ~ n_cis, data = dat_r))
10^sum_r$coefficients["n_cis",1] #beta = 1.20 
sum_r$coefficients["n_cis",4] #P = 6.0x10^-19

gridExtra::grid.arrange(p7.cis, p7.trans, p7, p8_join, p8.2, nrow = 1, widths = c(0.1,0.1,0.26,0.26,0.26), top = "Average pleiotropy of rare or common cis-IVs for a given gene (cis-QTL shared genes)")

## Flagged genes:
#pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(assoc_cis_genes == "SIGLEC1") |> pull(assoc_trans_genes)
#pqtls_assocs_joined_relaxed_toppleiotropic |> dplyr::filter(assoc_cis_genes == "PLG") |> pull(assoc_trans_genes)
```

```{r}
#| label: rare-v-common-pleiotropy-truecis
#| fig.height: 4
#| fig.width: 13.5

# Restricting to exonic varaints with true-cis effects
shared_truecis <- intersect(unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |> dplyr::filter(proximal_gene == paste0("'",protein_target, "'")) |> dplyr::pull(proximal_gene)),
                            unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare |> dplyr::filter(proximal_gene == paste0("'",protein_target, "'")) |> dplyr::pull(proximal_gene)))

## Filter to top pleiotropic vartaint per gene (of exome:rare and genome:commmon shared true cis sets, n = 301)
pqtls_assocs_joined_relaxed_truecis <- pqtls_assocs_joined_relaxed_sharedcis |>
  dplyr::mutate(all_cis_genes = assoc_cis_genes) |>
  tidyr::separate_rows(assoc_cis_genes, sep = ";") |>
  dplyr::filter(IV_set %in% c("exome:rare","exome:common"),
                proximal_gene %in% shared_truecis,
                proximal_gene == paste0("'",assoc_cis_genes, "'")) |>
  dplyr::arrange(assoc_cis_genes, desc(n_assocs)) |>
  dplyr::group_by(IV_set) |>
  dplyr::filter(duplicated(assoc_cis_genes) == FALSE) |> dplyr::ungroup()

## Variant pleiotropy
p7.2.cis <- pqtls_assocs_joined_relaxed_truecis  |>
  ggplot(aes(x = IV_set, y = n_cis)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of cis-pQTL associations\n per cis-IV")

p7.2.trans <- pqtls_assocs_joined_relaxed_truecis  |>
  ggplot(aes(x = IV_set, y = n_trans + n_cis)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations\n per cis-IV")

## Gene wise comparison
p7.2.plotdat <- pqtls_assocs_joined_relaxed_truecis[,c("assoc_cis_genes", "IV_set", "n_cis", "all_cis_genes")] |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(n_cis, all_cis_genes),
    names_glue = "{IV_set}_{.value}"
  ) |>
  dplyr::filter(!(is.na(`exome:rare_n_cis`)) & !(is.na(`exome:common_n_cis`))) |> 
  rowwise() |>
  dplyr::mutate( # Calc prpoprtion of shared cis genes
    prop_shared = {
      s1 <- strsplit(`exome:rare_all_cis_genes`, ";")[[1]]
      s2 <- strsplit(`exome:common_all_cis_genes`, ";")[[1]]
      length(intersect(s1, s2)) / length(union(s1, s2))
    }) |> ungroup()

p7.2 <- p7.2.plotdat |>
  ggplot(aes(x = `exome:rare_n_cis`, y = `exome:common_n_cis`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_jitter(size = 2) +
  scale_x_continuous(limits = c(0,7), breaks = 1:7) +
  scale_y_continuous(limits = c(0,7), breaks = 1:7) +
  scale_colour_gradient(low = "grey90", high = "red") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. cis-pQTL associations \n (rare cis-IV for gene)",
       y = "No. cis-pQTL associations \n (common cis-IV for gene)")

p8.2.plotdat <- pqtls_assocs_joined_relaxed_truecis[,c("assoc_cis_genes", "IV_set", "n_trans", "assoc_trans_genes")] |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(n_trans, assoc_trans_genes),
    names_glue = "{IV_set}_{.value}"
  ) |>
  dplyr::filter(!(is.na(`exome:rare_n_trans`)) & !(is.na(`exome:common_n_trans`))) |> 
  rowwise() |>
  dplyr::mutate( # Calc prpoprtion of shared trans genes
    prop_shared = {
      s1 <- strsplit(`exome:rare_assoc_trans_genes`, ";")[[1]]
      s2 <- strsplit(`exome:common_assoc_trans_genes`, ";")[[1]]
      length(intersect(s1, s2)) / length(union(s1, s2))
    }) |> ungroup()
  
p8.2 <- p8.2.plotdat |> dplyr::arrange(prop_shared) |>
  ggplot(aes(x = `exome:rare_n_trans`, y = `exome:common_n_trans`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_point(size = 2) +
  scale_x_continuous(limits = c(0,350)) +
  scale_y_continuous(limits = c(0,350)) +
  scale_colour_gradient(low = "grey", high = "red") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. trans-pQTL associations \n (rare cis-IV for gene)",
       y = "No. trans-pQTL associations \n (common cis-IV for gene)")
  
p8.2_inset <- p8.2.plotdat |> dplyr::arrange(prop_shared) |>
  ggplot(aes(x = `exome:rare_n_trans`, y = `exome:common_n_trans`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_jitter(size = 1) +
  scale_x_continuous(limits = c(0,20)) +
  scale_y_continuous(limits = c(0,20)) +
  scale_colour_gradient(low = "grey", high = "red") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.background = element_rect(fill = "white", colour = "grey")) +
  labs(x = "No. trans-pQTL associations \n (rare cis-IV for gene)",
       y = "No. trans-pQTL associations \n (common cis-IV for gene)")

p8.2_join <- cowplot::ggdraw() + cowplot::draw_plot(p8.2) + cowplot::draw_plot(p8.2_inset, x = 0.5, y = 0.6, width = 0.95 - 0.5, height = 0.95 - 0.6)
  
# Correlation between cis and trans associations 

p8.2.2 <-  pqtls_assocs_joined_relaxed_truecis |>
  ggplot(aes(x = n_cis, y = n_trans, group = IV_set, colour = IV_set)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_log10() +
  scale_colour_manual(name = "", values = cols_IVsets) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. cis-pQTL associations",
       y = "No. trans-pQTL associations")

gridExtra::grid.arrange(p7.2.cis, p7.2.trans, p7.2, p8.2_join, p8.2.2, nrow = 1, widths = c(0.1,0.1,0.26,0.26,0.26), top = "Average pleiotropy of rare or common cis-IVs for a given gene (true cis-QTL shared genes)")

```

Are the higher numbers of cis-gene and trans-gene (within the same chromosome) associations observed for common IVs driven by variants that fall in gene dense regions (and is this more striking than for rare IVs)? Although this may represent proximal regulation, it could also be an indication of confounding through LD, particularly if rare and common vartiants show disparate trends:

```{r}
#| label: pleiotropy-vs-genedensity
#| fig.height: 4
#| fig.width: 15
#| cache: TRUE

# Annotate pQTLs (common IVs P<5e-8, rare IVs P<1e-4) with number of genes in proximity
count_genes_1mb <- function(geneinfo, variant_ids){
  out <- data.frame(variant = variant_ids,
                    chr = as.numeric(sub("-.*","",variant_ids)),
                    pos = as.numeric(sub("^.*-(.*)-.*-.*$","\\1",variant_ids))) |>
    dplyr::arrange(chr, pos)
  
  for (i in 1:nrow(out)){
    # 1Mb window about variant
    window_min <- out$pos[i] - 1000000
    window_max <- out$pos[i] + 1000000 
    
    # Extract genes in window (overlap okay)
    window_genes <- geneinfo |> 
      dplyr::filter(as.numeric(chr) == out$chr[i],
                    (as.numeric(start) >= window_min | as.numeric(stop) >= window_min) & (as.numeric(start) <= window_max | as.numeric(stop) <= window_max),
                    gene_biotype == "protein_coding")
    out$n_genes[i] = length(unique(window_genes$gene))
  }
  
  return(out[,c("variant","n_genes")])
}

annotate_highld <- function(variant_ids){
  out <- data.frame(variant = variant_ids,
                    chr = as.numeric(sub("-.*","",variant_ids)),
                    pos = as.numeric(sub("^.*-(.*)-.*-.*$","\\1",variant_ids))) |>
    dplyr::arrange(chr, pos)
  
  for (i in 1:nrow(out)){
    # High LD regions on chromosome
    ldset <- highld_regions |> dplyr::filter(V1 == paste0("chr", out$chr[i])) |> data.frame()
    inset <- any(apply(ldset, 1, function(x){out$pos[i] > as.numeric(x["V2"]) & out$pos[i] < as.numeric(x["V3"])}))
    out$longrange_ld[i] <- inset
  }
  
  return(out[,c("variant","longrange_ld")])
}

variant_ids <- c(pqtls_assocs_joined_relaxed_toppleiotropic|> dplyr::filter(IV_set == "exome:rare") |> dplyr::pull(variant),
                 sub(":","-",gsub("_","-",pqtls_assocs_joined_relaxed_toppleiotropic|> dplyr::filter(IV_set == "genome:common") |> dplyr::pull(variant)))) |> unique()

pqtls_assocs_joined_relaxed_toppleiotropic <- pqtls_assocs_joined_relaxed_toppleiotropic |>
  dplyr::mutate(variant_formatted = ifelse(IV_set == "genome:common", sub(":","-",gsub("_","-",variant)), variant))

## Annotate variants with genes in cis and flag those in high LD region
pqtls_assocs_joined_relaxed_toppleiotropic <- pqtls_assocs_joined_relaxed_toppleiotropic |> 
  left_join(count_genes_1mb(geneinfo = geneinfo, variant_ids = variant_ids), by = c("variant_formatted"="variant")) |>
  dplyr::rename("n_codinggenes_1mb" = "n_genes")

pqtls_assocs_joined_relaxed_toppleiotropic  <- pqtls_assocs_joined_relaxed_toppleiotropic|> 
  left_join(count_genes_1mb(geneinfo = geneinfo_olink, variant_ids = variant_ids), by = c("variant_formatted"="variant")) |>
  dplyr::rename("n_olinkgenes_1mb" = "n_genes")

pqtls_assocs_joined_relaxed_toppleiotropic  <- pqtls_assocs_joined_relaxed_toppleiotropic|> 
  left_join(annotate_highld(variant_ids = variant_ids), by = c("variant_formatted"="variant"))

# Plot
p9 <- pqtls_assocs_joined_relaxed_toppleiotropic |> 
  ggplot(aes(x = n_olinkgenes_1mb, y = n_cis, colour = IV_set)) +
  geom_hline(yintercept = c(0,5), col = "grey70") + 
  geom_jitter(width = 0.2, height = 0, size = 2) +
  facet_grid(IV_set ~ .) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "Number of O'link protein-coding genes within 1Mb of instrument", 
       y = "Number of pQTL associations with \n O'link proteins within 1Mb",
       title = "IVs in genes with both common and rare WES instruments")

p10 <- pqtls_assocs_joined_relaxed_toppleiotropic |> 
  ggplot(aes(x = n_codinggenes_1mb, y = n_cis, colour = IV_set)) +
  geom_hline(yintercept = c(0,5), col = "grey30") + 
  geom_jitter(width = 0.2, height = 0, size = 2) +
  facet_grid(IV_set ~ .) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "Number of protein-coding genes within 1Mb of instrument", 
       y = "Number of pQTL associations with \n O'link proteins within 1Mb",
       title = "IVs in genes with both common and rare WES instruments")

p11 <- pqtls_assocs_joined_relaxed_toppleiotropic |> 
  ggplot(aes(x = longrange_ld, y = n_cis)) +
  facet_grid(.~IV_set) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  #scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "IV in long-range LD region", y = "Number of cis-pQTL associations\n per cis-IV")

p11.2 <- pqtls_assocs_joined_relaxed_toppleiotropic |> 
  ggplot(aes(x = longrange_ld, y = n_trans + n_cis)) +
  facet_grid(.~IV_set) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "IV in long-range LD region", y = "Number of pQTL associations\n per cis-IV")

gridExtra::grid.arrange(p9, p10, p11, p11.2, nrow = 1, widths = c(0.35,0.35,0.15,0.15))

```

How often do genes instrumented with either common or rare variants show consistant trans effects? We may anticipate that true vertical pleiotropy is detectable through both instrumentation strategies, whereas problematic trans effects, arising due to LD, will result in associations seen when using common variant instrumentation alone...

What proportion of cis and trans associations are shared between rare and common instrumentation strategies (on the gene level)?

```{r}
#| label: shared-signals
#| eval: FALSE

# Define:
# cis (< +- 1mb), cis_chr (on the same chr), trans (> +- 1mb), trans_genome (on another chr)
# cis/trans signals shared by rare and common varinat instrumentaiton of genes (WES IVs only)

shared_signals <- dplyr::full_join(
  lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned |>
    dplyr::select(protein_target, proximal_gene, cis_trans, protein_target_chr, proximal_gene_chr) |> unique() |>
    dplyr::mutate("exome:rare" = TRUE),
  lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |>
    dplyr::select(protein_target, proximal_gene, cis_trans, protein_target_chr, proximal_gene_chr) |> unique() |>
    dplyr::mutate("exome:common" = TRUE), by = c("protein_target", "proximal_gene", "cis_trans", "protein_target_chr", "proximal_gene_chr")) |>
  dplyr::filter(!is.na(proximal_gene_chr),
                protein_target != gsub("'","",proximal_gene)) |>
  dplyr::mutate(
    shared = (`exome:rare` == `exome:common`),
    cis_1mb = (cis_trans == "cis"),
    cis_chr = (proximal_gene_chr == protein_target_chr),
    trans_chr = (cis_trans == "trans" & proximal_gene_chr == protein_target_chr),
    trans_genome = (proximal_gene_chr != protein_target_chr))

shared_signals$shared[is.na(shared_signals$shared)] <- FALSE

## Plot this
# shared_signals |> ggplot(aes(x = trans_genome, fill = shared)) +
#   geom_bar(position = "fill")

## Loop through genes containg both common and rare instruments and quantify shared trans signals

shared_transsignals <- data.frame(proximal_gene = shared_genes, 
                                  n_trans_total = NA,
                                  n_trans_shared = NA,
                                  n_trans_rare = NA,
                                  n_trans_common = NA,
                                  genes_shared = NA,
                                  genes_rare = NA,
                                  genes_common = NA)
rownames(shared_transsignals) <- shared_genes

for (gene in shared_genes){
  trans_rare <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned |> 
    dplyr::filter(proximal_gene == gene) |>
    dplyr::pull(protein_target) |> unique()
  trans_rare <- trans_rare[trans_rare != gsub("'","",gene)]
  
  trans_common <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |> 
    dplyr::filter(proximal_gene == gene) |>
    dplyr::pull(protein_target) |> unique()
  trans_common <- trans_common[trans_common != gsub("'","",gene)]

  shared_transsignals[gene, "n_trans_total"] <- length(unique(c(trans_rare, trans_common)))
  shared_transsignals[gene, "n_trans_shared"] <- length(intersect(trans_rare, trans_common))
  shared_transsignals[gene, "n_trans_rare"] <- length(setdiff(trans_rare, trans_common))
  shared_transsignals[gene, "n_trans_common"] <- length(setdiff(trans_common, trans_rare))
  shared_transsignals[gene, "genes_shared"] <- paste(intersect(trans_rare, trans_common), collapse = ";")
  shared_transsignals[gene, "genes_rare"] <- paste(setdiff(trans_rare, trans_common), collapse = ";")
  shared_transsignals[gene, "genes_common"] <- paste(setdiff(trans_common, trans_rare), collapse = ";")
}

# Bind number of gene within +- 1Mb
shared_transsignals <- shared_transsignals |>
  left_join(pqtls_assocs_joined_relaxed[,c("proximal_gene", "n_codinggenes_1mb")] |>
              dplyr::arrange(desc(n_codinggenes_1mb)) |> 
              dplyr::filter(duplicated(proximal_gene) == FALSE), by = "proximal_gene")

# Proportion of the total trans hits seen with variants in a gene which are detected using both common and rare instruments?
shared_transsignals$prop_shared = shared_transsignals$n_trans_shared/shared_transsignals$n_trans_total

# shared_transsignals |> dplyr::filter(n_trans_total != 0) |>
#   ggplot(aes(x = n_codinggenes_1mb, y = prop_shared)) +
#   geom_point()

```

### IV heterogenity

Heterogenity in SNP effects can be a result of variants acting within differing biological pathways to infuence an outcome (pleiotropy), and violates the third IV assumption (an instrument only influences an outcome via its effect on an exposure). Do rare and common, cis and trans IVs exhibit differing heterogenity in causal effect estimates? When rare and common instruments are combined into one MR analysis, is there a relationship between exposure effect and SNP heterogenity (definied as the deviation between the wald ratio for an IV and the IVW estimated from the metanalysis of all IVs)?...

```{r}
#| label: IV-heterogenity
#| fig.height: 20
#| fig.width: 5

# Read in results of mr_heterogenity() for common and rare variant IV sets
files_het <- list.files(res_dir, pattern = "heterogenity", full.names = T)
het <- do.call("rbind", lapply(files_het, function(x){data.table::fread(x)}))

het_set <- het |> 
  dplyr::filter(method == "Inverse variance weighted" & n_snps > 5) |>
  dplyr::mutate(IV_set = factor(IVsets[variant_set], levels = variant_IVsets))

het_set|> 
  ggplot(aes(x = variant_type, y = Q)) +
  facet_grid(outcome ~ IV_set) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + 
  labs(x = "", y = "Q statistic", title = "Heterogenity in causal estimates (proteins with >5 IVs)")

# Restrict to the same proteins across instrument sets (sensitivity analysis)

# comp_hetcis <- het_set |> dplyr::filter(variant_type == "cis") |> with(pairwise.wilcox.test(Q, IV_set, p.adjust.method = "none"))
# comp_hetcis$p.value
# comp_hettrans <- het_set |> dplyr::filter(variant_type == "trans") |> with(pairwise.wilcox.test(Q, IV_set, p.adjust.method = "none"))
# comp_hettrans$p.value
```

## Causal effect estimates

### Comparison of causal effects

**Significant causal effects detected across instrument types**

```{r}
#| label: full-results
#| fig.width: 12
#| fig.height: 33
#| cache: TRUE

# P-value threshold for cis-MR (multiple testing correction for proteins instrumentable with at least 3 IVs in cis)
alpha_cis3 <- 0.05/max(avail_IVs_relaxed |> dplyr::filter(cis_trans == "cis") |> pull(n_prots_3IV)) # 5x10^-5

# Plot "manhattan" for protein causal effects on a given outcome, restricting to IV sets of at least 3 variants (masks can have 1)
plot_fullresults <- function(results_MR, outcome, nsnps = 3, cistrans = "cis"){
  df <- do.call("rbind", lapply(results_MR[[outcome]], function(x){
    dat <- x |> 
      dplyr::filter(method %in% c("Wald ratio", "Inverse variance weighted"),
                    cis_trans == cistrans,
                    (IV_set %in% names(variant_IVsets) & nsnp >= nsnps) | !(IV_set %in% names(variant_IVsets)),
                    IV_set %in% names(variant_IVsets))
  }))
  # Format for plotting
  df <- df |>
    dplyr::arrange(pval) |>
    dplyr::mutate(
      set = IVsets[IV_set],
      col = factor(ifelse(pval <= 0.01, set, NA), levels = IVsets),
      label = ifelse(pval <= alpha, exposure, NA))
    #dplyr::mutate(label = ifelse(duplicated(label), NA, label))
  
  df |> dplyr::arrange(desc(pval)) |>
    ggplot(aes(x = exposure, y = -log10(pval))) +
    geom_hline(yintercept = -log10(alpha_cis3), col = "grey50", lty = 2) +
    geom_point(aes(colour = col), size = 2.5) +
    ggrepel::geom_text_repel(aes(label = label)) +
    scale_colour_manual(name = "", values = cols_IVsets, na.value = "grey90") +
    theme_minimal() +
    theme(legend.position = "bottom",
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 10),
          axis.title = element_text(size = 12),
          panel.grid = element_blank(),
          plot.title = element_text(hjust = 0.5, size = 14),
          legend.text = element_text(size = 12)) +
    guides(colour = guide_legend(nrow = 1)) +
    labs(x = "Protein exposure", y = "IVW/WR -log10(P)", title = outcome)
}

gridExtra::grid.arrange(grobs = 
  lapply(X = names(res_MR_relaxed), FUN = plot_fullresults, results_MR = res_MR_relaxed, nsnps = 3, cistrans = "cis"), nrow = 11)


```

How many new highly significant causal associations are detected with rare variants or gene based masks that a. cannot be instrumented in cis with common variants or b. show no causal effect when using common-IVs (suggesting rare variants perturb the gene in a more relevant fashion)?:

```{r}
#| label: causal-effect-overlap
#| cache: TRUE

# Pull significant MR results (at p<5e-05 for comparison of cis IV sets)
# >=3 IVs for variant-based instruments, 1 sufficient for masks (exclude masks for now?)

sig_MR_relaxed <- lapply(res_MR_relaxed, function(x){
  do.call("rbind", lapply(x, function(y){
    dat <- y |> 
      dplyr::filter(method %in% c("Wald ratio", "Inverse variance weighted"),
                    cis_trans == "cis",
                    (IV_set %in% names(variant_IVsets) & nsnp >= 3) | !(IV_set %in% names(variant_IVsets)),
                    IV_set %in% names(paper_IVsets),
                    pval <= alpha_cis3)})) |> # multiple testing correction
    dplyr::mutate(IVset = IVsets[IV_set])})

# Upset plot of overlap in support for exposure --> outcome association
upset_sig_cis <- lapply(sig_MR_relaxed, function(x){
  list("mask:pLOF|miss" = x |> dplyr::filter(IVset %in% c("mask:pLOF", "mask:miss", "mask:pLOF|miss")) |>  dplyr::pull(pair) |> unique(),
       "IV:rare" = x |> dplyr::filter(IVset %in% c("exome:ultrarare","exome:rare")) |>  dplyr::pull(pair) |> unique(),
       "IV:common" = x |> dplyr::filter(IVset %in% c("exome:common", "genome:common")) |>  dplyr::pull(pair) |> unique())})
upset_sig_cis <- purrr::map(names(upset_sig_cis[[1]]), ~ purrr::flatten_chr(purrr::map(upset_sig_cis, .x)))
names(upset_sig_cis) <- c("mask:pLOF|miss","variant:rare","variant:common")

# Pull significant MR results (at p<0.01 for comparison of cis IV sets)
# >=3 IVs for variant-based instruments, 1 sufficient for masks (exclude masks for now?)

sig_MR_relaxed_0.01 <- lapply(res_MR_relaxed, function(x){
  do.call("rbind", lapply(x, function(y){
    dat <- y |> 
      dplyr::filter(method %in% c("Wald ratio", "Inverse variance weighted"),
                    cis_trans == "cis",
                    (IV_set %in% names(variant_IVsets) & nsnp >= 3) | !(IV_set %in% names(variant_IVsets)),
                    IV_set %in% names(paper_IVsets),
                    pval <= 0.01)})) |> # multiple testing correction
    dplyr::mutate(IVset = IVsets[IV_set])})

# Upset plot of overlap in support for exposure --> outcome association
upset_sig_cis_0.01 <- lapply(sig_MR_relaxed_0.01, function(x){
  list("mask:pLOF|miss" = x |> dplyr::filter(IVset %in% c("mask:pLOF", "mask:miss", "mask:pLOF|miss")) |>  dplyr::pull(pair) |> unique(),
       "IV:rare" = x |> dplyr::filter(IVset %in% c("exome:ultrarare","exome:rare")) |>  dplyr::pull(pair) |> unique(),
       "IV:common" = x |> dplyr::filter(IVset %in% c("exome:common", "genome:common")) |>  dplyr::pull(pair) |> unique())})
upset_sig_cis_0.01 <- purrr::map(names(upset_sig_cis_0.01[[1]]), ~ purrr::flatten_chr(purrr::map(upset_sig_cis_0.01, .x)))
names(upset_sig_cis_0.01) <- c("mask:pLOF|miss","variant:rare","variant:common")

## Exposure:outcome pairs with sufficient cis-IVs to test
common_pairs <- num_IVs_relaxed |> dplyr::filter(cis_trans == "cis", grepl("common", IV_set), value >= 3) |> pull(name) |> unique()
rare_pairs <- num_IVs_relaxed |> dplyr::filter(cis_trans == "cis", grepl("rare_pruned", IV_set), value >= 3) |> pull(name) |> unique()

## Overlap in sig exposure-outcome pairs (P<5e-5)
g1 <- ggvenn::ggvenn(upset_sig_cis[c(2,3)], fill_color = c("steelblue1","orange"), set_name_size = 5) + 
  labs(title = "Exposure:outcome pair \nP(IVW)<5e-5") +
  theme(plot.title = element_text(hjust = 0.5))
### Of rare unique:
### Lack of common IVs
table(g1$data |> dplyr::filter(`variant:rare` == TRUE & `variant:common` == FALSE) |> pull(`_key`) %in% common_pairs)
### Significant at lower threshold?
table(g1$data |>
  dplyr::filter(`variant:rare` == TRUE & `variant:common` == FALSE) |>
  dplyr::filter(`_key` %in% common_pairs) |>
  pull(`_key`) %in% upset_sig_cis_0.01$`variant:common`) # 1 yes, 10 no

### Of common unique:
### Lack of rare IVs
table(g1$data |> dplyr::filter(`variant:rare` == FALSE & `variant:common` == TRUE) |> pull(`_key`) %in% rare_pairs)
### Significant at lower threshold?
table(g1$data |>
  dplyr::filter(`variant:rare` == FALSE & `variant:common` == TRUE) |>
  dplyr::filter(`_key` %in% rare_pairs) |>
  pull(`_key`) %in% upset_sig_cis_0.01$`variant:rare`) # 18 yes, 81 no
### Of shared:
### Concordant vs discordant direction of effect
shared_5e5 <- g1$data |> dplyr::filter(`variant:rare` == TRUE & `variant:common` == TRUE) |> pull(`_key`)
concordant_5e5 <- do.call("rbind",lapply(sig_MR_relaxed, function(x){x |> dplyr::filter(pair %in% shared_5e5)})) |>
  dplyr::filter(nsnp >= 3) |>
  group_by(pair) |>
  summarise(length(unique(sign(b))))
## Concordant sign?
table(concordant_5e5[,2]==1)

## Overlap in sig exposure-outcome pairs (P<5e-5)
g1_all <- ggvenn::ggvenn(upset_sig_cis, fill_color = c("mediumorchid","steelblue1","orange"), set_name_size = 5) + 
  labs(title = "Exposure:outcome pair \nP(IVW)<5e-5") +
  theme(plot.title = element_text(hjust = 0.5))

## Overlap in sig exposure-outcome pairs (P<0.01)
g2 <- ggvenn::ggvenn(upset_sig_cis_0.01[c(2,3)], fill_color = c("steelblue1","orange"), set_name_size = 5) +
  labs(title = "Exposure:outcome pair \nP(IVW)<0.01") +
  theme(plot.title = element_text(hjust = 0.5))
### Of rare unique:
### Lack of common IVs
table(g2$data |> dplyr::filter(`variant:rare` == TRUE & `variant:common` == FALSE) |> pull(`_key`) %in% common_pairs)
### Which pairs are these?:
# g2$data |>
#   dplyr::filter(`variant:rare` == TRUE & `variant:common` == FALSE) |>
#   dplyr::mutate(commonIVs_exist = `_key` %in% common_pairs)
### Of common unique:
### Lack of rare IVs
table(g2$data |> dplyr::filter(`variant:rare` == FALSE & `variant:common` == TRUE) |> pull(`_key`) %in% rare_pairs)
### Which pairs are these?:
# g2$data |>
#   dplyr::filter(`variant:rare` == FALSE & `variant:common` == TRUE) |>
#   dplyr::mutate(rareIVs_exist = `_key` %in% rare_pairs)
### Of shared:
### Concordant vs discordant direction of effect
shared_0.01 <- g2$data |> dplyr::filter(`variant:rare` == TRUE & `variant:common` == TRUE) |> pull(`_key`)
concordant_0.01 <- do.call("rbind",lapply(sig_MR_relaxed_0.01, function(x){x |> dplyr::filter(pair %in% shared_0.01)})) |>
  dplyr::filter(nsnp >= 3) |>
  group_by(pair) |>
  summarise(length(unique(sign(b))))
## Concordant sign?
table(concordant_0.01[,2]==1)

# 38 causal protein --> outcome relationships when instrumenting using rare cis variants (p<5x10^-5)
# 31 undetected when using common cis-IVs at this threshold
## Of these 31:
### 20 lacked sufficient common-IVs and were not tested
### 1 are significant using common-IVs at a relaxed threshold 0.01>p>5x10^-5
### 10 were tested but using common-IVs but no causal associaton was detected

# 131 causal protein --> outcome relationships when instrumenteing using common cis variants
# 124 undetected when using rare cis-IVs at this threshold
### 25 lacked sufficient rare-IVs and were not teted
### 18 are significant using rare-IVs at a relaxed threshold 0.01>p>5x10^-5
### 81 were tested but using rare-IVs but no causal associaton was detected

```

```{r}
#| label: venns
#| fig-width: 10
#| fig-height: 7

gridExtra::grid.arrange(grobs = list(g1,g2), nrow = 1)
gridExtra::grid.arrange(grobs = list(g1_all), nrow = 1)
```

### Example results

Plot effect estimates across IV sets for a given exposure-outcome pair

```{r}
#| label: plot-mr-results

plot_results <- function(exposure, outcome){
  
  pair <- paste(exposure, outcome, sep = "_")
  # Individual SNP results
  dat_snps <- do.call("rbind",lapply(harmonised_snps_relaxed[[outcome]][[pair]][names(paper_IVsets)], function(x){
    x |> dplyr::filter(cis_trans == "cis", mr_keep == TRUE) |>
      dplyr::select(SNP, beta.exposure, beta.outcome, pval.exposure, pval.outcome, se.exposure, se.outcome)})) |>
    tibble::rownames_to_column("IV_set") |>
    dplyr::mutate(IV_set = factor(IVsets[sub("\\..*","",IV_set)], levels = IVsets)) |>
    dplyr::filter(IV_set %in% paper_IVsets) |>
    dplyr::arrange(desc(IV_set))
  
  # MR estimates
  dat_res <- res_MR_relaxed[[outcome]][[pair]] |> 
    dplyr::filter(cis_trans == "cis", method %in% c("Wald ratio", "Inverse variance weighted"), IV_set %in% names(paper_IVsets)) |>
    dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets))
  
  ## IVW plot
  p_slope <- dat_snps |> dplyr::filter(IV_set %in% variant_IVsets) |> 
    ggplot(aes(x = beta.exposure, y = beta.outcome, colour = IV_set)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_errorbar(aes(xmin = beta.exposure - 1.96*se.exposure, xmax = beta.exposure + 1.96*se.exposure), height = 0, alpha = 0.3) +
    geom_errorbar(aes(ymin = beta.outcome - 1.96*se.outcome, ymax = beta.outcome + 1.96*se.outcome), height = 0, alpha = 0.3) +
    geom_abline(data = dat_res |> dplyr::filter(nsnp > 1), aes(slope = b, intercept = 0, colour = IV_set)) +
    geom_point(size = 2) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5)) +
    guides(colour = guide_legend(nrow = 2)) +
    labs(x = expression(beta["exposure"]), y = expression(beta["outcome"]), title = paste("Exposure:", exposure, "Outcome:", outcome))
    
  ## Forest plot
  p_forest_vars <- dat_res |> dplyr::filter(IV_set %in% variant_IVsets) |>
    ggplot(aes(x = b, y = IV_set, colour = IV_set)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(aes(size = nsnp)) +
    geom_errorbarh(aes(xmin = b-1.96*se, xmax = b+1.96*se, group = cis_trans), height = 0.4) +
    scale_colour_manual(values = cols_IVsets) +
    scale_size_area() +
    theme_minimal() + guides(colour = "none") +
    theme(plot.title = element_text(hjust = 0.5), 
          axis.text.x = element_text(size = 7),
          legend.position = "right") +
    labs(x = "IVW", y = "", size = "n(IVs)")
  
  p_forest_mask <- dat_res |> dplyr::filter(!(IV_set %in% variant_IVsets)) |>
    ggplot(aes(x = b, y = IV_set, colour = IV_set)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    geom_errorbarh(aes(xmin = b-1.96*se, xmax = b+1.96*se, group = cis_trans), height = 0.4) +
    scale_colour_manual(values = cols_IVsets) +
    scale_size_area() +
    theme_minimal() + guides(colour = "none") +
    theme(plot.title = element_text(hjust = 0.5), 
          axis.text.x = element_text(size = 7),
          legend.position = "right") +
    labs(x = "WR", y = "", size = "n(IVs)")
  
  cowplot::ggdraw() + cowplot::draw_plot(p_slope,0,0,0.5,1) + 
    cowplot::draw_plot(p_forest_vars,0.5,0.2,0.5,0.4) +
    cowplot::draw_plot(p_forest_mask,0.5,0.6,0.4,0.2)
}
```

```{r}
#| label: plot-mr-results-explore
#| include: FALSE
#| eval: FALSE

## Plot, common variant associations lacking rare evidence
common_only <- g2$data |> dplyr::filter(`variant:rare` == FALSE, `variant:common` == TRUE) |> dplyr::pull(`_key`)
common_only <- common_only[common_only %in% upset_sig_cis$`variant:common`] # Significant w FDR correction

apply(as.data.frame(do.call("rbind",(strsplit(common_only, "_")))), MARGIN = 1,
                    FUN = function(x){plot_results(exposure = x["V1"], outcome = x["V2"])})

# CLMP-SBP
# HDDC2-RBCcount (four rare variants in gene)
# MPO-CorpVol

## Plot, rare variant associations lacking common evidence
rare_only <- g2$data |> dplyr::filter(`variant:rare` == TRUE, `variant:common` == FALSE) |> dplyr::pull(`_key`)
rare_only <- rare_only[rare_only %in% upset_sig_cis$`variant:rare`] # Significant w FDR correction

apply(as.data.frame(do.call("rbind",(strsplit(rare_only, "_")))), MARGIN = 1,
                    FUN = function(x){plot_results(exposure = x["V1"], outcome = x["V2"])})

# Shared
shared <- g1$data |> dplyr::filter(`variant:rare` == TRUE, `variant:common` == TRUE) |> dplyr::pull(`_key`)

apply(as.data.frame(do.call("rbind",(strsplit(shared, "_")))), MARGIN = 1,
                    FUN = function(x){plot_results(exposure = x["V1"], outcome = x["V2"])})
```

```{r}
#| label: plot-mr-results-examples
#| fig.width: 14
#| fig.height: 8.5

## Example of concordant results from instrument sets (PRSS53 in BMI)
#plot_results(exposure = "PRSS53", outcome = "BMI")
#ex1 <- plot_results(exposure = "AMPD3", outcome = "HbA1c")
ex1 <- plot_results(exposure = "PCSK9", outcome = "LDL") ## Lack of common IVs

## Example of rare eviedence only 
#plot_results(exposure = "GMPR", outcome = "RBCcount") ## Lack of common IVs
ex2 <- plot_results(exposure = "SMPD1", outcome = "SBP")
ex3 <- plot_results(exposure = "DBH", outcome = "SBP")
## Example of common evidence only
#ex4 <- plot_results(exposure = "PRSS53", outcome = "LDL")
#ex4 <- plot_results(exposure = "CDSN", outcome = "PlatVol")
ex4 <- plot_results(exposure = "HDDC2", outcome = "RBCcount")

gridExtra::grid.arrange(grobs = list(ex1,ex2,ex3,ex4), nrow = 2)
```

**Forest plots of wald ratios**

```{r}
#| label: plot-IV-details

# Pull IVs used for MR (across variant sets), derive wald ratios and annotate exome variants with amino acid change
annotate_ivs <- function(outcome, exposure, snpset){
  # Pull cis variants
  ivs_exposure <- lapply(snpset[[outcome]][[paste(exposure,outcome,sep = "_")]], function(x){x |> dplyr::filter(cis_trans == "cis")})
  ivs_exposure <- ivs_exposure[unlist(lapply(ivs_exposure, nrow)) > 0]
  
  # Derive wald ratios for all cis-IVs, all frequencies
  keep_cols <- c("SNP","variant","chr.exposure","pos.exposure", 
                 "effect_allele.exposure", "other_allele.exposure",
                 "beta.exposure","beta.outcome",
                 "se.exposure","se.outcome","pval.exposure","pval.outcome",
                 "eaf.exposure","cis_trans","gene.outcome", "ANNOTATON", "mr_keep")
  
  ## Preformat data
  ivs_exposure_all <- do.call("bind_rows", mapply(FUN = function(x,y){
    keep_cols_df <- intersect(keep_cols, colnames(x))
    x <- x[,keep_cols_df]
    x$IV_set <- y
    return(x)}, x = ivs_exposure[names(variant_IVsets)], y = variant_IVsets, SIMPLIFY = FALSE)) |>
    dplyr::mutate(IV_set = factor(IV_set, levels = variant_IVsets[c(1,2,4,3)]),
                  chr.exposure = ifelse(grepl("rs", SNP), sub(":.*","",variant), chr.exposure),
                  pos.exposure = ifelse(grepl("rs", SNP), sub(".*:(.*)_.*_.*","\\1",variant), pos.exposure),
                  SNP = toupper(ifelse(grepl("rs", SNP), variant, SNP))) |>
    dplyr::arrange(pos.exposure, IV_set) |>
    dplyr::filter(duplicated(SNP) == FALSE, mr_keep == TRUE, !is.na(SNP))
  
  ## WRs
  wrs_all <- t(apply(ivs_exposure_all, MARGIN = 1, function(x){
    TwoSampleMR::mr_wald_ratio(b_exp = as.numeric(x["beta.exposure"]),
                               b_out = as.numeric(x["beta.outcome"]), 
                               se_exp = as.numeric(x["se.exposure"]),
                               se_out = as.numeric(x["se.outcome"])) |> unlist()
  }))
    
  colnames(wrs_all) <- paste0("wald_ratio_", colnames(wrs_all))
  wrs_all <- cbind(ivs_exposure_all, wrs_all)
  
  # Annotate exome variants
  wrs_all <- left_join(wrs_all, 
                       unique(all_cisIVs_relaxed[,c("SNP_ID",
                                             "effect_allele.exposure","other_allele.exposure","GENE","Exon rank",
                                             "cDNA change","Amino acid change","Variant_group","effect_allele.vep","other_allele.vep","ea_matches_annot")]),
                       by = c("SNP" = "SNP_ID", "effect_allele.exposure", "other_allele.exposure")) |>
    dplyr::mutate(exposure = exposure, outcome = outcome)
  
  return(wrs_all)
}

# Plot
# ivs = harmonised SNP data with appendend wald ratios and variant details
# orientate = TRUE (orientate to expression increasing allele)
plot_forest_snps <- function(iv_effects, orientate){
  iv_effects$SNP_name = ifelse(nchar(iv_effects$GENE) > 1,
                           paste0("chr",sub("_.*","",iv_effects$SNP)," (",iv_effects$GENE,") \n", iv_effects$`Amino acid change`),
                           paste0("chr",sub("_.*","",iv_effects$SNP)))
  # Flag varaints for which the annotated protein change does not match the effect direction
  iv_effects$SNP_name = ifelse(nchar(iv_effects$`Amino acid change`) > 1 & iv_effects$ea_matches_annot == FALSE, paste(iv_effects$SNP_name, "#"), iv_effects$SNP_name)
    
   if(orientate == TRUE){
     flip_sign <- ifelse(iv_effects$beta.exposure > 0, TRUE, FALSE)
     iv_effects$beta.exposure[flip_sign] <- iv_effects$beta.exposure[flip_sign]*-1
     iv_effects$beta.outcome[flip_sign] <- iv_effects$beta.outcome[flip_sign]*-1
     
     # Flag varaints for which the annotated protein change does not match the effect direction
     iv_effects$SNP_name = ifelse(nchar(iv_effects$`Amino acid change`) > 1 & iv_effects$ea_matches_annot == flip_sign, paste(iv_effects$SNP_name, "#"), iv_effects$SNP_name)
   }
  
  ntests <- nrow(iv_effects)
  exposure <- unique(iv_effects$exposure)
  outcome <- unique(iv_effects$outcome)
  
  # Draw gene
  chr <- paste0("chr",unique(iv_effects$chr.exposure))
  regionstart <- as.numeric(min(iv_effects$pos.exposure))
  regionstop <- as.numeric(max(iv_effects$pos.exposure))
  
  marker_positions <- data.frame(chr = chr,
                                 pos = as.numeric(iv_effects$pos.exposure),
                                 IV_set = iv_effects$IV_set)
  
  gf <- plot_genomefeatures_gg(chr = chr, regionstart = regionstart, regionstop = regionstop, marker_positions = marker_positions)
  gf <- patchwork::patchworkGrob(gf)
  
  # Forest plots
  f1 <- iv_effects |>
    dplyr::mutate(col_points = ifelse(pval.exposure < 1, as.character(IV_set), NA)) |>
    ggplot(aes(x = SNP_name, y = beta.exposure, colour = col_points)) +
    geom_hline(yintercept = 0, lty = 2) +
    geom_point(size = 2.5) +
    geom_errorbar(aes(ymin = beta.exposure - 1.96*se.exposure,
                      ymax = beta.exposure + 1.96*se.exposure), width = 0.3) +
    scale_colour_manual(values = cols_IVsets, na.value = "grey90") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), legend.position = "none") +
    labs(x = "", y = "Bx")
  
  f2 <- iv_effects |>
    dplyr::mutate(col_points = ifelse(pval.outcome < 1, as.character(IV_set), NA)) |>
    ggplot(aes(x = SNP_name, y = beta.outcome, colour = col_points)) +
    geom_hline(yintercept = 0, lty = 2) +
    geom_point(size = 2.5) +
    geom_errorbar(aes(ymin = beta.outcome - 1.96*se.outcome,
                      ymax = beta.outcome + 1.96*se.outcome), width = 0.3) +
    scale_colour_manual(values = cols_IVsets, na.value = "grey90") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), legend.position = "none") +
    labs(x = "", y = "By")
  
  f3 <- iv_effects |>
    dplyr::mutate(col_points = as.character(IV_set)) |>
    ggplot(aes(x = SNP_name, y = wald_ratio_b, colour = col_points)) +
    geom_hline(yintercept = 0, lty = 2) +
    geom_point(size = 2.5) +
    geom_errorbar(aes(ymin = wald_ratio_b - 1.96*wald_ratio_se,
                      ymax = wald_ratio_b + 1.96*wald_ratio_se), width = 0.3) +
    scale_colour_manual(values = cols_IVsets, na.value = "grey90") +
    theme_minimal() +
    theme(axis.text.x =element_blank(),
          legend.position = "none") +
    labs(x = "", y = "By/Bx")
  
  # Variant annotation
  v <- iv_effects |>
    ggplot(aes(x = SNP_name, y = 1)) +
    geom_tile(aes(fill = Variant_group), colour = "white", lwd = 2) +
    theme_minimal() +
    scale_fill_manual(values = cols_consequences) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 75, hjust = 1),
          axis.text.y = element_text(colour = "white"),
          axis.title.y = element_text(colour = "white")) +
    labs(x = "")
  
  #gridExtra::grid.arrange(gf, f1, f2, f3, v, heights = c(0.2, 0.2, 0.2, 0.2, 0.2), top = paste0(exposure, " IVs ", exposure, ":", outcome))
  aligned <- cowplot::align_plots(gf, f1, f2, f3, v, align = "h", axis = "lr")
  
  title <- cowplot::ggdraw() +
    cowplot::draw_label(paste0(exposure, " IVs ", exposure, ":", outcome),fontface = "bold")
  
  cowplot::plot_grid(
    title,
    cowplot::plot_grid(aligned[[1]], aligned[[2]], aligned[[3]], aligned[[4]], aligned[[5]],
      ncol = 1,
      rel_heights = c(0.2, 0.1, 0.1, 0.1, 0.2)),
    ncol = 1,
    rel_heights = c(0.05, 1)
  )
}
```

**Pleitropy profiles**

```{r}
#| label: plot-pleiotropy signature
#| 
plot_pleiotropy <- function(iv_effects){
  # Variants
  iv_effects$SNP <- gsub("_","-",sub(":","-",iv_effects$SNP))
  vars <- split(iv_effects$SNP, f = iv_effects$IV_set)
  vars_ivsets <- names(vars)
  # Full pQTL association lookup
  lookup_ivsets <- lookup_pQTLs_MRset[names(IVsets[match(vars_ivsets, IVsets)])]
  # Pull all pQTL associations for vars in list
  pqtls_vars <- do.call("rbind", mapply(FUN = function(x,y){x |> dplyr::filter(gsub("_","-",sub(":","-",variant)) %in% y)}, 
                       x = lookup_ivsets, y = vars, SIMPLIFY = F))
  pqtls_vars$variant <- factor(gsub("_","-",sub(":","-",pqtls_vars$variant)), levels = iv_effects$SNP)
  pqtls_vars$pvalue <- ifelse(pqtls_vars$pvalue == 0, 1e-300, pqtls_vars$pvalue)
  
  plot_hits <- ggplot(pqtls_vars, aes(x = variant, y = protein_target)) +
    geom_tile(aes(fill = -log10(pvalue))) +
    scale_fill_gradient(low = "steelblue1", high = "tomato") +
    theme_minimal() +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 75, hjust = 1),
          axis.title = element_blank())
  
  plot_hits
}

```


```{r}
#| label: AMPD3
#| fig.width: 5
#| fig.height: 9

plot_forest_snps(iv_effects = annotate_ivs(outcome = "HbA1c", exposure = "AMPD3", snpset = harmonised_snps_relaxed), orientate = FALSE)
plot_pleiotropy(iv_effects = annotate_ivs(outcome = "HbA1c", exposure = "AMPD3", snpset = harmonised_snps_relaxed))
```

```{r}
#| label: PCSK9
#| fig.width: 4
#| fig.height: 9

plot_forest_snps(iv_effects = annotate_ivs(outcome = "LDL", exposure = "PCSK9", snpset = harmonised_snps_relaxed), orientate = TRUE)
plot_pleiotropy(iv_effects = annotate_ivs(outcome = "LDL", exposure = "PCSK9", snpset = harmonised_snps_relaxed))
```

```{r}
#| label: SMPD1
#| fig.width: 6
#| fig.height: 11

plot_forest_snps(iv_effects = annotate_ivs(outcome = "SBP", exposure = "SMPD1", snpset = harmonised_snps_relaxed), orientate = FALSE)
plot_pleiotropy(iv_effects = annotate_ivs(outcome = "SBP", exposure = "SMPD1", snpset = harmonised_snps_relaxed))
```

```{r}
#| label: HDDC2
#| fig.width: 4
#| fig.height: 9

plot_forest_snps(iv_effects = annotate_ivs(outcome = "RBCcount", exposure = "HDDC2", snpset = harmonised_snps_relaxed), orientate = TRUE)
plot_pleiotropy(iv_effects = annotate_ivs(outcome = "RBCcount", exposure = "HDDC2", snpset = harmonised_snps_relaxed))

HDDC2_cIVs <- harmonised_snps_relaxed$RBCcount$HDDC2_RBCcount$sun_gwas_variant_common_clumped |> dplyr::filter(cis_trans == "cis")

# GPMAP pleiotropy profile of common variants:
coloc_count <- sapply(HDDC2_cIVs$SNP, function(x){gpmapr::search_gpmap(x)}, simplify = F)
# Two of six variants fall in coloc groups:
# rs3799738 (6:125232804 G/A 8131994)n - search for proxy
snp1 <- gpmapr::variant(8132004)$coloc_groups
# rs45438294 (6:125373298 C/T 8132435) - search for proxy
snp2 <- gpmapr::variant(8132437)$coloc_groups

```

```{r}
#| label: DBH
#| fig.width: 14
#| fig.height: 14

plot_forest_snps(iv_effects = annotate_ivs(outcome = "SBP", exposure = "DBH", snpset = harmonised_snps_relaxed), orientate = FALSE)
plot_pleiotropy(iv_effects = annotate_ivs(outcome = "SBP", exposure = "DBH", snpset = harmonised_snps_relaxed))
```

**Comparison of IVW effect estimates across instrument sets**

Given the relationships shown above, particularly that reducing the stringency of significance thresholding for rare IVs does not reduce instrument specificity (pleiotropy profiles remain consistant), I have compared IVW effect estimates for protein --\> outcome relationships instrumented using at least three common (at P\<5e-8) and at least three rare (at P\<1e-4) cis or trans instruments.

```{r}
#| label: compare estimates
#| cache: TRUE
#| message: FALSE

# Functions
# Generate a table of IVW estimates for a pair of instrument sets
compare_estimates <- function(results_MR, IVset1, IVset2, n_snps=3, cis_trans="all"){
  # Pull out IVW estimates and SE for instrument sets
  if(cis_trans == "all"){
    df <- lapply(results_MR, function(x){
      x |> dplyr::filter(method == "Inverse variance weighted", 
                         cis_trans == "cis_trans",
                         IV_set %in% c(IVset1,IVset2))
    })
  }
  if(cis_trans == "cis"){
    df <- lapply(results_MR, function(x){
      x |> dplyr::filter(method == "Inverse variance weighted", 
                         cis_trans == "cis",
                         IV_set %in% c(IVset1,IVset2))
    })
  }
  if(cis_trans == "trans"){
    df <- lapply(results_MR, function(x){
      x |> dplyr::filter(method == "Inverse variance weighted", 
                         cis_trans == "trans",
                         IV_set %in% c(IVset1,IVset2))
    })
  }
  df_out_b <- do.call("rbind",df) |> 
    dplyr::filter(nsnp >= n_snps) |>
    tidyr::pivot_wider(names_from = "IV_set", values_from = c("b"), id_cols = "exposure") |>
    na.exclude() |>
    dplyr::relocate(exposure,IVset1,IVset2)
  
  df_out_se <- do.call("rbind",df) |> 
    dplyr::filter(nsnp >= n_snps) |>
    tidyr::pivot_wider(names_from = "IV_set", values_from = "se", id_cols = "exposure") |>
    na.exclude() |>
    dplyr::relocate(exposure,IVset1,IVset2)
  
  df_out_p <- do.call("rbind",df) |> 
    dplyr::filter(nsnp >= n_snps) |>
    tidyr::pivot_wider(names_from = "IV_set", values_from = "pval", id_cols = "exposure") |>
    na.exclude() |>
    dplyr::relocate(exposure,IVset1,IVset2)
  colnames(df_out_p)[2:3] <- paste0(colnames(df_out_p)[2:3],".pval")
  
  ## Caluclate effect heterogeneity
  if (!(identical(df_out_b$exposure, df_out_se$exposure) &&
        identical(colnames(df_out_b), colnames(df_out_se)))) {
    message("Error - mismatch of beta and SE dataframes")
    return(NULL)
  }
  
  beta_vecs <- lapply(split(df_out_b[,c(2,3)], 1:nrow(df_out_b)), as.numeric)
  se_vecs <- lapply(split(df_out_se[,c(2,3)], 1:nrow(df_out_se)), as.numeric)
  
  het_res <- do.call("rbind", mapply(effect_heterogeneity, beta_vec = beta_vecs, se_vec = se_vecs, SIMPLIFY = F)) |>
    as.data.frame() |>
    dplyr::mutate(exposure = df_out_b$exposure, .before = "Q") |>
    dplyr::mutate(Qpval_bon = p.adjust(Qpval, method = "bonferroni"),
                  Qpval_fdr = p.adjust(Qpval, method = "fdr"))
  
  df_out <- dplyr::left_join(df_out_b, df_out_se, by = "exposure", suffix = c(".beta",".se")) |>
    dplyr::left_join(df_out_p, by = "exposure") |>
    dplyr::left_join(het_res, by = "exposure")
  
  return(df_out)
}

# Generate a table of cis vs trans IVW estimates for given instrument set
compare_cistrans <- function(results_MR, IVset, n_snps=3){
  # Pull out IVW estimates and SE for instrument sets
  df <- lapply(results_MR, function(x){
    x |> 
      dplyr::filter(method == "Inverse variance weighted",
                    IV_set %in% IVset,
                    nsnp >= n_snps,
                    cis_trans %in% c("cis","trans")) |>
      dplyr::mutate(IV_set = paste(IV_set, cis_trans, sep = "_")) |>
      dplyr::arrange(cis_trans)})
  df <- df[lapply(df, nrow) == 2]
  
  df_out_b <- do.call("rbind",df) |> 
    tidyr::pivot_wider(names_from = "IV_set", values_from = "b", id_cols = "exposure") |>
    na.exclude()
  
  df_out_se <- do.call("rbind",df) |> 
    tidyr::pivot_wider(names_from = "IV_set", values_from = "se", id_cols = "exposure") |>
    na.exclude()
  
  df_out_p <- do.call("rbind",df) |> 
    tidyr::pivot_wider(names_from = "IV_set", values_from = "pval", id_cols = "exposure") |>
    na.exclude()
  colnames(df_out_p)[2:3] <- paste0(colnames(df_out_p)[2:3],".pval")
  
  ## Caluclate effect heterogeneity
  if (!(identical(df_out_b$exposure, df_out_se$exposure) &&
        identical(colnames(df_out_b), colnames(df_out_se)))) {
    message("Error - mismatch of beta and SE dataframes")
    return(NULL)
  }
  
  beta_vecs <- lapply(split(df_out_b[,c(2,3)], 1:nrow(df_out_b)), as.numeric)
  se_vecs <- lapply(split(df_out_se[,c(2,3)], 1:nrow(df_out_se)), as.numeric)
  
  het_res <- do.call("rbind", mapply(effect_heterogeneity, beta_vec = beta_vecs, se_vec = se_vecs, SIMPLIFY = F)) |>
    as.data.frame() |>
    dplyr::mutate(exposure = df_out_b$exposure, .before = "Q") |>
    dplyr::mutate(Qpval_bon = p.adjust(Qpval, method = "bonferroni"),
                  Qpval_fdr = p.adjust(Qpval, method = "fdr"))
  
  df_out <- dplyr::left_join(df_out_b, df_out_se, by = "exposure", suffix = c(".beta",".se")) |>
    dplyr::left_join(df_out_p, by = "exposure") |>
    dplyr::left_join(het_res, by = "exposure")
  
  return(df_out)
}

# Proportion of concordant/discordant MR estimates (of results significant in at least one IV set) based on heterogenity estimates and direction of effect
prop_concordant <- function(estimates, pval_thresh = 0.05){
  
  estimate_cols <- grep("\\.beta",colnames(estimates), value = T)
  pval_cols <- grep("\\.pval",colnames(estimates), value = T)
  sets <- sub(".beta", "", estimate_cols) # IV sets being compared
  
  # Number of tests sig. for at least one IV set
  rows_sig <- apply(estimates[,pval_cols] <= pval_thresh, 1, any)
  
  # Heterogenity, of tests sig. for at least one IV set
  df_counts <- data.frame(n_instrumented_genes = nrow(estimates),
                       n_nosig = nrow(estimates) - sum(rows_sig),
                       n_sig_nohet = sum(rows_sig & estimates$Qpval >= pval_thresh),
                       n_sig_het_concordant = sum(rows_sig & estimates$Qpval <= pval_thresh & sign(estimates[estimate_cols[1]]) == sign(estimates[estimate_cols[2]])),
                       n_sig_het_discordant = sum(rows_sig & estimates$Qpval <= pval_thresh & sign(estimates[estimate_cols[1]]) != sign(estimates[estimate_cols[2]])))
  
  df_props <- df_counts[,2:5]/df_counts[,1]
  colnames(df_props) <- sub("n_","prop_",colnames(df_props))
  
  df_out <- cbind(df_counts, df_props)
  
  # Discordant results
  estimates_het <- estimates |> dplyr::filter(rows_sig & Qpval <= pval_thresh)
  
  df_counts_het <- data.frame(n_het_genes = nrow(estimates_het),
                              sig1 = sum(estimates_het[pval_cols[1]] < estimates_het[pval_cols[2]] & estimates_het[pval_cols[2]] > pval_thresh),
                              sig2 = sum(estimates_het[pval_cols[2]] < estimates_het[pval_cols[1]] & estimates_het[pval_cols[1]] > pval_thresh),
                              n_both_sig_concordant = sum(estimates_het[pval_cols[1]] < pval_thresh & estimates_het[pval_cols[2]] < pval_thresh & 
                                                    sign(estimates_het[estimate_cols[1]]) == sign(estimates_het[estimate_cols[2]])),
                              n_both_sig_discordant = sum(estimates_het[pval_cols[1]] < pval_thresh & estimates_het[pval_cols[2]] < pval_thresh & 
                                                    sign(estimates_het[estimate_cols[1]]) != sign(estimates_het[estimate_cols[2]])))
  
  colnames(df_counts_het)[2] <- paste0("n_",sets[1],"_sig")
  colnames(df_counts_het)[3] <- paste0("n_",sets[2],"_sig")
  
  df_het_props <- df_counts_het[,2:5]/df_counts_het[,1]
  colnames(df_het_props) <- sub("n_","prop_",colnames(df_het_props))
  
  df_out_het <- cbind(df_counts_het, df_het_props)
  
  return(list(all = df_out, sigQ = df_out_het))
}

plot_concordance <- function(concordance, title){
  
  plot_dat_1 <- concordance$all |> 
    tibble::rownames_to_column("outcome") |> 
    dplyr::mutate(n_het = n_sig_het_concordant + n_sig_het_discordant) |>
    dplyr::select(c(1,3,4,11)) |> tidyr::pivot_longer(cols = c(2:4)) |>
    dplyr::mutate(name = factor(name, levels = rev(c("n_nosig","n_sig_nohet","n_het"))))
  
  p1 <- plot_dat_1 |> ggplot(aes(x = outcome, y = value, fill = name)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = "Outcome", 
         y = " Proportion of MR tests \n(protein --> outcome)") +
    scale_fill_manual(name = "",
      values = c("tomato","grey30","grey90"),
      labels = c("Qpval<thresh","Qpval>thresh","NS")) +
    geom_text(dat = plot_dat_1 |> dplyr::filter(name == "n_het"), aes(x = outcome, y = 1.05, label = value), colour = "tomato") +
    geom_text(dat = plot_dat_1 |> dplyr::filter(name == "n_sig_nohet"), aes(x = outcome, y = 0.92, label = value), colour = "white") +
    geom_text(dat = plot_dat_1 |> dplyr::filter(name == "n_nosig"), aes(x = outcome, y = 0.5, label = value), colour = "grey30") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 10),
          legend.key.size = unit(0.6, "cm"),
          legend.position = "bottom") +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
  
  plot_dat_2 <- concordance$Qsig |> 
    tibble::rownames_to_column("outcome") |> 
    dplyr::select(c(1,3:6)) |> tidyr::pivot_longer(cols = c(2:5))
  
  sets = unique(plot_dat_2$name)[1:2]
  sets = sub("_sig$","",sub("^n_","",sets))
  
  if(all(grepl("cis|trans", sets))){
    fill_names = c(paste(IVsets[sub("_(cis|trans)$","",sets)],c("(cis)","(trans)")), "Sig. conc.", "Sig. dis.")
    names(fill_names) <- unique(plot_dat_2$name)
    fill_col = c("grey", "grey30","tomato","steelblue1")
  }else{
    fill_names = c(IVsets[sets], "Sig. conc.", "Sig. dis.")
    names(fill_names) <- unique(plot_dat_2$name)
    fill_col = c(as.character(cols_IVsets[fill_names[1:2]]), "grey", "grey30")
    names(fill_col) <- names(fill_names)
  }
  
  p2 <- plot_dat_2 |> ggplot(aes(x = outcome, y = value, fill = name)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = "Outcome", 
         y = " Proportion of heterogenous MR results \n(protein --> outcome)") +
    scale_fill_manual(name = "",
      values = fill_col,
      labels = fill_names) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 10),
          legend.position = "bottom",
          legend.key.size = unit(0.6, "cm")) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
  
  gridExtra::grid.arrange(p1,p2,nrow = 1, top = title)
  
}

plot_estimates <- function(estimates, name, pval_thresh = 0.05){
  
  estimates <- as.data.frame(estimates)
  col_names <- names(estimates)
  
  if(any(grepl("cis|trans",col_names))){
    lab_x <- "cis-IVs"
    lab_y <- "trans-IVs"
  } else {
    lab_x <- IVsets[sub("\\.beta","",col_names[2])]
    lab_y <- IVsets[sub("\\.beta","",col_names[3])]
  }
  
  CI_x <- data.frame(LCI = estimates[,2] - 1.96*estimates[,4], UCI = estimates[,2] + 1.96*estimates[,4])
  CI_y <- data.frame(LCI = estimates[,3] - 1.96*estimates[,5], UCI = estimates[,3] + 1.96*estimates[,5])
  
  mod_lm <- lm(estimates[,3] ~ estimates[,2])
  slope <- round(coefficients(mod_lm)[2],2)
  
  flag_point <- estimates$Qpval <= pval_thresh
  
  range_min <- min(min(estimates[,2]), min(estimates[,3])) #min(na.exclude(errorbar_xmin, errorbar_ymin)))
  range_max <- max(max(estimates[,2]), max(estimates[,3])) #min(na.exclude(errorbar_xmax, errorbar_ymax)))
  
  estimates_plot <- estimates |>
    dplyr::mutate(exposure = factor(exposure, levels = estimates$exposure[order(estimates$Qpval, decreasing = TRUE)]),
                  errorbar_xmin = ifelse(flag_point, CI_x$LCI, NA),
                  errorbar_xmax = ifelse(flag_point, CI_x$UCI, NA),
                  errorbar_ymin = ifelse(flag_point, CI_y$LCI, NA),
                  errorbar_ymax = ifelse(flag_point, CI_y$UCI, NA)) |>
    dplyr::arrange(exposure)
  
  estimates_plot |>
    ggplot(aes_string(x = col_names[2], y = col_names[3])) +
    geom_abline(slope = 1, intercept = 0, colour = "steelblue", lty = 2, lwd = 0.5) +
    geom_point(aes(colour = estimates_plot$Qpval <= pval_thresh), size = 2, shape = 16) +  
    geom_errorbar(xmin = estimates_plot$errorbar_xmin,
                  xmax = estimates_plot$errorbar_xmax, colour = "tomato") +
    geom_errorbar(ymin = estimates_plot$errorbar_ymin,
                  ymax = estimates_plot$errorbar_ymax, colour = "tomato") +
    geom_smooth(method = "lm", se = FALSE, colour = "tomato", lwd = 0.5) +
    annotate(geom = "text", x = range_min/2, y = range_max, label = paste0("Slope=",slope), size = 3, colour = "tomato") +
    scale_colour_manual(values = c("black","tomato")) +
    scale_x_continuous(limits = c(range_min, range_max)) +
    scale_y_continuous(limits = c(range_min, range_max)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 10), legend.position = "none") +
    labs(x=lab_x, y=lab_y, title = name)
}

# Upset plot comparing signifiant causal estimates across variant sets
# Restrtcited to those proteins that could be instrumented with >= 3 variants across the comparison sets

plot_upset <- function(IVsets_list, cis_trans = "all", pval_thresh = 0.05){
  
  if(cis_trans == "all"){inlist = summary_sig}
  if(cis_trans == "cis"){inlist = summary_sig_cis}
  if(cis_trans == "trans"){inlist = summary_sig_trans}
  
  set_sig <- lapply(inlist, function(x){
    dat <- x$min_p[,IVsets_list]
    # Extract only those tests that could be conducted in all IV sets
    tests <- rownames(dat) # Exposure_outcome
    shared_tests <- tests[apply(dat, MARGIN = 1, function(x){sum(!(is.na(x)))}) == 2]
    dat <- dat[shared_tests,]
    dat_sig <- dat <= pval_thresh
    
    tests <- rownames(dat_sig)
    list_sig <- apply(dat_sig, MARGIN = 2, FUN = function(x){tests[x]}, simplify = F) # Names of significant exposures
    names(list_sig) <- IVsets[names(list_sig)]
    list_sig <- list_sig[lapply(list_sig, length) > 0] # Remove empty exposure sets
    return(list_sig)
  })
  
  # Upset plots
  plots <- mapply(FUN = function(list_sig, outcome){
    if(length(list_sig) == 1){return(NULL)}
    upset(fromList(list_sig), order.by = "freq", text.scale = 1.5, mainbar.y.label = paste0("Protein:",outcome,"\n (IVW p<", pval, ")"))
  }, list_sig = set_sig, outcome = names(set_sig), SIMPLIFY = F)
  
  plots <- plots[!sapply(plots, is.null)]
  
  # for (v in names(plots)) {
  #   print(plots[[v]])
  #   #grid.text(v, x = 0.65, y=0.97, gp = gpar(fontsize = 20))
  #   grid.edit('arrange', name = v)
  #   vp <- grid.grab()
  #   plots[[v]] <- vp
  # }
  return(plots)
}
```

1a) Common vs rare variant cis-IVs (ExWAS)\
1b) Common vs rare variant trans-IVs (ExWAS)

2a) Common vs ultra-rare variant cis-IVs (ExWAS)\
2b) Common vs ultra-rare variant trans-IVs (ExWAS)

```{r}
#| label: fig-compIVs-commonvrare
#| cache: TRUE
#| message: FALSE
#| fig.width: 14
#| fig.height: 6

comp_IVWs_1a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_1a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_1a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_1a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_1a, name = names(comp_IVWs_1a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and rare cis-pQTL instrumentation strategies")

comp_IVWs_1b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_1b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_1b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_1b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_1b, name = names(comp_IVWs_1b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and rare trans-pQTL instrumentation strategies")

comp_IVWs_2a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_2a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_2a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_2a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_2a, name = names(comp_IVWs_2a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and ultra-rare cis-pQTL instrumentation strategies")

comp_IVWs_2b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_2b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_2b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_2b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_2b, name = names(comp_IVWs_2b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and ultra-rare trans-pQTL instrumentation strategies")
```

```{r}
#| label: fig-compIVs-commonvrare-concordance
#| fig.width: 7
#| fig.height: 5

plot_concordance(concordance = comp_IVWs_1a_conc, title = "IVW estimate concordance: WES common vs rare cis-IVs")
plot_concordance(concordance = comp_IVWs_1b_conc, title = "IVW estimate concordance: WES common vs rare trans-IVs") 

plot_concordance(concordance = comp_IVWs_2a_conc, title = "IVW estimate concordance: WES common vs ultrarare cis-IVs")
plot_concordance(concordance = comp_IVWs_2b_conc, title = "IVW estimate concordance: WES common vs ultrarare trans-IVs") 

# Mean concordance/discordance (across outcomes)
#colMeans(comp_IVWs_1a_conc$all) # cis-common v rare
#colMeans(comp_IVWs_2a_conc$all) # cis-common v ultrarare
```

3a) Common (GWAS) vs rare variant cis-IVs (ExWAS)\
3b) Common (GWAS) vs rare variant trans-IVs (ExWAS)

4a) Common (GWAS) vs ultra-rare variant cis-IVs (ExWAS)\
4b) Common (GWAS) vs ultra-rare variant trans-IVs (ExWAS)

```{r}
#| label: fig-compIVs-commonGWASvrare
#| cache: TRUE
#| message: FALSE
#| fig.width: 14
#| fig.height: 6

comp_IVWs_3a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Multiple test correct for max number of exposures tested across outcomes
alpha_Q_3a <- 0.05/max(unlist(lapply(comp_IVWs_3a, nrow))) #7.6x10^-5

# Proportion concordant
comp_IVWs_3a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_3a, FUN = prop_concordant, pval = 0.05), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_3a, FUN = prop_concordant, pval = 0.05), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_3a, name = names(comp_IVWs_3a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and WES rare cis-pQTL instrumentation strategies")

comp_IVWs_3b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_3b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_3b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_3b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_3b, name = names(comp_IVWs_3b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS  common and WES rare trans-pQTL instrumentation strategies")

comp_IVWs_4a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_4a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_4a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_4a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_4a, name = names(comp_IVWs_4a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and WES ultra-rare cis-pQTL instrumentation strategies")

comp_IVWs_4b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_4b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_4b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_4b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_4b, name = names(comp_IVWs_4b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and WES ultra-rare trans-pQTL instrumentation strategies")

# Extract list of genes with significant heterogenity between cis common and rare/ultrarare variant IVW estimates (at least one of which is significant)
het_genes <- lapply(
  Map(c, 
      #lapply(comp_IVWs_1a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      #lapply(comp_IVWs_2a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_3a, function(x){x |> dplyr::filter(Qpval <= 0.05 & apply(x[,grepl("\\.pval",colnames(x))] < 0.05,1,any)) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_4a, function(x){x |> dplyr::filter(Qpval <= 0.05 & apply(x[,grepl("\\.pval",colnames(x))] < 0.05,1,any)) |> dplyr::pull(exposure)})), FUN = unique)

het_genes <- mapply(FUN = function(x,y){paste(x,y,sep = "_")}, x = het_genes, y = names(het_genes), SIMPLIFY = F)

# Highly significant MR results (from at least 1 IV set)
het_genes_strong <- lapply(
  Map(c, 
      #lapply(comp_IVWs_1a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      #lapply(comp_IVWs_2a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_3a, function(x){x |> dplyr::filter(Qpval <= 0.05 & apply(x[,grepl("\\.pval",colnames(x))] < alpha_cis3,1,any)) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_4a, function(x){x |> dplyr::filter(Qpval <= 0.05 & apply(x[,grepl("\\.pval",colnames(x))] < alpha_cis3,1,any)) |> dplyr::pull(exposure)})), FUN = unique)
het_genes_strong <- het_genes_strong[lapply(het_genes_strong, length) > 0]

het_genes_strong <- mapply(FUN = function(x,y){paste(x,y,sep = "_")}, x = het_genes_strong, y = names(het_genes_strong), SIMPLIFY = F)
```

```{r}
#| label: fig-compIVs-GWAScommonvrare-concordance
#| fig.width: 8
#| fig.height: 5

plot_concordance(concordance = comp_IVWs_3a_conc, title = "IVW estimate concordance: GWAS common vs rare cis-IVs")
plot_concordance(concordance = comp_IVWs_3b_conc, title = "IVW estimate concordance: GWAS common vs rare trans-IVs") 

plot_concordance(concordance = comp_IVWs_4a_conc, title = "IVW estimate concordance: GWAS common vs ultrarare cis-IVs")
plot_concordance(concordance = comp_IVWs_4b_conc, title = "IVW estimate concordance: GWAS common vs ultrarare trans-IVs") 

# Proportion of significant results (sig in at least 1 IV set) that are heterogeneous
apply(comp_IVWs_3a_conc$all, 1, FUN = function(x){(x["n_sig_het_concordant"] + x["n_sig_het_discordant"])/(x["n_instrumented_genes"] - x["n_nosig"])}) * 100 # cis-common v rare
apply(comp_IVWs_4a_conc$all, 1, FUN = function(x){(x["n_sig_het_concordant"] + x["n_sig_het_discordant"])/(x["n_instrumented_genes"] - x["n_nosig"])}) * 100 # cis-common v ultrarare

```

7)  Common cis-IVs vs trans-IVs (GWAS)\
8)  Common cis-IVs vs trans-IVs (ExWAS)\
9)  Rare cis-IVs vs trans-IVs (ExWAS)\
10) Ultra-rare cis-IVs vs trans-IVs\

```{r}
#| label: fig-compIVs-cisvtrans
#| cache: TRUE
#| message: FALSE
#| fig.width: 14
#| fig.height: 6

comp_IVWs_5 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                      IVset = "sun_gwas_variant_common_clumped",
                      n_snps = 3)

# Proportion concordant
comp_IVWs_5_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_5, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_5, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_5, name = names(comp_IVWs_5), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and cis- and trans-pQTL instrumentation strategies")

comp_IVWs_6 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                      IVset = "dhindsa_exwas_variant_common",
                      n_snps = 3)

# Proportion concordant
comp_IVWs_6_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_6, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_6, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_6, name = names(comp_IVWs_6), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from ExWAS common and cis- and trans-pQTL instrumentation strategies")

comp_IVWs_7 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                      IVset = "dhindsa_exwas_variant_rare_pruned",
                      n_snps = 3)

# Proportion concordant
comp_IVWs_7_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_7, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_7, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_7, name = names(comp_IVWs_7), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from ExWAS rare and cis- and trans-pQTL instrumentation strategies")

comp_IVWs_8 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                       IVset = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3)

# Proportion concordant
comp_IVWs_8_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_8, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_8, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_8, name = names(comp_IVWs_8), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from ExWAS ultra-rare and cis- and trans-pQTL instrumentation strategies")

# Extract list of genes with significant heterogenity between cis and trans estimates
het_genes_cistrans <- lapply(
  Map(c, lapply(comp_IVWs_5, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_6, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_7, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_8, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)})), function(x){unique(x)})

het_genes_cistrans <- mapply(FUN = function(x,y){paste(x,y,sep = "_")}, x = het_genes_cistrans , y = names(het_genes_cistrans), SIMPLIFY = F)

```

```{r}
#| label: fig-compIVs-cisvtans-concordance
#| fig.width: 9
#| fig.height: 5

plot_concordance(concordance = comp_IVWs_5_conc, title = "IVW estimate concordance: GWAS common cis vs trans-IVs")
plot_concordance(concordance = comp_IVWs_6_conc, title = "IVW estimate concordance: WES common cis vs trans-IVs") 
plot_concordance(concordance = comp_IVWs_7_conc, title = "IVW estimate concordance: WES rare cis vs trans-IVs")
plot_concordance(concordance = comp_IVWs_8_conc, title = "IVW estimate concordance: WES common ultrarare cis vs trans-IVs") 
```

## Causes of IVW heterogeneity

Differeneces in IVW estimates between IV sets could be caused by differences in pleiotropy profile, differences in IV number, the number of IVs that are true-cis, heterogenity in SNP effects within each IV set??...

Plots of highly significant heterogeneous results:

```{r}
#| label: top-heterogeous
#| fig.width: 13
#| fig.height: 44
#| message: FALSE

# Qpval < 0.05, at least 1 genome:common or exome:rare/ultrarare IVW P < 5x10^-5 (FDR corrected)
gridExtra::grid.arrange(grobs = lapply(strsplit(unlist(het_genes_strong), "_"), function(x){plot_results(exposure = x[1], outcome = x[2])}), ncol = 2)

```

### Heterogenity vs pleiotropy profile

If the cis-pleiotropy profile of common and rare IVs for a gene is more consistant (greater overlap in the genes instrumented), are the MR effect estimates more likely to be consistent?

```{r}
#|label: heterogenity-vs-pleiotropy

# Pull list of all cis pqtls used as instruments in the relaxed threshold analysis
# exome::rare and genome::common for this analysis

varIDs_relaxed <- list("dhindsa_exwas_variant_rare_pruned" = all_cisIVs_relaxed |> dplyr::filter(IV_set == "exome:rare") |> pull(SNP_ID),
                       "sun_gwas_variant_common_clumped" = all_cisIVs_relaxed |> dplyr::filter(IV_set == "genome:common") |> pull(SNP_ID))
# Correct SNP ID formatting
varIDs_relaxed <- lapply(varIDs_relaxed, function(x){toupper(gsub("_","-",sub(":","-",x)))})

# Summarise associations per variant (pQTL) - pleiotropy assessment
pQTLs_assocs_MRset <- mapply(function(x, y){
  x <- x |> dplyr::group_by(variant, proximal_gene) |> 
    dplyr::summarise(n_assocs = n(), 
                     n_cis = length(which(cis_trans == "cis")),
                     n_truecis = sum(proximal_gene == paste0("'",protein_target,"'")),
                     n_trans = length(which(cis_trans == "trans")),
                     assoc_genes = paste(unique(protein_target), collapse = ";"),
                     assoc_cis_genes = paste(unique(protein_target[cis_trans == "cis"]), collapse = ";"),
                     assoc_trans_genes = paste(unique(protein_target[cis_trans == "trans"]), collapse = ";"))
  x$IV_set <- y
  return(x)
}, x = lookup_pQTLs_MRset, y = as.list(names(lookup_pQTLs_MRset)), SIMPLIFY = F)

pqtls_assocs_joined_MRset <- do.call("rbind", pQTLs_assocs_MRset ) |>
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |>
  dplyr::filter(IV_set %in% c("exome:rare","genome:common")) |>
  dplyr::mutate(all_cis_genes = assoc_cis_genes,
                variant = gsub("_","-",sub(":","-",variant))) |>
  tidyr::separate_rows(assoc_cis_genes, sep = ";") |>
  dplyr::filter(variant %in% c(varIDs_relaxed$dhindsa_exwas_variant_rare_pruned, varIDs_relaxed$sun_gwas_variant_common_clumped)) # Restrict to cis-IVs used in MR

# Calculate the overlap in pleiotropy profile for all rare and common cis-IVs for a given protein
# That is, of all proteins in the cis window cis-IVs for a given protein associate with, how many are shared between common and rare IV sets?
cispleiotropy_overlap <- pqtls_assocs_joined_MRset[,c("assoc_cis_genes", "IV_set", "n_cis", "all_cis_genes")] |>
  dplyr::group_by(assoc_cis_genes, IV_set) |>
  dplyr::summarise(total_cisgenes = paste(unique(unlist(strsplit(all_cis_genes, ";"))), collapse = ";"),
                   count_cisgenes = length(unique(unlist(strsplit(all_cis_genes, ";"))))) |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(count_cisgenes, total_cisgenes),
    names_glue = "{IV_set}_{.value}"
  ) |>
  rowwise() |>
  dplyr::mutate( # Calc prpoprtion of shared cis genes
    prop_shared = {
      s1 <- strsplit(`exome:rare_total_cisgenes`, ";")[[1]]
      s2 <- strsplit(`genome:common_total_cisgenes`, ";")[[1]]
      length(intersect(s1, s2)) / length(union(s1, s2))
    },
   n_total_cisgenes = {
      s1 <- strsplit(`exome:rare_total_cisgenes`, ";")[[1]]
      s2 <- strsplit(`genome:common_total_cisgenes`, ";")[[1]]
      length(unique(c(s1, s2)))}
   ) |> ungroup() |> na.exclude()

# Overlap between all assocaited genes
globalpleiotropy_overlap <- pqtls_assocs_joined_MRset[,c("assoc_cis_genes", "IV_set", "n_cis", "assoc_genes")] |>
  dplyr::group_by(assoc_cis_genes, IV_set) |>
  dplyr::summarise(total_genes = paste(unique(unlist(strsplit(assoc_genes, ";"))), collapse = ";"),
                   count_genes = length(unique(unlist(strsplit(assoc_genes, ";"))))) |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(count_genes, total_genes),
    names_glue = "{IV_set}_{.value}"
  ) |>
  rowwise() |>
  dplyr::mutate( # Calc prpoprtion of shared cis genes
    prop_shared = {
      s1 <- strsplit(`exome:rare_total_genes`, ";")[[1]]
      s2 <- strsplit(`genome:common_total_genes`, ";")[[1]]
      length(intersect(s1, s2)) / length(union(s1, s2))
    },
   n_total_genes = {
      s1 <- strsplit(`exome:rare_total_genes`, ";")[[1]]
      s2 <- strsplit(`genome:common_total_genes`, ";")[[1]]
      length(unique(c(s1, s2)))}
   ) |> ungroup() |> na.exclude()

# Append pleiotropy overlap to MR estimate heterogeniety results
# Common genome vs rare exome cis-MR
comp_IVWs_3a <- lapply(comp_IVWs_3a, function(x){
  dplyr::left_join(x, globalpleiotropy_overlap, by = c("exposure" = "assoc_cis_genes"))
})

# Append information on num IVs in set and evidence of heterogeneity
## Add column to link by
comp_IVWs_3a <- mapply(function(x,y){
  x$pair <- paste(x$exposure, y, sep = "_") 
  return(x)}, x = comp_IVWs_3a, y = as.list(names(comp_IVWs_3a)), SIMPLIFY = F)

## Extract IV heterogeneity results (imported above)
het_rare <- het |> dplyr::filter(variant_type == "cis" & n_snps >= 3 & method == "Inverse variance weighted" & variant_set %in% "dhindsa_exwas_variant_rare_pruned")
het_rare <- split(het_rare, f = het_rare$outcome)
het_common <- het |> dplyr::filter(variant_type == "cis" & n_snps >= 3 & method == "Inverse variance weighted" & variant_set %in% "sun_gwas_variant_common_clumped")
het_common <- split(het_common, f = het_common$outcome)

## Append
comp_IVWs_3a <- mapply(function(x,y){
  x$dhindsa_exwas_variant_rare_pruned.snphet.pval = y[match(x$pair, y$iv_set)]$Q_pval
  x$dhindsa_exwas_variant_rare_pruned.nsnps = y[match(x$pair, y$iv_set)]$n_snps
  return(x)
}, x = comp_IVWs_3a, y = het_rare, SIMPLIFY = F)

comp_IVWs_3a <- mapply(function(x,y){
  x$sun_gwas_variant_common_clumped.snphet.pval = y[match(x$pair, y$iv_set)]$Q_pval
  x$sun_gwas_variant_common_clumped.nsnps = y[match(x$pair, y$iv_set)]$n_snps
  return(x)
}, x = comp_IVWs_3a, y = het_common, SIMPLIFY = F)
```

Look at which factors associate with the likelihood of two IVW estimates exhibiting heterogenity:

```{r}
#|label: associations-w-heterogenity

comp_IVWs_3a_plotdat <- do.call("rbind", comp_IVWs_3a) |>
  dplyr::filter(sun_gwas_variant_common_clumped.pval < 0.05 | dhindsa_exwas_variant_rare_pruned.pval < 0.05) |> # IVW is significant for at least one set
  dplyr::mutate(sig_het = ifelse(Qpval < 0.05, 1, 0),
                low_nsnps = ifelse(dhindsa_exwas_variant_rare_pruned.nsnps == 3 | sun_gwas_variant_common_clumped.nsnps == 3, 1, 0),
                het_snpeffects = ifelse(dhindsa_exwas_variant_rare_pruned.snphet.pval < 0.05 | sun_gwas_variant_common_clumped.snphet.pval < 0.05, 1, 0))

# Explore relationship
lm(Q ~ prop_shared, data = comp_IVWs_3a_plotdat) |> summary()
lm(Q ~ n_total_genes > 1, data = comp_IVWs_3a_plotdat) |> summary()
lm(Q ~ low_nsnps, data = comp_IVWs_3a_plotdat) |> summary()
lm(Q ~ het_snpeffects, data = comp_IVWs_3a_plotdat) |> summary()

# Heterogenity Q statistics vs pleiotropy sharing
# set.seed(4)
# comp_IVWs_3a_plotdat |> ggplot(aes(x = prop_shared < 1, y = Q)) +
#   geom_jitter(aes(colour = Qpval < alpha_Q_3a)) +
#   geom_boxplot(outliers = F, width = 0.3) +
#   scale_colour_manual(values = c("grey50", "red")) +
#   theme_minimal()

```

Repeat the variant level pleiotropy comparison above (which uses the top most pleiotropic common and rare variant per gene), on the gene level (aggregating pleiotropic associations across all IVs for a gene):

```{r}
#| label: rare-v-common-pleiotropy-genelevel
#| fig.height: 4
#| fig.width: 13.5

# Using the all cis-IVs for shared genes

## Gene pleiotropy
p7gene.cis <- cispleiotropy_overlap  |>
  dplyr::select(1:3) |>
  tidyr::pivot_longer(cols = 2:3) |> 
  dplyr::mutate(IV_set = gsub("_.*","",name)) |>
  ggplot(aes(x = IV_set, y = value)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of unique cis-pQTL associations\n across cis-IV for protein")

#w1.gene <- wilcox.test(cispleiotropy_overlap$`exome:rare_count_cisgenes`, cispleiotropy_overlap$`genome:common_count_cisgenes`)

p7gene.global <- globalpleiotropy_overlap  |>
  dplyr::select(1:3) |>
  tidyr::pivot_longer(cols = 2:3) |> 
  dplyr::mutate(IV_set = gsub("_.*","",name)) |>
  ggplot(aes(x = IV_set, y = value)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  ggpubr::stat_compare_means(label.x.npc = 0.5, label = "p.format", size = 2) + #Wilcoxon test
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of unique pQTL associations\n across cis-IV for protein")

#w2.gene <- wilcox.test(globalpleiotropy_overlap$`exome:rare_count_genes`, globalpleiotropy_overlap$`genome:common_count_genes`)

## Gene wise comparison
p7gene <- cispleiotropy_overlap |>
  ggplot(aes(x = `exome:rare_count_cisgenes`, y = `genome:common_count_cisgenes`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_jitter(size = 1) +
  scale_x_continuous(limits = c(0,20), breaks = 1:20) +
  scale_y_continuous(limits = c(0,20), breaks = 1:20) +
  scale_colour_gradient(low = "grey90", high = "red") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. unique cis-pQTL associations \n (across all rare cis-IV for protein)",
       y = "No. unique cis-pQTL associations \n (across all common cis-IV for protein)")
  
p8gene <- globalpleiotropy_overlap |> dplyr::arrange(prop_shared) |>
  ggplot(aes(x = `exome:rare_count_genes` , y = `genome:common_count_genes`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_point(size = 1) +
  scale_x_continuous(limits = c(0,650)) +
  scale_y_continuous(limits = c(0,650)) +
  scale_colour_gradient(low = "grey", high = "red") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. pQTL associations \n (across all rare cis-IV for protein)",
       y = "No. pQTL associations \n (across all common cis-IV for protein)")
  
p8gene_inset <- globalpleiotropy_overlap |> dplyr::arrange(prop_shared) |>
  ggplot(aes(x = `exome:rare_count_genes` , y = `genome:common_count_genes`, colour = prop_shared)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey50") +
  geom_jitter() +
  scale_x_continuous(limits = c(0,30)) +
  scale_y_continuous(limits = c(0,30)) +
  scale_colour_gradient(low = "grey", high = "red") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        plot.background = element_rect(fill = "white", colour = "grey")) +
  labs(x = "No. pQTL associations \n (across all rare cis-IV for protein)",
       y = "No. pQTL associations \n (across all common cis-IV for protein)")

p8gene_join <- cowplot::ggdraw() + cowplot::draw_plot(p8gene) + cowplot::draw_plot(p8gene_inset, x = 0.5, y = 0.6, width = 0.95 - 0.5, height = 0.95 - 0.6)

## Average proportion of shared cis genes
summary(cispleiotropy_overlap$prop_shared)
## Average proportion of shared tranas genes
summary(globalpleiotropy_overlap$prop_shared)
  
# Correlation between cis and trans associations 
p8gene.2.plotdat <- dplyr::left_join(cispleiotropy_overlap[,1:3], globalpleiotropy_overlap[,1:3], by = "assoc_cis_genes")

p8gene.2 <- p8gene.2.plotdat |>
  ggplot() +
  geom_jitter(aes(x = `exome:rare_count_cisgenes`, y = `exome:rare_count_genes`-`exome:rare_count_cisgenes`), 
              width = 0.2, height = 0, size = 1, col = cols_IVsets["exome:rare"]) +
  geom_jitter(aes(x = `genome:common_count_cisgenes`, y = `genome:common_count_genes`-`genome:common_count_cisgenes`), 
              width = 0.2, height = 0, size = 1, col = cols_IVsets["genome:common"]) +
  geom_smooth(aes(x = `exome:rare_count_cisgenes`, y = `exome:rare_count_genes`-`exome:rare_count_cisgenes`), 
              method = "lm", se = FALSE,  col = cols_IVsets["exome:rare"]) +
  geom_smooth(aes(x = `genome:common_count_cisgenes`, y = `genome:common_count_genes`-`genome:common_count_cisgenes`), 
              method = "lm", se = FALSE,  col = cols_IVsets["genome:common"]) +
  scale_y_log10() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "No. cis-pQTL associations",
       y = "No. trans-pQTL associations")

gridExtra::grid.arrange(p7gene.cis, p7gene.global, p7gene, p8gene_join, p8gene.2, nrow = 1, widths = c(0.1,0.1,0.26,0.26,0.26), top = "Pleiotropy profile of gene based on aggregted pleiotropy profile of all rare or common cis-IVs for protein")

```




## Allelic series

Assessing for monotonically increasing causal effect estimates (Wald Ratios) across rare-variant series of alleles:\
- Rare variants (\<1% MAF): benign missense variants (BMV) - damaging missense variants (DMVs) - protein truncating variants (PTVs)\

Using the [COAST-SS](https://github.com/insitro/AllelicSeries) approach (developed for allelic series association tests), which has been developed to handle GWAS summary statistics, but here we are passing Wald Ratios (variant outcome effect/ variant protein effect)

What is needed here?:\
- Summary statistics for all rare variants (capped at p\<0.01 by the AzPheWAS portal for pQTLs) falling in the coding region of genes encoding the O'link plasma proteome assayed proteins\
- The corresponding variant associations for outcomes of interest (with which to calculate Wald ratios)  - Variant predicted functional effects (with which to categorise SNPs)

**Magnitude of variant class effects**

Do cis-variants predicted to be more deleterious in nature (loss of function) actually have larger effects on protein expression?

```{r}
#| label: vep-classes
#| fig.width: 7
#| fig.height: 3

# Read in variant effect estimates, wald ratios and VEP annotations
load(file.path(res_dir,"results_molecular_waldratios_ciscodingvariants_VEPannotated.rda"))

wald_ratios <- lapply(wald_ratios, function(x){
  x$annots_relaxed <- as.factor(x$annots_relaxed)
  x})

vep_class <- c("1" = "syn./noncoding", 
               "2" = "benign missense",
               "3" = "del. missense",
               "4" = "putative LoF")

cols_vep <- c("1" = "steelblue", 
              "2" = "olivedrab3",
              "3" = "orange",
              "4" = "tomato")

# Plot variant consequnce call on cis pQTL effect
wald_ratios$LDL |> 
  dplyr::filter(!is.na(annots_relaxed)) |> 
  ggplot(aes(x = beta.exposure, 
             y = annots_relaxed)) +
  ggridges::geom_density_ridges(aes(fill = annots_relaxed), trim = TRUE) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  scale_y_discrete(labels = vep_class) +
  scale_fill_manual(values = cols_vep) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "cis-pQTL effect (beta)", y = "Variant annotation")

# Plot wald ratios split by variant consequence
plot_wald_vep <- function(gene, outcome){
  dat <- wald_ratios[[outcome]] |> 
    dplyr::filter(exposure == gene) |>
    dplyr::arrange(annots_relaxed, abs(wald_ratio_b)) |>
    dplyr::mutate(SNP = factor(SNP, levels = SNP))
  
  res <- coastss_annots_relaxed[[outcome]] |>
    dplyr::filter(exposure == gene)
  
  p_theme <- theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          legend.position = "bottom",
          legend.key.height = unit(0.2, 'cm'))
  
  # Exposure
  p_exp <- ggplot(dat, aes(x = beta.exposure, y = SNP, colour = annots_relaxed)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point() +
    scale_colour_manual(values = cols_vep, labels = vep_class, name = "") +
    geom_errorbarh(aes(xmin = beta.exposure-1.96*se.exposure, xmax = beta.exposure+1.96*se.exposure)) +
    labs(x = "Bx", y = "", 
         title = paste("Exposure:", gene),
         subtitle = "") +
    guides(colour = guide_legend(nrow=2, byrow=TRUE)) +
    p_theme
  
  # Outcome
  p_out <- ggplot(dat, aes(x = beta.outcome, y = SNP, colour = annots_relaxed)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point() +
    scale_colour_manual(values = cols_vep, labels = vep_class, name = "") +
    geom_errorbarh(aes(xmin = beta.outcome-1.96*se.outcome, xmax = beta.outcome+1.96*se.outcome)) +
    labs(x = "By", y = "", 
         title = paste("Outcome:", outcome),
         subtitle = "") +
    guides(colour = guide_legend(nrow=2, byrow=TRUE)) +
    p_theme
  
  # Wald Ratio
  p_wald <- ggplot(dat, aes(x = wald_ratio_b, y = SNP, colour = annots_relaxed)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point() +
    scale_colour_manual(values = cols_vep, labels = vep_class, name = "") +
    geom_errorbarh(aes(xmin = wald_ratio_b-1.96*wald_ratio_se, xmax = wald_ratio_b+1.96*wald_ratio_se)) +
    labs(x = "Wald Ratio (By/Bx)", y = "", 
         title = paste("Exposure:", gene, "Outcome:", outcome),
         subtitle = paste("COAST-SS p-value:", signif(res$alleleic_skat_p, 2))) +
    guides(colour = guide_legend(nrow=2, byrow=TRUE)) +
    p_theme
  
  gridExtra::grid.arrange(grobs = list(p_wald, p_exp, p_out), nrow = 1)
}

# Run COAST-SS analysis on select exposure or outcome betas
run_coastss <- function(wrs, annot_col, weights, protein, outcome="BMI", betas="exposure"){
  
  x <- wrs[[outcome]] |> 
    dplyr::filter(exposure == protein,
                  allele_match == TRUE)
  
    count_vars <- table(x[,annot_col])
    count_vars_string <- paste(paste0(names(count_vars), "=", count_vars),collapse=";")
    
  if(betas == "exposure"){
      mean_betas <- x |> dplyr::group_by_at(annot_col) |> dplyr::summarise(mean(beta.exposure))
      mean_betas_string = paste(paste0(mean_betas[[1]], "=", round(mean_betas[[2]],5)), collapse=";")
      mean_absbetas <- x |> dplyr::group_by_at(annot_col) |> dplyr::summarise(mean(abs(beta.exposure)))
      mean_absbetas_string = paste(paste0(mean_absbetas[[1]], "=", round(mean_absbetas[[2]],5)), collapse=";")
    
    coastss <- AllelicSeries::COASTSS(
      anno = x[,annot_col],
      beta = abs(x[,"beta.exposure"]), # Absolute betas to assess magnitude of effect
      se = x[,"se.exposure"],
      maf = x[,"eaf.exposure"],
      weights = weights
    )
  }
  
    if(betas == "outcome"){
      mean_betas <- x |> dplyr::group_by_at(annot_col) |> dplyr::summarise(mean(beta.outcome))
      mean_betas_string = paste(paste0(mean_betas[[1]], "=", round(mean_betas[[2]],5)), collapse=";")
      mean_absbetas <- x |> dplyr::group_by_at(annot_col) |> dplyr::summarise(mean(abs(beta.outcome)))
      mean_absbetas_string = paste(paste0(mean_absbetas[[1]], "=", round(mean_absbetas[[2]],5)), collapse=";")
      
      coastss <- AllelicSeries::COASTSS(
        anno = x[,annot_col],
        beta = abs(x[,"beta.outcome"]), # Absolute betas to assess magnitude of effect
        se = x[,"se.outcome"],
        maf = x[,"eaf.outcome"],
        weights = weights
      )
    }
    
    res <- data.frame(
      "exposure" = protein,
      "outcome" = outcome,
      "annot_count" = count_vars_string, # number of variants in each annotation class
      "annot_meanbeta" = mean_betas_string, # mean wald ratio of variants in class
      "annot_meanabsbeta" = mean_absbetas_string,
      "alleleic_skat_p" = coastss@Pvals[coastss@Pvals$test == "allelic_skat","pval"])
    
  return(res)
}

```

### COAST-SS analysis

```{r}
#| label: COAST-SS 
#| fig.width: 12
#| fig.height: 20

# Load results from COAST-SS allelic series anlysis 
# 1 = synonymous/noncoding, 2 = benign missense, 3 = deleterious missense, 4 = putative loss of function

load(file.path(res_dir,"results_molecular_COASTSS_relaxed.rda")) # coastss_annots_relaxed

## Filter out exposure genes for which there are less than two variants available for each of at least three classes, and calculate adjusted p-vals
coastss_annots_relaxed <- lapply(coastss_annots_relaxed, function(x){
  x |> 
    dplyr::filter(!is.na(alleleic_skat_p)) |>
    dplyr::filter(
      map_int(stringr::str_split(annot_count, ";"), function(pairs) {
        counts <- as.integer(stringr::str_extract(pairs, "(?<==)\\d+"))
        sum(counts >= 2, na.rm = TRUE)
      }) >= 3) |>
    dplyr::mutate(adj_p = p.adjust(alleleic_skat_p, method = "BH"))
})

outcome_names <- c("LDL direct" = "LDL",
                    "Body mass index (BMI)" = "BMI",
                    "Vitamin D" = "VitD",
                    "Triglycerides" = "Trig",
                    "Glycated haemoglobin (HbA1c)" = "HbA1c",
                    "Mean platelet (thrombocyte) volume" = "PlatVol",
                    "IGF-1" = "IGF1",
                    "Waist-to-hip ratio adjusted for body mass index (WHRadjBMI)" = "WHRadjBMI",
                    "Red blood cell (erythrocyte) count" = "RBCcount",
                    "Mean corpuscular volume" = "CorpVol",
                    "Systolic blood pressure, automated reading" = "SBP")

## Top results (P<0.05/859; 5.8x10^-5)
alpha_coastss = 0.05/82
coastss_annots_relaxed_top <- do.call("rbind", lapply(coastss_annots_relaxed, function(x){
  x |> dplyr::filter(adj_p < alpha_coastss) |>
    dplyr::mutate(outcome_name = outcome_names[outcome])}))

do.call("rbind", coastss_annots_relaxed) |>
  dplyr::mutate(label = ifelse(adj_p < 0.05, exposure, NA),
                outcome = outcome_names[outcome]) |>
  ggplot(aes(x = exposure, y = -log10(alleleic_skat_p))) +
  facet_grid(outcome ~ ., scale = "free_y") +
  geom_point(aes(colour = adj_p < 0.05)) +
  scale_colour_manual(name = "P(FDR) < 0.05", values = c("grey","tomato")) +
  ggrepel::geom_text_repel(aes(label = label), size = 4, max.overlaps = Inf, force = 10) +
  theme_minimal() +
  theme(strip.text = element_text(size = 10),
        panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.title = element_text(size = 14),
        legend.position = "bottom") +
  labs(x = "Protein exposure", y = "-log10(P) Allelic series")
```

```{r}
#| label: COASTSS-results 
#| fig.width: 12
#| fig.height: 4.5

mapply(FUN = function(x,y){plot_wald_vep(gene = x, outcome = y)}, 
       x = as.list(coastss_annots_relaxed_top$exposure), 
       y = as.list(coastss_annots_relaxed_top$outcome_name), SIMPLIFY = F)

# Split exposure/outcome beta analysis
# run_coastss(wrs = wald_ratios, protein = "CELSR2", outcome = "LDL", betas = "exposure", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# run_coastss(wrs = wald_ratios, protein = "CELSR2", outcome = "LDL", betas = "outcome", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# 
# run_coastss(wrs = wald_ratios, protein = "LY75", outcome = "PlatVol", betas = "exposure", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# run_coastss(wrs = wald_ratios, protein = "LY75", outcome = "PlatVol", betas = "outcome", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# run_coastss(wrs = wald_ratios, protein = "LY75", outcome = "RBCcount", betas = "outcome", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# run_coastss(wrs = wald_ratios, protein = "LY75", outcome = "CorpVol", betas = "outcome", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# 
# run_coastss(wrs = wald_ratios, protein = "LRIG1", outcome = "CorpVol", betas = "exposure", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# run_coastss(wrs = wald_ratios, protein = "LRIG1", outcome = "CorpVol", betas = "outcome", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# 
# run_coastss(wrs = wald_ratios, protein = "DBH", outcome = "SBP", betas = "exposure", weights = c(1,2,3,4), annot_col = "annots_relaxed")
# run_coastss(wrs = wald_ratios, protein = "DBH", outcome = "SBP", betas = "outcome", weights = c(1,2,3,4), annot_col = "annots_relaxed")
```

Do missense mutations have larger pQTL effects than synonymous mutations across proteins? On average expectation would be yes, but there is likely between variant variability. Does COAST-SS analysis support monotonicity of absolute pQTL effects scaling with class (with and without pLOF)? Note that we do not have rare missense variants in this analysis that do not have pQTL effects due to thresholding of summary statistics at p<0.01.

```{r}
# Run COAST-SS analyses for protein expression (restrict to proteins with sufficient variants across classes to test)
coastss_exposures <- do.call("rbind", sapply(coastss_annots_relaxed$BMI$exposure, function(x){run_coastss(protein = x, wrs = wald_ratios, weights = c(1,2,3,4), annot_col = "annots_relaxed")}, simplify = F))

coastss_exposures_noptv <- do.call("rbind", sapply(coastss_annots_relaxed$BMI$exposure, function(x){run_coastss(protein = x, wrs = lapply(wald_ratios, function(x){x |> dplyr::filter(as.numeric(annots_relaxed) < 4)}), weights = c(1,2,3), annot_col = "annots_relaxed")}, simplify = F))
```

## Gene based analysis

In gene based analysis, variant aggregate approaches are used to test for the effect of carrying any single or combination of rare genetic variants in a gene on a phenotype. Variant masks are often generated based on variant properties (e.g putative loss of function or missense variants), allowing the consequence of varying degrees of gene perturbation to be assessed. Here gene-based masks have been testing for cis and trans associations with plasma protein levels and complex traits, allowing them to be used in an MR framework. It is perhaps most straight forward to handle gene-level associations as single instuments and interpret Wald ratios, although in some case multiple gene-level instruments can exist for a given exposure.

```{r}
#| label: extract-masks

# Extract gene-based burden mask and calculate Wald ratios
res_masks <- lapply(harmonised_snps, function(x){
  masks <- lapply(x, function(y){
    dat <- y[IV_sets[1:4]] # gene mask results
    # Remove empty results dfs
    dat <- dat[unlist(lapply(dat, function(x){all(!(is.na(x$SNP)))}))]
    
    dat <- do.call("rbind", lapply(dat, function(z){
      z |> dplyr::select(
        mask = test.exposure, 
        mask_gene = SNP, 
        chr = chr.outcome, 
        pos = pos.outcome, 
        exposure, outcome, 
        beta.exposure, beta.outcome, 
        se.exposure, se.outcome, 
        pval.exposure, pval.outcome, 
        cis_trans)
    }))
  })
})

res_masks <- lapply(res_masks, function(x){
  out <- do.call("rbind",x)
  rownames(out) <- NULL
  return(out)})

# Wald ratios
res_masks <- lapply(res_masks, function(x){
  wrs <- apply(x, MARGIN = 1, FUN = function(x){
    TwoSampleMR::mr_wald_ratio(
    b_exp = as.numeric(x["beta.exposure"]),
    b_out = as.numeric(x["beta.outcome"]),
    se_exp = as.numeric(x["se.exposure"]),
    se_out = as.numeric(x["se.outcome"]))
  })

  wrs_df <- do.call("rbind",lapply(wrs, unlist))
  colnames(wrs_df) <- paste("wald_ratio", colnames(wrs_df), sep = "_")

  # Bind to harmonised data
  out <- cbind(x,wrs_df)
  return(out)
})
```

**Gene-level cis-mask associations**

Focusing on aggregate tests combining putative loss of function (protein truncating) and damaging rare varaints, which genes show significant causal effects on outcomes of interest:

```{r}
#| label: fig-maskcis_IVs
#| fig.width: 22
#| fig.height: 8
  
res_masks_cis <- lapply(res_masks, function(x){
  out <- x |> 
    dplyr::filter(cis_trans == "cis",
                            mask == "ptvraredmg") |> 
    dplyr::mutate(wald_ratio_pval_adj = p.adjust(wald_ratio_pval, method = "BH")) |>
    dplyr::arrange(wald_ratio_pval_adj)
  return(out)
})

# Plot Wald ratios for all gene maks associated with an outcome
plot_wald_masks <- function(results_masks, outcome, pthresh){
  dat <- results_masks[[outcome]] |> 
    dplyr::filter(wald_ratio_pval_adj < pthresh) |>
    dplyr::arrange(wald_ratio_b) |>
    dplyr::mutate(
      mask_gene = toupper(mask_gene),
      mask_gene = factor(mask_gene, levels = mask_gene))

  p_wald <- ggplot(dat, aes(x = wald_ratio_b, y = mask_gene)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(colour = "steelblue4", size = 2) +
    geom_errorbarh(aes(xmin = wald_ratio_b-1.96*wald_ratio_se, xmax = wald_ratio_b+1.96*wald_ratio_se),
                   colour = "steelblue4",
                   height = 0.4) +
    labs(x = "Wald Ratio (By/Bx)", y = "Burden test gene \n (instrumenting cis-protein expression)", 
         title = paste("Outcome:", outcome),
         subtitle = "cis-IVs (pLOF and rare damaging)") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
  
  return(p_wald)}

gridExtra::grid.arrange(grobs = lapply(names(res_masks_cis)[-1], plot_wald_masks, results_masks = res_masks_cis, pthresh = 0.1),
                        nrow = 2)
```

**Cis vs trans gene-level mask associations**

```{r}

res_masks_trans <- lapply(res_masks, function(x){
  out <- x |> 
    dplyr::filter(cis_trans == "trans",
                            mask == "ptvraredmg") |> 
    dplyr::mutate(wald_ratio_pval_adj = p.adjust(wald_ratio_pval, method = "BH")) |>
    dplyr::arrange(wald_ratio_pval_adj)
  return(out)
})

# Extract trans IVs for gene showing significant cis-instrumented causal effects
comp_masks_cistrans <- mapply(x = res_masks_cis, y = res_masks_trans, function(x,y){
  x <- x[,c("mask", "mask_gene", "exposure", "outcome", "wald_ratio_b", "wald_ratio_se", "wald_ratio_pval", "wald_ratio_pval_adj")]
  y <- y[,c("mask", "mask_gene", "exposure", "outcome", "wald_ratio_b", "wald_ratio_se", "wald_ratio_pval", "wald_ratio_pval_adj")]
  join_res <- x |> left_join(y, by = c("exposure", "outcome", "mask"), suffix = c("_cis","_trans"))
  return(join_res)
}, SIMPLIFY = F)

# Plot comparison of wald ratios for significant cis or trans IV associations
# lapply(comp_masks_cistrans, function(x){
#   dat <- x |>
#     dplyr::filter(wald_ratio_pval_adj_cis < 0.1 | wald_ratio_pval_adj_trans < 0.1) |>
#     na.exclude()
# 
#   dat |> ggplot(aes(x = wald_ratio_b_cis, y = wald_ratio_b_trans)) +
#     geom_point()
# })
```

### Rare copy number variants (CNVs)

Rare copy number variants (CNVs), including gene deletion (or complete ablation of function through loss of function mutations), and duplication (or higher order) events have been identified. As anticipated, these have been shown to have a monotonic effect on protein expression levels; that is, gene deletions reduce measured gene expression, and duplication events increase gene expression, often substantially [(Milind and Pritchard et al.)](https://pubmed.ncbi.nlm.nih.gov/40832387/). Rare genetic variation with disruptive effects on protein expression, in both directional extremes, present opportunities for protein instrumentation for molecular MR. However, it is of interest whether instrumenting gene expression using deletion and duplication events returns consistent causal estimates for exposure --\> outcome effects, or non-monotonicity in genetic effects exist. The latter would suggest divergent pathways through which cis-perturbatuions of a gene may act to influence a phentoype, or potentially indicate that gene expression is not an appropriate measure of the true functional consequences of such perturbations.

Wald ratios have been derived using published gene-level summary statistics (Milind et al.) for the effect of gene deletions, duplications and pLOF variants (burdan masks) on O'link plasma protein levels and a series of 56 complex traits measured in UKB individuals. Note, Wald ratios were derived for all available cis and trans gene masks and CNV events irrespective of whether the event showed a significant protein expression effect. P-value filtered results are imported below.

Data: https://zenodo.org/records/16800547

```{r}
#| label: import-cnv-results

# Relaxed exposure p-value threshold: <0.05
#load(file = file.path(res_dir, "results_molecular_waldratios_CNVs_alloutcomes_p0.05.rda"))
#res_CNVs <- CNV_relaxed
```

**Magnitude of CNV effects**

```{r}
#| label: cis-cnvs
#| fig.width: 13
#| fig.height: 3

# Merge all results for each outcome into a single dataframe and retain cis variants
# res_CNVs_cis <- lapply(lapply(res_CNVs, function(x){
#   lapply(x, function(x){
#     do.call("rbind",x)})}), function(y){
#       out <- do.call("rbind", y) |>
#         dplyr::filter(cis_trans == "cis")
#       rownames(out) <- NULL
#       return(out)})
# 
# save(res_CNVs_cis, file = file.path(res_dir, "results_molecular_waldratios_CNVs_alloutcomes_p0.05_cis.rda"))

# Check direction of effect of cis-acting duplications, deletions and pLOFs on protein expression

load(file.path(res_dir, "results_molecular_waldratios_CNVs_alloutcomes_p0.05_cis.rda")) #res_CNVs_cis

# Single representitive outcome data to plot exposure associations from w no p-value filtering
load(file.path(res_dir, "results_molecular_waldratios_CNVs_Standing-height.rda")) #wald_ratios
dat_cnvs <- wald_ratios
dat_cnvs_cis <- do.call("rbind",lapply(dat_cnvs, function(x){
  do.call("rbind", x) |> dplyr::filter(cis_trans == "cis")}))
  
cnv_class <- c("milind_plofs" = "pLoF gene burden",
               "milind_lofs" = "LoF gene burden",
               "milind_duplications" = "gene duplication",
               "milind_deletions" = "gene deletion")

cols_cnv <- c("milind_deletions" = "steelblue4",
              "milind_lofs" = "#74ADD1",
              "milind_plofs" = "#ABD9E9",
              "milind_duplications" = "orange")

# Plot variant CNV consequnce call on cis pQTL effect
p7 <- dat_cnvs_cis |> 
  dplyr::mutate(test = factor(test, 
                              levels = c("milind_deletions","milind_lofs","milind_plofs","milind_duplications"))) |>
  ggplot(aes(x = beta.exposure, 
             y = test)) +
  ggridges::geom_density_ridges(aes(fill = test), trim = TRUE) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  scale_y_discrete(labels = cnv_class) +
  scale_fill_manual(values = cols_cnv) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "cis-pQTL expression effect (beta)", y = "CNV annotation")

# pLOF v duplication (effect significant at p<0.05 only)

p8 <- dat_cnvs_cis |>
  dplyr::select(exposure, test, beta.exposure, se.exposure) |>
  tidyr::pivot_wider(
    names_from  = test,
    values_from = c(beta.exposure, se.exposure),
    names_glue  = "{test}_{.value}") |>
  ggplot(aes(x = milind_lofs_beta.exposure, y = milind_duplications_beta.exposure)) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_hline(yintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_abline(slope = -1, intercept = 0, colour ="grey30", linewidth = 0.5) +
  geom_point(size = 2, colour = "steelblue4") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(-3,3)) +
  theme_minimal() +
  labs(x = "cis-LoF burden effect \non protein expression", y = "cis-duplication burden effect \non protein expression")

p9 <- dat_cnvs_cis |>
  dplyr::select(exposure, test, beta.exposure, se.exposure) |>
  tidyr::pivot_wider(
    names_from  = test,
    values_from = c(beta.exposure, se.exposure),
    names_glue  = "{test}_{.value}") |>
  ggplot(aes(x = milind_deletions_beta.exposure, y = milind_duplications_beta.exposure)) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_hline(yintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_abline(slope = -1, intercept = 0, colour ="grey30", linewidth = 0.5) +
  geom_point(size = 2, colour = "steelblue4") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(-3,3)) +
  theme_minimal() +
  labs(x = "cis-deletion burden effect \non protein expression", y = "cis-duplication burden effect \non protein expression")

gridExtra::grid.arrange(grobs = list(p7, p8, p9), nrow = 1, widths = c(0.5,0.25,0.25))

```

**Comparison of Wald Ratios**

Does instrumenting a gene using a rare loss of function mutation or a rare deletion give consistant causal estiamtes?

```{r}
#| label: compare-walds
#| fig.width: 14
#| fig.height: 52

comp_WRs_lofdups <- lapply(res_CNVs_cis, function(x){
  
  if(!(all(c("milind_lofs","milind_duplications") %in% x$test))){
    return(NULL)
    next
  }
  # Data frame of paired wald ratios for each gene
  dat <- x |> dplyr::filter(test == c("milind_lofs","milind_duplications")) |>
    dplyr::select(exposure, test, wald_ratio_b, wald_ratio_se, wald_ratio_pval) |>
    tidyr::pivot_wider(
      names_from  = test,
      values_from = c(wald_ratio_b, wald_ratio_se, wald_ratio_pval),
      names_glue  = "{test}_{.value}") |>
    as.data.frame() |>
    na.exclude()
  
  hettest <- apply(dat, MARGIN = 1, function(y){
    effect_heterogeneity(
      beta_vec = as.numeric(y[2:3]), 
      se_vec = as.numeric(y[4:5]))}) |> 
    t() |> as.data.frame() |>
    dplyr::mutate(Qpval_adj = p.adjust(Qpval, method = "BH"))
  
  dat <- cbind(dat, hettest) |>
    dplyr::mutate(outcome = unique(x$outcome)) |>
    dplyr::arrange(Qpval_adj)
  
  return(dat)
})
    
p10 <- lapply(comp_WRs_lofdups[sapply(comp_WRs_lofdups, is.null) == FALSE], function(x){
  outcome <- unique(x$outcome)
  
  x$col_point <- factor(ifelse(x$Qpval_adj < 0.05, "Het. Q pval <0.05 (FDR)",
                      ifelse((p.adjust(x$milind_lofs_wald_ratio_pval) < 0.05 | p.adjust(x$milind_duplications_wald_ratio_pval) < 0.05), "WR pval <0.05 (FDR)",
                      "NS")), levels = c("Het. Q pval <0.05 (FDR)", "WR pval <0.05 (FDR)", "NS"))
  
  x |> ggplot(aes(x = milind_lofs_wald_ratio_b, y = milind_duplications_wald_ratio_b)) +
    geom_abline(slope = 1, intercept = 0, colour ="grey30", linewidth = 0.5) +
    geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
    geom_hline(yintercept = 0, colour ="grey30", linewidth = 1) + 
    geom_point(aes(colour = col_point, size = col_point)) +
    scale_colour_manual(values = c("tomato","black","grey"), drop = FALSE) +
    scale_size_manual(values = c(3,3,2), drop = FALSE) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "none") +
    labs(x = "Wald Ratio \n(cis-LoF instrumantaion)", 
         y = "Wald Ratio \n(cis-duplication instrumentation)",
         title = paste("Protein ->\n", outcome))
})

gridExtra::grid.arrange(grobs = p10, nrow = 14)

## Pull out instances of effect heterogeneity between LoF and duplication instruments
het_cnv_wrs <- do.call("rbind", lapply(comp_WRs_lofdups[sapply(comp_WRs_lofdups, is.null) == FALSE], function(x){x |> dplyr::filter(Qpval_adj < 0.05)}))
```

**Heterogeneity in Wald Ratios**

```{r}
#| label: compare-walds-hets
#| fig.width: 15
#| fig.height: 20
#| 
# Plot wald ratios of differing cnv instrumentation strategies
plot_wald_cnvs <- function(gene, outcome){
  
  dat <- res_CNVs_cis[[outcome]] |> 
    dplyr::filter(exposure == gene)
  
  p_theme <- theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
  
  labs <- c("milind_lofs" = "cis-LoF", "milind_duplications" = "cis-Dup.", "milind_deletions" = "cis-Del.")
  
  exp_range <- max(abs(dat$beta.exposure) + 1.96*dat$se.exposure) + 0.1
  out_range <- max(abs(dat$beta.outcome) + 1.96*dat$se.outcome) + 0.1
  wr_range <- max(abs(dat$wald_ratio_b) + 1.96*dat$wald_ratio_se) + 0.1
  
  # Exposure
  p_exp <- ggplot(dat, aes(x = beta.exposure, y = labs[test], colour = test)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    scale_colour_manual(values = cols_cnv, labels = cnv_class, name = "") +
    scale_x_continuous(limits = c(-exp_range, exp_range)) +
    geom_errorbarh(aes(xmin = beta.exposure-1.96*se.exposure, xmax = beta.exposure+1.96*se.exposure), height = 0.4) +
    labs(x = "Bx", y = "") +
    p_theme +
    theme(legend.position = "none")
  
  # Outcome
  p_out <- ggplot(dat, aes(x = beta.outcome, y = labs[test], colour = test)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    scale_colour_manual(values = cols_cnv, labels = cnv_class, name = "") +
    scale_x_continuous(limits = c(-out_range, out_range)) +
    geom_errorbarh(aes(xmin = beta.outcome-1.96*se.outcome, xmax = beta.outcome+1.96*se.outcome), height = 0.4) +
    labs(x = "By", y = "") +
    p_theme +
    theme(legend.position = "none")
  
  # Wald Ratio
  p_wald <- ggplot(dat, aes(x = wald_ratio_b, y = labs[test], colour = test)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    scale_colour_manual(values = cols_cnv, labels = cnv_class, name = "") +
    scale_x_continuous(limits = c(-wr_range, wr_range)) +
    geom_errorbarh(aes(xmin = wald_ratio_b-1.96*wald_ratio_se, xmax = wald_ratio_b+1.96*wald_ratio_se), height = 0.4) +
    labs(x = "Wald Ratio (By/Bx)", y = "", 
         title = paste("Exposure:", gene, "\nOutcome:", outcome)) +
    p_theme +
    theme(legend.position = "none", plot.title = element_text(size = 8))

  return(gridExtra::arrangeGrob(grobs = list(p_wald, p_exp, p_out), nrow = 3, heights = c(0.38,0.31,0.31)))
}

gridExtra::grid.arrange(grobs = mapply(plot_wald_cnvs, gene = het_cnv_wrs$exposure, outcome = het_cnv_wrs$outcome, SIMPLIFY = F), nrow = 5)
```
