---
title: "Comparing common and rare single variant and gene aggregate instrumentation strategies for molecular MR"
author: "Aimee Hanson"
date: "`r format(Sys.time(), '%d %B %Y')`"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 4
    toc-location: left
    page-layout: full
    code-fold: true
---

```{r}
#| label: setup
#| message: FALSE
#| echo: FALSE

library(here)
library(dplyr)
library(ggplot2)
library(dotenv)
library(UpSetR)
library(grid)
library(gridExtra)

dotenv::load_dot_env(file = here("config.env"))
data_dir <- Sys.getenv("data_dir")
res_dir <- Sys.getenv("results_dir")

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")

source(here("scripts/04_analysis_proteins/functions.R"))
```

## Introduction

Classically, Mendelian Randomisation (MR) methods utilising trait-associated genetic variants from GWAS studies have employed common polymorphisms (population MAF \> 1%) targeted by genotyping chips, or reliably imputed from reference populations, to instrument modifiable exposures. However, common variants tested in GWAS typically explain a very small fraction of the variability in a measured trait, potentially exhibit pleiotropic effects acted upon by balancing selection or as a consequent of genetic linkage, and are rarely causal. Rare variants, which typically show large biological effects (e.g. through abolishing protein expression) provide a means of more unambiguously instrumenting relevant molecular processes. Comparison of causal estimates derived using differing methods of genetically instrumenting modifiable exposures may enhance the interpretation of the biological mechanisms underlying exposure-outcome relationships. This includes using variants from across the allele frequency spectrum, but also leveraging rare variant aggregate approaches to instrument gene-level perturbations in expression and function.

Here, caual estimates for pairwises combinations of **2937** molecular exposures (O-link plasma proteins) and **11** health related compelex traits (below) have been derived using nine instrumentation strategies, with further instrument splitting into cis (variants within 1MB of the encoding gene start and end site), trans and full variant sets (a total of **27** instrument sets per exposure).

**Outcomes:** Low density lipoprotein levels (LDL), Body Mass Index (BMI), Vitamin D (VitD), Triglycerides (Trig), Glycated Haemoglobin (HbA1c), Mean Platelet Volume (PlatVol), IGF-1 (IGF1), Waist-to-Hip Ratio (WHRadjBMI), Red Blood Cell Erythrocyte Count (RBCCount), Mean Corpuscular Volume (CorpVol) and Systolic Blood Pressure (SBP)

### Instrumental variable (IV) sets

Nine sets of IVs for each protein exposure have been extracted from across two studies:\
GWAS common variants: "Plasma proteomic associations with genetics and health in the UK Biobank" (Sun, B. 2023, Nature).\
WES variants/gene aggregate tests: "Rare variant associations with plasma protein levels in the UK Biobank" (Dhindsa, R. 2023, Nature).\

**Common GWAS**\
Common genome-wide fine-mapped variants from the GPMAP (\>1% MAF) (cis, trans and combined)\
Common genome-wide clumped and pruned variants (\>1% MAF) (cis, trans and combined)

**WES (variants)**\
Common exome-wide (\>1% MAF) (cis, trans and combined)\
Rare exome-wide (0-1% MAF) (cis, trans and combined)\
Ultra-rare exome-wide (0-0.1% MAF) (cis, trans and combined)\
*LD clumping was performed for common variants (for which LD matrices have been derived) only. For rare variants, instrument sets were stored both pre and post pruning to variants independent at r2<0.1.*

**WES (masks)**\
pLOF burden mask (cis, trans and combined)\
missense/low-confidence burden mask (cis, trans and combined)\
pLOF/missense/low-confidence burden mask (cis, trans and combined)\
synonymous burden mask (cis, trans and combined)

## Data import

For direct comparison of the performance of instrument sets, the theshold for pQTL significance has been set at 5x10\^-8 for both rare and common variants. MR results for all exposure-outcome pairs, derived using differing IV sets, and the harmonised summary statistics of those variant sets, are read in below.

Read in MR results from `scripts/04_analysis_proteins/001_performmolecularMR.R` and harmonised summary statistics:

```{r}
#| label: import-mr-results

# ============== Import MR results (final set) ============= #
# Origininal results (rare IVs P<5e-8, common IVs P<5e-8)
files <- list.files(res_dir, pattern = "results_molecular_MR_GW5e-08_EW5e-08_[A-Z]", full.names = T)
files_snps <- list.files(file.path(data_dir, "harmonised"), pattern = "harmonised_studies_GW5e-08_EW5e-08_[A-Z]", full.names = T)

# Repeat with rare variant LD pruning (rare IVs P<5e-8, common IVs P<5e-8)
files_pruned <- list.files(res_dir, pattern = "results_molecular_MR_GW5e-08_EW5e-08_rarepruned", full.names = T)
files_pruned <- files_pruned[-grep("COMBINED", files_pruned)]
files_snps_pruned <- list.files(file.path(data_dir, "harmonised"), pattern = "harmonised_studies_GW5e-08_EW5e-08_rarepruned", full.names = T)
files_snps_pruned <- files_snps_pruned[-grep("COMBINED", files_snps_pruned)]

res_MR <- list()
for(i in 1:length(files)){
  load(files[i])
  res_MR[[i]] <- results_MR
}
names(res_MR) <- sub(".*08_(.*)\\.rda","\\1", files)

### Append results from pruned rare IVs
res_MR_pruned <- list()
for(i in 1:length(files_pruned)){
  load(files_pruned[i])
  res_MR_pruned[[i]] <- lapply(results_MR, function(x){x$IV_set = paste0(x$IV_set, "_pruned"); return(x)})
}
names(res_MR_pruned) <- sub(".*rarepruned_(.*)\\.rda","\\1", files_pruned)

res_MR_joined <- Map(function(x, y) {
  Map(rbind, x, y)
}, res_MR, res_MR_pruned)

rm(res_MR_pruned)

# ============== Import MR results (relaxed set) ============= #
# Less strict thresholding for rare variants w LD pruning (rare IVs P<1e-4, common IVs P<5e-8)
files_relaxed_pruned <- list.files(res_dir, pattern = "results_molecular_MR_rarepruned", full.names = T)
files_snps_relaxed_pruned <- list.files(file.path(data_dir, "harmonised"), pattern = "harmonised_studies_rarepruned", full.names = T)

### Append results from pruned rare IVs
res_MR_relaxed_pruned <- list()
for(i in 1:length(files_relaxed_pruned)){
  load(files_relaxed_pruned[i])
  res_MR_relaxed_pruned[[i]] <- lapply(results_MR, function(x){x$IV_set = paste0(x$IV_set, "_pruned"); return(x)})
}
names(res_MR_relaxed_pruned) <- sub(".*_(.*)\\.rda","\\1", files_relaxed_pruned)

res_MR_relaxed <- Map(function(x, y) {
  Map(rbind, x, y)
}, res_MR, res_MR_relaxed_pruned)

rm(results_MR)
rm(res_MR_relaxed_pruned)
res_MR <- res_MR_joined
rm(res_MR_joined)

# ========== Import harmonised SNP data ========= #
harmonised_snps <- list()
for(i in 1:length(files_snps)){
  load(files_snps[i])
  harmonised_snps[[i]] <- harmonised_studies
}
names(harmonised_snps) <- sub(".*08_(.*)\\.rda","\\1", files_snps)

### Append pruned rare IVs
harmonised_snps_pruned <- list()
for(i in 1:length(files_snps_pruned)){
  load(files_snps_pruned[i])
  harmonised_snps_pruned[[i]] <- lapply(harmonised_studies_pruned, function(x){names(x) <- paste0(names(x), "_pruned"); return(x)})
}
names(harmonised_snps_pruned) <- sub(".*rarepruned_(.*)\\.rda","\\1", files_snps_pruned)

# Less strict thresholding for rare variants w LD pruning (rare IVs P<1e-4, common IVs P<5e-8)
harmonised_snps_relaxed <- list()
for(i in 1:length(files_snps_relaxed_pruned)){
  load(files_snps_relaxed_pruned[i])
  harmonised_snps_relaxed[[i]] <- lapply(harmonised_studies_pruned, function(x){names(x) <- paste0(names(x), "_pruned"); return(x)})
}
names(harmonised_snps_relaxed) <- sub(".*pruned_(.*)\\.rda","\\1", files_snps_relaxed_pruned)

harmonised_snps_relaxed <- Map(function(a1, b1) {
  Map(function(a2, b2) {
    c(a2, b2)   # append entire gene-level lists
  }, a1, b1)
}, harmonised_snps, harmonised_snps_relaxed) # Rare variants pruned with relaxed threshold

harmonised_snps <- Map(function(a1, b1) {
  Map(function(a2, b2) {
    c(a2, b2)   # append entire gene-level lists
  }, a1, b1)
}, harmonised_snps, harmonised_snps_pruned) # Rare variants pruned

rm(harmonised_studies)
rm(harmonised_snps_pruned)
rm(harmonised_studies_pruned)

# --------------------------------------------------------------------------------------- #
# List of LD pruned rare variants:
# P<5e-8
prunein_files <- grep("prune.in", list.files(file.path(data_dir,"sumstats/proteomics/protein_pQTL_pruning"), pattern = "p5e-8_.*", full = TRUE), value = T)
prunein <- do.call("rbind", lapply(prunein_files, function(x){data.table::fread(x, header = F, data.table = F)}))[,1]
# Correct variant name formatting
prunein <- unlist(lapply(strsplit(prunein, ":"), function(x){paste0(x[1],":",x[2],"_",tolower(x[3]),"_",tolower(x[4]))}))

# P<1e-4
prunein_files_relaxed <- grep("prune.in", list.files(file.path(data_dir,"sumstats/proteomics/protein_pQTL_pruning"), pattern = "wes_pQTLs_chr",, full = TRUE), value = T)
prunein_relaxed <- do.call("rbind", lapply(prunein_files_relaxed, function(x){data.table::fread(x, header = F, data.table = F)}))[,1]
# Correct variant name formatting
prunein_relaxed <- unlist(lapply(strsplit(prunein_relaxed, ":"), function(x){paste0(x[1],":",x[2],"_",tolower(x[3]),"_",tolower(x[4]))}))

# --------------------------------------------------------------------------------------- #
# Data filtering

## Retain the following instrument sets for final analysis:
IVsets <- c("dhindsa_exwas_mask_ptv" = "mask:pLOF",
            "dhindsa_exwas_mask_raredmg" = "mask:miss",
            "dhindsa_exwas_mask_ptvraredmg" = "mask:pLOF|miss",
            "dhindsa_exwas_mask_syn" = "mask:synonymous",
            "dhindsa_exwas_variant_ultrarare_pruned" = "exome:ultrarare",
            "dhindsa_exwas_variant_rare_pruned" = "exome:rare",
            "dhindsa_exwas_variant_common" = "exome:common",
            "sun_gwas_variant_common_clumped" = "genome:common")

variant_IVsets <- c("dhindsa_exwas_variant_ultrarare_pruned" = "exome:ultrarare",
                    "dhindsa_exwas_variant_rare_pruned" = "exome:rare",
                    "dhindsa_exwas_variant_common" = "exome:common",
                    "sun_gwas_variant_common_clumped" = "genome:common")

unpruned_IVsets <- c("dhindsa_exwas_mask_ptv" = "mask:pLOF",
                    "dhindsa_exwas_mask_raredmg" = "mask:miss",
                    "dhindsa_exwas_mask_ptvraredmg" = "mask:pLOF|miss",
                    "dhindsa_exwas_mask_syn" = "mask:synonymous",
                    "dhindsa_exwas_variant_ultrarare" = "exome:ultrarare",
                    "dhindsa_exwas_variant_rare" = "exome:rare",
                    "dhindsa_exwas_variant_common" = "exome:common",
                    "sun_gwas_variant_common_clumped" = "genome:common")

# Varaint and gene mask comparison for manuscript
paper_IVsets <- c("dhindsa_exwas_mask_ptvraredmg" = "mask:pLOF|miss",
                  "dhindsa_exwas_variant_ultrarare_pruned" = "exome:ultrarare",
                  "dhindsa_exwas_variant_rare_pruned" = "exome:rare",
                  "dhindsa_exwas_variant_common" = "exome:common",
                  "sun_gwas_variant_common_clumped" = "genome:common")

IV_sets <- names(IVsets)

res_MR <- lapply(res_MR, function(x){
  lapply(x, function(y){
    y |> dplyr::filter(IV_set %in% names(IVsets))
    })
  })

res_MR_relaxed <- lapply(res_MR_relaxed, function(x){
  lapply(x, function(y){
    y |> dplyr::filter(IV_set %in% names(IVsets))
    })
  })

harmonised_snps <- lapply(harmonised_snps, function(x){lapply(x, function(y){y[names(IVsets)]})})
harmonised_snps_relaxed <- lapply(harmonised_snps_relaxed, function(x){lapply(x, function(y){y[names(IVsets)]})})

# Save
#save(res_MR, file = file.path(res_dir, "results_molecular_MR_GW5e-08_EW5e-08_rarepruned_COMBINED.rda"))
#save(res_MR_relaxed, file = file.path(res_dir, "results_molecular_MR_GW5e-08_EW1e-04_rarepruned_COMBINED.rda"))
#save(harmonised_snps, file = file.path(data_dir, "harmonised/harmonised_studies_GW5e-08_EW5e-08_rarepruned_COMBINED.rda"))
#save(harmonised_snps_relaxed, file = file.path(data_dir, "harmonised/harmonised_studies_GW5e-08_EW1e-04_rarepruned_COMBINED.rda"))

# --------------------------------------------------------------------------------------- #
# Variables and aesthetics

n_genes <- length(harmonised_snps[[1]]) # 2937
names_genes <- sub("_.*", "", names(harmonised_snps[[1]]))
n_IVsets <- length(harmonised_snps[[1]][[1]]) #11 IV sets
names_IVsets <- names(harmonised_snps[[1]][[1]])

# 0.05/2937 proteins tested per outcome
n_proteins <- length(res_MR[[1]])
alpha <- signif(0.05/n_proteins,2)

# IVsets
cols_IVsets <- c("darkorchid4","darkorchid3","darkorchid2","orchid",
                 "steelblue4","steelblue1","darkorange2","orange")
names(cols_IVsets) <- IVsets

# Read in GRCh38 gene ranges:
geneinfo <- data.table::fread(file.path(data_dir, "variant_annotation/gene_info_GRCh38.tsv"), data.table = F) |>
  dplyr::arrange(as.numeric(chr),as.numeric(start))
# Read in Olink gene ranges:
geneinfo_olink <- data.table::fread(file.path(data_dir, "sumstats/proteomics/olink_generanges_Sunetal.csv"), data.table = F) |>
  dplyr::rename(
    "chr" = "Gene CHROM",
    "start" = "Gene start",
    "stop" = "Gene end") |>
  dplyr::mutate(
    gene_biotype = "protein_coding",
    gene = `Gene symbol`)

```

## Instrument characteristics

## Comparison of IV sets

**The average number of genetic instruments (pQTLs) per protein across instrument sets**

Rare and common variant instruments (pQTLs) have been partitioned according to their proximity to the instrumented protein-coding gene into cis (+- 1MB from gene start site) and trans (\>1MB distant) sets. A similar number of cis-IVs per protein can be derived from each of the IV sets (typically \<5). Proteins that are instrumented with a large number of rare cis variants fall typically gene dense genomic regions. More trans-IVs are availble in the common-varaint instrument sets (both GWAS and ExWAS derived), suggesting there are fewer rare and ultra-rare variants that associate with protein levels in trans.

```{r}
#| label: fig-numivs
#| cache: TRUE
#| fig.width: 11
#| fig.height: 6.5

# Number of instruments per protein for each IV set
# num_IVs <- lapply(res_MR, function(x){
#   df_out <- data.frame(expand.grid(IV_set = IV_sets, cis_trans = c("cis","trans")))
#   
#   for(i in 1:length(x)){
#     df <- x[[i]] |> dplyr::summarise(unique(nsnp), .by = c(IV_set, cis_trans))
#     colnames(df)[3] <- names(x[i])
#     df_out <- left_join(df_out, df, by = c("IV_set","cis_trans"))
#   }
#   
#   df_out <- df_out |> tidyr::pivot_longer(cols = 3:length(df_out)) |> na.exclude()
#   return(df_out)
# })
# 
# num_IVs <- do.call("rbind", num_IVs) |> dplyr::mutate(outcome = sub(".*_","",name))

#write.table(num_IVs, file = file.path(res_dir, "numIVs_GW5e-08_EW5e-08_rarepruned.txt"), row.names = F, col.names = T, quote = F)
num_IVs <- data.table::fread(file.path(res_dir, "numIVs_GW5e-08_EW5e-08_rarepruned.txt"))

# Plot
p1 <- mapply(FUN = function(x,y){
  x <- x |> dplyr::filter(cis_trans == y) |>
    dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets)) |>
    dplyr::filter(label %in% variant_IVsets)
  
  ggplot(x, aes(x = label, y = value, fill = label)) +
    facet_grid(.~outcome) +
    stat_boxplot(geom ='errorbar', width = 0.2) + 
    geom_boxplot(width = 0.7) +
    scale_y_log10() +
    scale_fill_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          legend.position = "none",
          plot.title = element_text(hjust = 0.5)) +
    labs(x = "IV set",
         y = "Num. instruments", 
         title = paste0("Number of ",y ," instruments per plasma protein (p<5e-8)"))
  
}, x = list(num_IVs,num_IVs), y = c("cis","trans"), SIMPLIFY = FALSE)

gridExtra::grid.arrange(grobs = p1)
```

A GWS theshold for significance is strict for rare variants, which must exhibit very strong effects in order to be detected despite their low minor allele frequency in the population. If this theshold is relaxed (for rare and ultrarare variants only), are more cis and trans IVs available, or common variants continue to show more protein associations in trans?

```{r}
#| label: fig-numivs-relaxed
#| cache: TRUE
#| fig.width: 11
#| fig.height: 7

# Number of instruments per protein for each IV set
# num_IVs_relaxed <- lapply(res_MR_relaxed, function(x){
#   df_out <- data.frame(expand.grid(IV_set = IV_sets, cis_trans = c("cis","trans")))
#   
#   for(i in 1:length(x)){
#     df <- x[[i]] |> dplyr::summarise(unique(nsnp), .by = c(IV_set, cis_trans))
#     colnames(df)[3] <- names(x[i])
#     df_out <- left_join(df_out, df, by = c("IV_set","cis_trans"))
#   }
#   
#   df_out <- df_out |> tidyr::pivot_longer(cols = 3:length(df_out)) |> na.exclude()
#   return(df_out)
# })
# 
# num_IVs_relaxed <- do.call("rbind", num_IVs_relaxed) |> dplyr::mutate(outcome = sub(".*_","",name))

#write.table(num_IVs_relaxed, file = file.path(res_dir, "numIVs_GW5e-08_EW1e-04_rarepruned.txt"), row.names = F, col.names = T, quote = F)
num_IVs_relaxed <- data.table::fread(file.path(res_dir, "numIVs_GW5e-08_EW1e-04_rarepruned.txt"))

# Plot
p1.2 <- mapply(FUN = function(x,y){
  x <- x |> dplyr::filter(cis_trans == y) |>
    dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets)) |>
    dplyr::filter(label %in% variant_IVsets)
  
  ggplot(x, aes(x = label, y = value, fill = label)) +
    facet_grid(.~outcome) +
    stat_boxplot(geom ='errorbar', width = 0.2) + 
    geom_boxplot(width = 0.7) +
    scale_y_log10() +
    scale_fill_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          legend.position = "none",
          plot.title = element_text(hjust = 0.5)) +
    labs(x = "IV set",
         y = "Num. instruments", 
         title = paste0("Number of ",y ," instruments per plasma protein \n(common p<5e-8; rare p<1e-4)"))
  
}, x = list(num_IVs_relaxed,num_IVs_relaxed), y = c("cis","trans"), SIMPLIFY = FALSE)

gridExtra::grid.arrange(grobs = p1.2)
```


**The number of proteins able to be instrumented with pQTLs across instrument sets**

Note, results below are for proteins with \>=3 independent IVs avialble, those considered instrumentable.

Although the power to detect highly significant rare-variant associations is expected to be lower than that for common variants due to to low alternate allele counts, an approximately equivalent percentage (\~25%) of the plasma proteome can be instrumented using GWAS common or ExWAS rare variants in cis (and \~6% with ultra-rare variants). ExWAS common variants provide less coverage, likely because fewer independent common cis-variants will typically be found in the coding region of a gene than within the 1MB surrounding putative regulatory region captured by GWAS. Alternatively, over 50% of the proteome can be instrumented with ExWAS and GWAS derived trans common variants relative to \<12% with rare and ultra-rare variants. This again suggests there are fewer (or less readily detected) rare trans-variant associations with plasma protein levels. This is slightly counterintuitive, as it would be anticipated that stronger rare varaint cis-effects on protein levels would also be detected as strong trans associations with proteins in interacting pathways.\

What are some potential explanations for this?: \
- The common variants that exist at intermediate frequencies in a population (i.e. have not reached fixation) are acted upon by competing balancing selective forces and show a greater number of trans assocaitions across diverse biological processes (i.e. are more plieotropic). Converse to this is the idea that rare variants can only be observed in genomic regions where the DNA sequence has been evolutionarially optimised (conserved) for a highly specific biological mechanism. \
- The nature of the genes instrumented by common and rare variants is fundamentally different. For example the proteins instrumented by common cis-IVs are more often involved in the regulation of diverse cellular pathways (e.g. variants in enhancers impacting transcriptional or cell signalling regulation), where strong-effect rare variants modify protein levels or function directly, with consequences for a smaller numebr of interacting/downstream proteins?
- It's a power thing (explored below)

Questions: \
- What is the relative strength (betas) of cis and trans, rare and common pQTL effects? \
- What is the intersection of the plasma proteins that have rare and common IVs available (significant pQTL associations)?

```{r}
#| label: fig-summary
#| warning: FALSE
#| fig.width: 11
#| fig.height: 6.5

# Number of proteins with at least n cis/trans IV
avail_IVs <- num_IVs |> group_by(IV_set, cis_trans, outcome) |> 
  dplyr::summarise(n_prots_1IV = sum(value >= 1),
                   prop_prots_1IV = sum(value >= 1)/n_proteins,
                   n_prots_3IV = sum(value >= 3),
                   prop_prots_3IV = sum(value >= 3)/n_proteins)

# Plot
p2 <- sapply(c("cis","trans"), function(x){
  avail_IVs |>
    dplyr::filter(IV_set %in% names(variant_IVsets) & cis_trans == x) |>
    dplyr::mutate(labels = factor(variant_IVsets[IV_set], levels = variant_IVsets)) |>
    ggplot(aes(x = labels)) +
    facet_grid(.~outcome) +
    geom_col(aes(y = prop_prots_1IV*100, colour = labels), fill = "white", width = 0.8) +
    geom_col(aes(y = prop_prots_3IV*100, fill = labels), width = 0.8) +
    scale_y_continuous(limits = c(0,100)) +
    scale_fill_manual(values = cols_IVsets) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          plot.title = element_text(hjust = 0.5),
          legend.position = "none") +
    labs(x = "IV set",
         y = "%", 
         title = paste0("Percentage of plasma proteins with ",x, "-IV sets available (with n>=1 or >=3 instruments; p<5e-8)"))}, simplify = FALSE)
  
gridExtra::grid.arrange(grobs = p2)
```

Repeat with relaxed thresholding for rare variants:

```{r}
#| label: fig-summary-relaxed
#| warning: FALSE
#| fig.width: 11
#| fig.height: 7

# Number of proteins with at least n cis/trans IV
avail_IVs_relaxed <- num_IVs_relaxed |> group_by(IV_set, cis_trans, outcome) |> 
  dplyr::summarise(n_prots_1IV = sum(value >= 1),
                   prop_prots_1IV = sum(value >= 1)/n_proteins,
                   n_prots_3IV = sum(value >= 3),
                   prop_prots_3IV = sum(value >= 3)/n_proteins)

# Plot
p2.2 <- sapply(c("cis","trans"), function(x){
  avail_IVs_relaxed |>
    dplyr::filter(IV_set %in% names(variant_IVsets) & cis_trans == x) |>
    dplyr::mutate(labels = factor(variant_IVsets[IV_set], levels = variant_IVsets)) |>
    ggplot(aes(x = labels)) +
    facet_grid(.~outcome) +
    geom_col(aes(y = prop_prots_1IV*100, colour = labels), fill = "white", width = 0.8) +
    geom_col(aes(y = prop_prots_3IV*100, fill = labels), width = 0.8) +
    scale_y_continuous(limits = c(0,100)) +
    scale_fill_manual(values = cols_IVsets) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 55, hjust = 1),
          plot.title = element_text(hjust = 0.5),
          legend.position = "none") +
    labs(x = "IV set",
         y = "%", 
         title = paste0("Percentage of plasma proteins with ",x, "-IV sets available (with n>=1 or >=3 instruments) \n(common p<5e-8; rare p<1e-4)"))}, simplify = FALSE)
  
gridExtra::grid.arrange(grobs = p2.2)

```

**Instrument overlap**

What is the overlap in the proteins that have at least 1 IV across instrument sets?

```{r}
#| label: instrument overlap
#| fig.width: 6
#| fig.height: 3

num_IVs$protein <- sub("_.*","",num_IVs$name)

upset_cis <- list(
  "exome:rare" = num_IVs |> 
    dplyr::filter(outcome == "BMI" & IV_set == "dhindsa_exwas_variant_rare_pruned" & cis_trans == "cis") |> 
    dplyr::pull(protein),
  "ex/genome:common" = num_IVs |> 
    dplyr::filter(outcome == "BMI" & IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "cis") |>
    dplyr::pull(protein) |> unique())

upset_trans <- list(
  "exome:rare" = num_IVs |> 
    dplyr::filter(outcome == "BMI" & IV_set == "dhindsa_exwas_variant_rare_pruned" & cis_trans == "trans") |> 
    dplyr::pull(protein),
  "ex/genome:common" = num_IVs |> 
    dplyr::filter(outcome == "BMI" & IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "trans") |>
    dplyr::pull(protein) |> unique())

# Upset plots
upset(fromList(upset_cis), order.by = "freq", text.scale = 1.5, mainbar.y.label = "cis-instrumented \nproteins (P<5e-8)")
upset(fromList(upset_trans), order.by = "freq", text.scale = 1.5, mainbar.y.label = "trans-instrumented \n proteins (P<5e-8)")

num_IVs_relaxed$protein <- sub("_.*","",num_IVs_relaxed$name)

upset_cis_relaxed <- list(
  "exome:rare" = num_IVs_relaxed |> 
    dplyr::filter(outcome == "BMI" & IV_set == "dhindsa_exwas_variant_rare_pruned" & cis_trans == "cis") |> 
    dplyr::pull(protein),
  "ex/genome:common" = num_IVs_relaxed |> 
    dplyr::filter(outcome == "BMI" & IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "cis") |>
    dplyr::pull(protein) |> unique())

upset_trans_relaxed <- list(
  "exome:rare" = num_IVs_relaxed |> 
    dplyr::filter(outcome == "BMI" & IV_set == "dhindsa_exwas_variant_rare_pruned" & cis_trans == "trans") |> 
    dplyr::pull(protein),
  "ex/genome:common" = num_IVs_relaxed |> 
    dplyr::filter(outcome == "BMI" & IV_set %in% c("dhindsa_exwas_variant_common", "sun_gwas_variant_common_clumped") & cis_trans == "trans") |>
    dplyr::pull(protein) |> unique())

# Upset plots
upset(fromList(upset_cis_relaxed), order.by = "freq", text.scale = 1.5, mainbar.y.label = "cis-instrumented \n proteins \n P<5e-8 common \n P<1e-4 rare")
upset(fromList(upset_trans_relaxed), order.by = "freq", text.scale = 1.5, mainbar.y.label = "trans-instrumented \n proteins \n P<5e-8 common \n P<1e-4 rare")

```

Do rare and common trans-IVs for the same proteins fall in the same genomic regions - i.e. suggesting regulation through a given transgene?

**The strength of IV associations**

What are the relative effect sizes of instruments (pQTLs) in the different IV sets?

First pull the pQTL associations for each IV set into a consistant format:

```{r}
#| label: pulling pQTLs

# Pull exposure associations from harmonised SNP sets

# pqtls <- lapply(seq_len(n_genes), function(gene){
#   out <- lapply(seq_len(n_IVsets), function(set){
# 
#     dfs <- lapply(harmonised_snps, function(x){
#       df <- x[[gene]][[set]]
#       if("gene.outcome" %in% colnames(df)){
#         df$gene <- df$gene.outcome
#       }else{df$gene <- toupper(df$SNP)}
#       
#       if(!("chr.exposure" %in% colnames(df))){
#         df$chr.exposure = sub(":.*","",df$variant)
#         df$pos.exposure = sub(".*:(.*)_.*_.*","\\1",df$variant)
#       }
# 
#       df <- df |> dplyr::select(SNP,
#                                 CHR = chr.exposure,
#                                 POS = pos.exposure,
#                                 EA = effect_allele.exposure,
#                                 OA = other_allele.exposure,
#                                 EAF = eaf.exposure,
#                                 beta = beta.exposure,
#                                 P = pval.exposure,
#                                 exposure,
#                                 cis_trans,
#                                 proximal_gene = gene)
#       return(df)
#     })
# 
#     dfs_join <- do.call("rbind", dfs) |> unique()
#     rownames(dfs_join) <- NULL
#     dfs_join
#   })
#   names(out) <- names_IVsets
#   out
# })
# names(pqtls) <- names_genes

#save(pqtls, file = file.path(data_dir,"sumstats/proteomics/protein_pQTLs_5e-08.rda"))

load(file = file.path(data_dir,"sumstats/proteomics/protein_pQTLs_5e-08.rda")) #pqtls
# 
# # Significant pQTLs, combined across all exposures, for each instrument sets
# pqtls_joined <- lapply(seq_len(n_IVsets), function(set){
#   dfs <- lapply(pqtls, function(x){
#     out <- x[[set]]
#     out$IVset <- names_IVsets[set]
#     return(out)
#   })
#   dfs_join <- do.call("rbind", dfs) |> dplyr::arrange(SNP)
#   rownames(dfs_join) <- NULL
#   return(na.exclude(dfs_join))
# })
# 
# names(pqtls_joined) <- names_IVsets
# 
# # Derive a set of variants that are pQTLs for at least one protein to look up in the raw summary statistics of all others (split by IV sets)
# snp_lookup <- lapply(pqtls_joined, function(x){
#   unique(x$SNP)
# })
# # Format as in summary stats
# snp_lookup[grepl("mask", names(snp_lookup))] <- lapply(snp_lookup[grepl("mask", names(snp_lookup))], function(x){toupper(x)})
# snp_lookup[grepl("exwas_variant", names(snp_lookup))] <- lapply(snp_lookup[grepl("exwas_variant", names(snp_lookup))], 
#                                                                 function(x){gsub(":","-",gsub("_","-",toupper(x)))})
# #save(snp_lookup, file = file.path(data_dir,"sumstats/proteomics/protein_pQTLstolookup_5e-08.rda"))
# # Variant lookup performed as in 002_lookuppQTLs.R (on P1 server)
```

```{r}
#| label: instrument strength
#| fig.height: 4
#| fig.width: 12

pqtls_joined <- sapply(names(unpruned_IVsets), function(x){do.call(rbind, lapply(pqtls, `[[`, x))} |> na.exclude(), simplify = F)
# Retain LD pruned rare variants
pqtls_joined$dhindsa_exwas_variant_rare <- pqtls_joined$dhindsa_exwas_variant_rare |> dplyr::filter(SNP %in% prunein)
pqtls_joined$dhindsa_exwas_variant_ultrarare <- pqtls_joined$dhindsa_exwas_variant_ultrarare |> dplyr::filter(SNP %in% prunein)
names(pqtls_joined)[grep("dhindsa_exwas_variant_rare", names(pqtls_joined))] <- "dhindsa_exwas_variant_rare_pruned"
names(pqtls_joined)[grep("dhindsa_exwas_variant_ultrarare", names(pqtls_joined))] <- "dhindsa_exwas_variant_ultrarare_pruned"

pqtls_joined_variants <- do.call("rbind", pqtls_joined[grep("variant", names(pqtls_joined))]) |>
  tibble::rownames_to_column("IV_set") |>
  dplyr::mutate(IV_set = sub("\\..*","",IV_set),
                cis_trans_true = ifelse(proximal_gene == exposure, "true_cis", cis_trans))

p3 <- pqtls_joined_variants |> 
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |> 
  ggplot(aes(x = IV_set, y = abs(beta))) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Instrument |exposure beta|")

p4 <- pqtls_joined_variants |> dplyr::select(-IV_set) |> unique() |>
  dplyr::mutate(MAF = ifelse(EAF>0.5, 1-EAF,EAF),
                beta_MA = ifelse(EAF>0.5, beta*-1, beta)) |>
  ggplot(aes(x = MAF , y = beta_MA)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_vline(xintercept = c(0.01,0.001), col = "red", lty = 2, lwd = 0.5) +
  theme_minimal() +
  labs(x = "MAF", y = "Instrument exposure beta")

gridExtra::grid.arrange(p3,p4,nrow = 1, widths = c(0.18,0.75), top = "Observed effect size of IVs for protein expression across instrument sets")
```

```{r}
#| label: instrument-strength-split
#| fig.height: 3.5
#| fig.width: 6

# Split cis and trans effects
pqtls_joined_variants |> 
  dplyr::filter(!(cis_trans_true == "")) |>
  dplyr::mutate(label = factor(IVsets[IV_set], levels = IVsets),
                cis_trans_true = factor(cis_trans_true, levels = c("trans","cis","true_cis"))) |>
  ggplot(aes(x = cis_trans_true, y = abs(beta), fill = label)) +
  facet_grid(.~label, scale = "free", space = "free") +
  geom_boxplot() +
  scale_fill_manual(values = cols_IVsets) +
  scale_x_discrete(labels = c("trans","proximal \ncis", "true \ncis")) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "", y = "Instrument |exposure beta|", title = "Effect size of IVs (P<5e-8)")

```

What is the cis effect of trans variants on the genes that they are situated in (for WES variants) or near (common variants)? That is, do all trans variants have a cis effect? 

### Pleiotropy profiles

**Cis and trans plieotropy profiles**

On average, how many molecular traits (protein levels) does a variant significantly associate with across instrument sets (e.g. ultra-rare, rare, common)?

To assess the pleiotropy profiles of variants in each IV set, all significant pQTL associations reported with a given variant (p\<=5e-8; or p<=1e-4 for rare varaints at a relaxed threshod) were extracted from the O-link protein GWAS/ExWAS summary statistics.

```{r}
#| label: pleiotropy-matrix

lookup_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_*", full.names = TRUE)

lookup_masks_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_dhindsa_exwas_mask_*", full.names = TRUE)
lookup_exome_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_dhindsa_exwas_variant_*", full.names = TRUE)
lookup_genome_files <- list.files(file.path(data_dir, "sumstats/proteomics"), pattern = "pQTL_lookup_sun_gwas_variant_*", full.names = TRUE)

# Pleiotropy matrices (for each IV set)
# How many proteins do IVs associated with?
lookup_pQTLs <- lapply(lookup_files, function(x){data.table::fread(x)})
names(lookup_pQTLs) <- sub(".*pQTL_lookup_(.*)\\.tsv","\\1",lookup_files)

# Format to keep significant variant-protein associations
lookup_pQTLs <- lapply(lookup_pQTLs, function(x){
  x <- as.data.frame(x)
  if(!("GENE" %in% colnames(x))){x$GENE <- NA}
  
  out <- data.frame(variant = sub(".*\\.tsv:","",x[,1]),
                    protein_target = x$protein_target,
                    proximal_gene = x$GENE,
                    pvalue = x[,which(colnames(x) %in% c("P","pValue"))],
                    cis_trans = x$cis_trans) |>
    dplyr::arrange(variant) |>
  unique()
  
  return(out)})

# Retain LD pruned common variants
#### [FIX]

# Relaxed theshold for rare variants (P<1e-4)
lookup_pQTLs_relaxed <- lapply(lookup_pQTLs, function(x){
  x |> dplyr::filter(!(cis_trans == "")) # Exclude composite protein targets with no cis_trans assignment
})

# Retain LD pruned rare variants
lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein_relaxed))))
lookup_pQTLs_relaxed$dhindsa_exwas_variant_ultrarare_pruned <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_ultrarare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein_relaxed))))
lookup_pQTLs_relaxed <- lookup_pQTLs_relaxed[names(IVsets)]
#lapply(lookup_pQTLs_relaxed, function(x){max(x$pvalue)})

lookup_pQTLs_relaxed_both <- lookup_pQTLs_relaxed # pre thresholding common exome varaints
# Threshold common exome variants
lookup_pQTLs_relaxed$dhindsa_exwas_variant_common <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |> dplyr::filter(pvalue <= 5e-8)

# Retain associations at 5x10^-8 (note gene masks are already thresholded at 1x10^-8)
lookup_pQTLs <- lapply(lookup_pQTLs, function(x){
  x |> dplyr::filter(pvalue <= 5e-8 & !(cis_trans == "")) # Exclude composite protein targets with no cis_trans assignment
})
# Prune rare variants
lookup_pQTLs$dhindsa_exwas_variant_rare_pruned <- lookup_pQTLs$dhindsa_exwas_variant_rare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein))))
lookup_pQTLs$dhindsa_exwas_variant_ultrarare_pruned <- lookup_pQTLs$dhindsa_exwas_variant_ultrarare |> 
  dplyr::filter(variant %in% gsub("_","-",sub(":","-",toupper(prunein))))
lookup_pQTLs <- lookup_pQTLs[names(IVsets)]
#lapply(lookup_pQTLs, function(x){max(x$pvalue)})

# Annotate proximal gene and protein target with the chromosome they are on
lookup_pQTLs <- lapply(lookup_pQTLs, function(x){
  x <- x |> dplyr::mutate(protein_target_chr = sub(";.*","",geneinfo_olink[match(x$protein_target, geneinfo_olink$`Assay Target`), "chr"]),
                          proximal_gene_chr = geneinfo[match(gsub("'", "", x$proximal_gene), geneinfo$gene), "chr"])
  return(x)
})

lookup_pQTLs_relaxed <- lapply(lookup_pQTLs_relaxed, function(x){
  x <- x |> dplyr::mutate(protein_target_chr = sub(";.*","",geneinfo_olink[match(x$protein_target, geneinfo_olink$`Assay Target`), "chr"]),
                          proximal_gene_chr = geneinfo[match(gsub("'", "", x$proximal_gene), geneinfo$gene), "chr"])
  return(x)
})

# Summarise associations per variant (pQTL)
pQTLs_assocs <- mapply(function(x, y){
  x <- x |> dplyr::group_by(variant, proximal_gene) |> 
    dplyr::summarise(n_assocs = n(), 
                     n_cis = length(which(cis_trans == "cis")),
                     n_trans = length(which(cis_trans == "trans")))
  x$IV_set <- y
  return(x)
}, x = lookup_pQTLs, y = as.list(names(lookup_pQTLs)), SIMPLIFY = F)

# Summarise the number of cis and trans genes each variant associates with
pQTLs_assocs_relaxed <- mapply(function(x, y){
  x <- x |> dplyr::group_by(variant, proximal_gene) |> 
    dplyr::summarise(n_assocs = n(), 
                     n_cis = length(which(cis_trans == "cis")),
                     n_trans = length(which(cis_trans == "trans")))
  x$IV_set <- y
  return(x)
}, x = lookup_pQTLs_relaxed, y = as.list(names(lookup_pQTLs_relaxed)), SIMPLIFY = F)

```

```{r}
#| label: cis-v-trans-pleiotropy
#| fig.height: 4
#| fig.width: 8

pqtls_assocs_joined <- do.call("rbind", pQTLs_assocs) |>
  dplyr::filter(grepl("variant",IV_set)) |>
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |>
  dplyr::group_by(IV_set) |> 
  dplyr::mutate(label = ifelse(n_trans == max(n_trans) & grepl("exome",IV_set), paste(variant,proximal_gene), NA)) |> dplyr::ungroup()

p5 <- pqtls_assocs_joined  |>
  ggplot(aes(x = IV_set, y = n_assocs)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations")

p6 <- pqtls_assocs_joined  |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-pQTL associations", y = "Number of trans-pQTL associations")

gridExtra::grid.arrange(p5,p6,nrow = 1, widths = c(0.2,0.7), top = "Pleiotropy profile of IV sets (P<5e-8)")
```

```{r}
#| label: cis-v-trans-pleiotropy-relaxed
#| fig.height: 4
#| fig.width: 8

pqtls_assocs_joined_relaxed <- do.call("rbind", pQTLs_assocs_relaxed) |>
  dplyr::filter(grepl("variant",IV_set)) |>
  dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets)) |>
  dplyr::group_by(IV_set) |> 
  dplyr::mutate(label = ifelse(n_trans == max(n_trans) & grepl("exome",IV_set), paste(variant,proximal_gene), NA)) |> dplyr::ungroup()

p5.2 <- pqtls_assocs_joined_relaxed  |>
  ggplot(aes(x = IV_set, y = n_assocs)) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "Number of pQTL associations")

p6.2 <- pqtls_assocs_joined_relaxed  |>
  ggplot(aes(x = n_cis, y = n_trans, colour = IV_set)) +
  facet_wrap(.~IV_set, nrow = 2) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = label), size = 3) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Number of cis-pQTL associations", y = "Number of trans-pQTL associations")

gridExtra::grid.arrange(p5.2,p6.2,nrow = 1, widths = c(0.2,0.7), top = "Pleiotropy profile of IV sets (common: P<5e-8; rare: P<1e-4)")

```

```{r}
#| label: pleiotropy-vs-pval
#| fig.height: 5
#| fig.width: 6

# Distribution of p-values
do.call("rbind", lookup_pQTLs_relaxed_both[names(variant_IVsets)]) |>
  tibble::rownames_to_column("IV_set") |>
  dplyr::mutate(IV_set = factor(variant_IVsets[sub("\\..*","",IV_set)], levels = variant_IVsets)) |>
  ggplot(aes(x = -log10(pvalue), group = cis_trans)) +
  geom_density(aes(fill = cis_trans), size = 0.5, alpha = 0.45) +
  geom_vline(xintercept = -log10(5e-8), colour = "grey30") +
  geom_vline(xintercept = -log10(1e-4), colour = "grey30") +
  scale_fill_manual(name = "", values = c("tomato","steelblue1")) +
  facet_wrap(IV_set ~ ., nrow = 4) +
  coord_cartesian(xlim = c(0,50)) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Genome-wide common variants are thresholded at p<5x10-8
  
```

Can we take genes that have both rare and common cis-IVs or trans-IVs available (using the relaxed p-value theshold for rare variants) and look at how many other genes, on average, IVs in these genes associate with? We might hypothesise that perturbations to a gene (either via a common or rare variant) will exhibit similar vertical pleiotropic effects on genes downstream. Although there is differences in the power to detect rare vs common variant effects, at relaxed p-value thresholds many more rare trans QTLs can be detected than common (pontentially due to the larger effect sizes of rare trans IVs), and almost all O'link proteins have rare-trans associations (so we can see them). But *how many* other proteins do these variants associate with (i.e. how specific are they?)? Lower specificity of common over rare IVs may suggest trans effects that are mediated through LD (genetic confounding) when instrumented with common variants, which is less problematic for rare variants:

```{r}
#| label: rare-v-common-pleiotropy
#| fig.height: 4
#| fig.width: 12

# Using the WES IVs (common IVs P<5e-8, rare IVs P<1e-4), as these can be directly annotated to genes:

# Collapse variant level cis and trans pQTL scores to the gene level (average)
pqtls_collapsed_relaxed <- pqtls_assocs_joined_relaxed |>
  dplyr::filter(IV_set %in% c("exome:rare","exome:common")) |>
  dplyr::group_by(proximal_gene, IV_set) |>
  dplyr::summarise(mean_n_assocs = mean(n_assocs),
                   mean_n_cis = mean(n_cis),
                   sd_n_cis = sd(n_cis),
                   mean_n_trans = mean(n_trans),
                   sd_n_trans = sd(n_trans))


p7 <- pqtls_collapsed_relaxed[,c(1,2,4,5)] |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(mean_n_cis, sd_n_cis),
    names_glue = "{IV_set}_{.value}"
  ) |>
  dplyr::filter(!(is.na(`exome:rare_mean_n_cis`)) & !(is.na(`exome:common_mean_n_cis`))) |>
  ggplot(aes(x = `exome:rare_mean_n_cis`, y = `exome:common_mean_n_cis`)) +
  geom_abline(intercept = 0, slope = 1, colour = "red") +
  geom_point() +
  scale_x_continuous(limits = c(0,10)) +
  theme_minimal() +
  labs(x = "Ave num. cis-gene associations \n (across rare IVs in gene)",
       y = "Ave num. cis-gene associations \n (across common IVs in gene)")


p8 <- pqtls_collapsed_relaxed[,c(1,2,6,7)] |>
  tidyr:: pivot_wider(
    names_from = IV_set,
    values_from = c(mean_n_trans, sd_n_trans),
    names_glue = "{IV_set}_{.value}"
  ) |>
  dplyr::filter(!(is.na(`exome:rare_mean_n_trans`)) & !(is.na(`exome:common_mean_n_trans`))) |>
  ggplot(aes(x = `exome:rare_mean_n_trans`, y = `exome:common_mean_n_trans`)) +
  geom_abline(intercept = 0, slope = 1, colour = "red") +
  geom_point() +
  scale_x_continuous(limits = c(0,250)) +
  theme_minimal() +
  labs(x = "Ave num. trans-gene associations \n (across rare IVs in gene)",
       y = "Ave num. trans-gene associations \n (across common IVs in gene)")

# Does a gene having aveage higher cis-pleiotropy result in an average higher trans-pleiotropy?

# Filter to genes that have both common and rare IVs in them
shared_genes <- intersect(unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_common$proximal_gene),
                          unique(lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned$proximal_gene))

p8.2 <-  pqtls_collapsed_relaxed |> dplyr::filter(proximal_gene %in% shared_genes) |>
  ggplot(aes(x = mean_n_cis, y = mean_n_trans, group = IV_set, colour = IV_set)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_log10() +
  scale_colour_manual(name = "", values = cols_IVsets) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(x = "Ave num. cis-gene associations \n (across IVs in gene)",
       y = "Ave num. trans-gene associations \n (across IVs in gene)")

gridExtra::grid.arrange(p7, p8, p8.2, nrow = 1, top = "Average pleiotropy of rare or common IVs in a given gene")

```

Are the higher numbers of cis-gene and trans-gene (within the same chromosome) associations observed for common IVs driven by variants that fall in gene dense regions (and is this more striking than for rare IVs)? Although this may represent proximal regulation, it could also be an indication of confounding through LD, particularly if rare and common vartiants show disparate trends:

```{r}
#| label: pleiotropy-vs-genedensity
#| fig.height: 4
#| fig.width: 16
#| cache: TRUE

# Annotate WES pQTLs (common IVs P<5e-8, rare IVs P<1e-4) with numbe.r of genes in proximity
count_genes_1mb <- function(geneinfo, variant_ids){
  out <- data.frame(variant = variant_ids,
                    chr = as.numeric(sub("-.*","",variant_ids)),
                    pos = as.numeric(sub("^.*-(.*)-.*-.*$","\\1",variant_ids))) |>
    dplyr::arrange(chr, pos)
  
  for (i in 1:nrow(out)){
    # 1Mb window about variant
    window_min <- out$pos[i] - 1000000
    window_max <- out$pos[i] + 1000000 
    
    # Extract genes in window (overlap okay)
    window_genes <- geneinfo |> 
      dplyr::filter(as.numeric(chr) == out$chr[i],
                    (as.numeric(start) >= window_min | as.numeric(stop) >= window_min) & (as.numeric(start) <= window_max | as.numeric(stop) <= window_max),
                    gene_biotype == "protein_coding")
    out$n_genes[i] = length(unique(window_genes$gene))
  }
  
  return(out[,c("variant","n_genes")])
}

variant_ids <- pqtls_assocs_joined_relaxed |> dplyr::filter(IV_set %in% c("exome:rare","exome:common")) |> dplyr::pull(variant)

pqtls_assocs_joined_relaxed <- pqtls_assocs_joined_relaxed |> 
  left_join(count_genes_1mb(geneinfo = geneinfo, variant_ids = variant_ids), by = "variant") |>
  dplyr::rename("n_codinggenes_1mb" = "n_genes")

pqtls_assocs_joined_relaxed <- pqtls_assocs_joined_relaxed |> 
  left_join(count_genes_1mb(geneinfo = geneinfo_olink, variant_ids = variant_ids), by = "variant") |>
  dplyr::rename("n_olinkgenes_1mb" = "n_genes")

p9 <- pqtls_assocs_joined_relaxed |> 
  dplyr::filter(IV_set %in% c("exome:rare","exome:common") &
                  proximal_gene %in% paste0("'", geneinfo_olink$gene, "'") & 
                  proximal_gene %in% shared_genes) |>
  ggplot(aes(x = n_olinkgenes_1mb, y = n_cis, colour = IV_set)) +
  geom_hline(yintercept = c(0,5), col = "grey70") + 
  geom_jitter(width = 0.2, height = 0) +
  facet_grid(IV_set ~ .) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "Number of O'link protein-coding genes within 1Mb of instrument", 
       y = "Number of pQTL associations with \n O'link proteins within 1Mb",
       title = "IVs in genes with both common and rare WES instruments")

p10 <- pqtls_assocs_joined_relaxed |> 
  dplyr::filter(IV_set %in% c("exome:rare","exome:common") &
                  proximal_gene %in% paste0("'", geneinfo_olink$gene, "'") & 
                  proximal_gene %in% shared_genes) |>
  ggplot(aes(x = n_codinggenes_1mb, y = n_cis, colour = IV_set)) +
  geom_hline(yintercept = c(0,5), col = "grey70") + 
  geom_jitter(width = 0.2, height = 0) +
  facet_grid(IV_set ~ .) +
  scale_colour_manual(values = cols_IVsets) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "Number of protein-coding genes within 1Mb of instrument", 
       y = "Number of pQTL associations with \n O'link proteins within 1Mb",
       title = "IVs in genes with both common and rare WES instruments")

gridExtra::grid.arrange(p9, p10, nrow = 1, top = "Average pleiotropy of rare or common IVs in a given gene")

```

How often do genes instrumented with either common or rare variants show consistant trans effects? We may anticipate that true vertical pleiotropy is detectable through both instrumentation strategies, whereas problematic trans effects, arising due to LD, will result in associations seen when using common variant instrumentation alone...

What proportion of cis and trans associations are shared between rare and common instrumentation strategies (on the gene level)?

```{r}
#| label: shared-signals
#| eval: FALSE

# Define:
# cis (< +- 1mb), cis_chr (on the same chr), trans (> +- 1mb), trans_genome (on another chr)
# cis/trans signals shared by rare and common varinat instrumentaiton of genes (WES IVs only)

shared_signals <- dplyr::full_join(
  lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned |>
    dplyr::select(protein_target, proximal_gene, cis_trans, protein_target_chr, proximal_gene_chr) |> unique() |>
    dplyr::mutate("exome:rare" = TRUE),
  lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |>
    dplyr::select(protein_target, proximal_gene, cis_trans, protein_target_chr, proximal_gene_chr) |> unique() |>
    dplyr::mutate("exome:common" = TRUE), by = c("protein_target", "proximal_gene", "cis_trans", "protein_target_chr", "proximal_gene_chr")) |>
  dplyr::filter(!is.na(proximal_gene_chr),
                protein_target != gsub("'","",proximal_gene)) |>
  dplyr::mutate(
    shared = (`exome:rare` == `exome:common`),
    cis_1mb = (cis_trans == "cis"),
    cis_chr = (proximal_gene_chr == protein_target_chr),
    trans_chr = (cis_trans == "trans" & proximal_gene_chr == protein_target_chr),
    trans_genome = (proximal_gene_chr != protein_target_chr))

shared_signals$shared[is.na(shared_signals$shared)] <- FALSE

## Plot this
# shared_signals |> ggplot(aes(x = trans_genome, fill = shared)) +
#   geom_bar(position = "fill")

## Loop through genes containg both common and rare instruments and quantify shared trans signals

shared_transsignals <- data.frame(proximal_gene = shared_genes, 
                                  n_trans_total = NA,
                                  n_trans_shared = NA,
                                  n_trans_rare = NA,
                                  n_trans_common = NA,
                                  genes_shared = NA,
                                  genes_rare = NA,
                                  genes_common = NA)
rownames(shared_transsignals) <- shared_genes

for (gene in shared_genes){
  trans_rare <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_rare_pruned |> 
    dplyr::filter(proximal_gene == gene) |>
    dplyr::pull(protein_target) |> unique()
  trans_rare <- trans_rare[trans_rare != gsub("'","",gene)]
  
  trans_common <- lookup_pQTLs_relaxed$dhindsa_exwas_variant_common |> 
    dplyr::filter(proximal_gene == gene) |>
    dplyr::pull(protein_target) |> unique()
  trans_common <- trans_common[trans_common != gsub("'","",gene)]

  shared_transsignals[gene, "n_trans_total"] <- length(unique(c(trans_rare, trans_common)))
  shared_transsignals[gene, "n_trans_shared"] <- length(intersect(trans_rare, trans_common))
  shared_transsignals[gene, "n_trans_rare"] <- length(setdiff(trans_rare, trans_common))
  shared_transsignals[gene, "n_trans_common"] <- length(setdiff(trans_common, trans_rare))
  shared_transsignals[gene, "genes_shared"] <- paste(intersect(trans_rare, trans_common), collapse = ";")
  shared_transsignals[gene, "genes_rare"] <- paste(setdiff(trans_rare, trans_common), collapse = ";")
  shared_transsignals[gene, "genes_common"] <- paste(setdiff(trans_common, trans_rare), collapse = ";")
}

# Bind number of gene within +- 1Mb
shared_transsignals <- shared_transsignals |>
  left_join(pqtls_assocs_joined_relaxed[,c("proximal_gene", "n_codinggenes_1mb")] |>
              dplyr::arrange(desc(n_codinggenes_1mb)) |> 
              dplyr::filter(duplicated(proximal_gene) == FALSE), by = "proximal_gene")

# Proportion of the total trans hits seen with variants in a gene which are detected using both common and rare instruments?
shared_transsignals$prop_shared = shared_transsignals$n_trans_shared/shared_transsignals$n_trans_total

# shared_transsignals |> dplyr::filter(n_trans_total != 0) |>
#   ggplot(aes(x = n_codinggenes_1mb, y = prop_shared)) +
#   geom_point()


```

### IV heterogenity

Heterogenity in SNP effects can be a result of variants acting within differing biological pathways to infuence an outcome (pleiotropy), and violates the third IV assumption (an instrument only influences an outcome via its effect on an exposure). Do rare and common, cis and trans IVs exhibit differing heterogenity in causal effect estimates? When rare and common instruments are combined into one MR analysis, is there a relationship between exposure effect and SNP heterogenity (definied as the deviation between the wald ratio for an IV and the IVW estimated from the metanalysis of all IVs)?...

```{r}
#| label: IV-heterogenity
#| fig.height: 3.5
#| fig.width: 6

# Read in results of mr_heterogenity() for common and rare variant IV sets
files_het <- list.files(res_dir, pattern = "heterogenity", full.names = T)
het <- do.call("rbind", lapply(files_het, function(x){data.table::fread(x)}))

het |> 
  dplyr::filter(method == "Inverse variance weighted" &
                  n_snps > 5) |>
  dplyr::mutate(IV_set = factor(IVsets[variant_set], levels = IVsets)) |>
  ggplot(aes(x = variant_type, y = -log10(Q_pval))) +
  facet_grid(. ~ IV_set) +
  geom_boxplot(aes(fill = IV_set)) +
  scale_fill_manual(values = cols_IVsets) +
  scale_y_log10() +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + 
  labs(x = "", y = "-log10(Q p-value)", title = "Heterogenity in causal estimates (proteins with >5 IVs)")

# Restrict to the same proteins across instrument sets (sensitivity analysis)
```

## Causal effect estimates

### Comparison of causal effects

**Significant causal effects detected across instrument types**

```{r}
#| label: full-results
#| fig.width: 18
#| fig.height: 33
#| cache: TRUE

# P-value threshold for cis-MR (multiple testing correction for proteins instrumentable with at least 3 IVs in cis)
alpha_cis3 <- 0.05/max(avail_IVs_relaxed |> dplyr::filter(cis_trans == "cis") |> pull(n_prots_3IV)) # 5x10^-5

# Plot "manhattan" for protein causal effects on a given outcome, restricting to IV sets of at least 3 variants (masks can have 1)
plot_fullresults <- function(results_MR, outcome, nsnps = 3, cistrans = "cis"){
  df <- do.call("rbind", lapply(results_MR[[outcome]], function(x){
    dat <- x |> 
      dplyr::filter(method %in% c("Wald ratio", "Inverse variance weighted"),
                    cis_trans == cistrans,
                    (IV_set %in% names(variant_IVsets) & nsnp >= nsnps) | !(IV_set %in% names(variant_IVsets)),
                    IV_set %in% names(paper_IVsets))
  }))
  # Format for plotting
  df <- df |>
    dplyr::arrange(pval) |>
    dplyr::mutate(
      set = IVsets[IV_set],
      col = factor(ifelse(pval <= 0.05, set, NA), levels = IVsets),
      label = ifelse(pval <= alpha, exposure, NA))
    #dplyr::mutate(label = ifelse(duplicated(label), NA, label))
  
  df |> dplyr::arrange(desc(pval)) |>
    ggplot(aes(x = exposure, y = -log10(pval))) +
    geom_hline(yintercept = -log10(alpha_cis3), col = "grey50", lty = 2) +
    geom_point(aes(colour = col), size = 2.5) +
    ggrepel::geom_text_repel(aes(label = label)) +
    scale_colour_manual(name = "", values = cols_IVsets, na.value = "grey90") +
    theme_minimal() +
    theme(legend.position = "bottom",
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 10),
          axis.title = element_text(size = 12),
          panel.grid = element_blank(),
          plot.title = element_text(hjust = 0.5, size = 14),
          legend.text = element_text(size = 12)) +
    guides(colour = guide_legend(nrow = 1)) +
    labs(x = "Protein exposure", y = "IVW/WR -log10(P)", title = outcome)
}

gridExtra::grid.arrange(grobs = 
  lapply(X = names(res_MR_relaxed), FUN = plot_fullresults, results_MR = res_MR_relaxed, nsnps = 3, cistrans = "cis"), nrow = 11)


```

How many new highly significant causal associations are detected with rare variants or gene based masks that a. cannot be instrumented in cis with common variants or b. show no causal effect when using common-IVs (suggesting rare variants perturb the gene in a more relevant fashion)?:

```{r}
#| label: new-causal-effects
#| cache: TRUE
#| include: FALSE

# Pull significant MR results (at p<0.05 for comparison of cis IV sets)
# >=3 IVs for variant-based instruments, 1 sufficient for masks

sig_MR_relaxed <- lapply(res_MR_relaxed, function(x){
  do.call("rbind", lapply(x, function(y){
    dat <- y |> 
      dplyr::filter(method %in% c("Wald ratio", "Inverse variance weighted"),
                    cis_trans == "cis",
                    (IV_set %in% names(variant_IVsets) & nsnp >= 3) | !(IV_set %in% names(variant_IVsets)),
                    IV_set %in% names(paper_IVsets),
                    pval <= alpha_cis3)})) |> # multiple testing correction
    dplyr::mutate(IVset = IVsets[IV_set])})

# All common IV results
res_MR_common <- lapply(res_MR_relaxed, function(x){
  do.call("rbind",lapply(x, function(y){
    dat <- y |> 
      dplyr::filter(method == "Inverse variance weighted",
                    cis_trans == "cis",
                    nsnp >= 3,
                    IV_set %in% c("dhindsa_exwas_variant_common","sun_gwas_variant_common_clumped")) |>
      dplyr::mutate(IVset = IVsets[IV_set])}))})

# Upset plot of overlap in support for exposure --> outcome association
upset_sig_cis <- lapply(sig_MR_relaxed, function(x){
  list("mask:pLOF|miss" = x |> dplyr::filter(IVset == "mask:pLOF|miss") |>  dplyr::pull(exposure),
       "exome:ultrarare" = x |> dplyr::filter(IVset == "exome:ultrarare") |>  dplyr::pull(exposure),
       "exome:rare" = x |> dplyr::filter(IVset == "exome:rare") |>  dplyr::pull(exposure),
       "exome:common" = x |> dplyr::filter(IVset == "exome:common") |>  dplyr::pull(exposure),
       "genome:common" = x |> dplyr::filter(IVset == "genome:common") |>  dplyr::pull(exposure))})

# Total significant (P<5e-05) causal protein --> outcome IVW estimates provided by rare IVs (variants and masks)
a <- sapply(upset_sig_cis, function(x){length(unique(c(x$`mask:pLOF|miss`, x$`exome:rare`, x$`exome:ultrarare`)))})
c(a, TOTAL = sum(a)) #65

# How many of these are undetected by common IV methods at this threshold
b <- sapply(upset_sig_cis, function(x){length(setdiff(c(x$`mask:pLOF|miss`, x$`exome:rare`, x$`exome:ultrarare`), c(x$`exome:common`, x$`genome:common`)))})
c(b, TOTAL = sum(b)) #55

# How many of these lack >=3 common variant cis-instruments for the tested exposure?
## Exposure:outcome pairs significant with rare instrumentation
a_pairs <- lapply(sig_MR_relaxed, function(x){x |> dplyr::filter(IVset %in% c("mask:pLOF|miss", "exome:rare", "exome:ultrarare")) |> pull(pair) |> unique()})
## Exposure:outcome pairs with sufficient common cis-IVs to test
common_pairs <- num_IVs_relaxed |> dplyr::filter(cis_trans == "cis", grepl("common", IV_set), value >= 3) |> pull(name) |> unique()
d <- sapply(a_pairs, function(x){sum(!(x %in% common_pairs))})
c(d, TOTAL=sum(d)) #35
## What are these pairs?
# sapply(a_pairs, function(x){x[!(x %in% common_pairs)]})

## How many have sufficient common cis-IVs but lack evidence for a causal association (at differing p-value thresholds)
## Common cis-IV results for signiciant rare cis-IV protein --> outcome results:
overlap <- lapply(res_MR_common, function(x){x |> dplyr::filter(pair %in% do.call("c", a_pairs)) |> 
    dplyr::arrange(pval) |> dplyr::filter(duplicated(pair) == FALSE)})
# Of these, pairs significant at differing theshold
e <- unlist(lapply(overlap, function(x){table(cut(x$pval, breaks = c(0,alpha_cis3,0.05,1)))[2]})) # 0.05>p>5x10^-5
c(e, TOTAL=sum(e)) # 5
f <- unlist(lapply(overlap, function(x){table(cut(x$pval, breaks = c(0,alpha_cis3,0.05,1)))[3]})) # p>0.05
c(f, TOTAL = sum(f))

# 65 causal protein --> outcome relationships when instrumenting using rare cis variants or masks (p<5x10^-5)
# 55 undetected when using common cis-IVs at this threshold
## Of these 55:
### 35 lacked sufficient common-IVs and were not tested (64%)
### 5 are significant using common-IVs at a relaxed threshold 0.05>p>5x10^-5 (9%)
### 15 were tested but using common-IVs but no causal associaton was detected  (27%)

## Which are these 15 discordant results?
# do.call("rbind",lapply(overlap, function(x){x |> dplyr::filter(pval > 0.05)}))

## Example of concordant results from instrument sets (PRSS53 in BMI)
## Example of rare eviedence only... GMPR
## Example of discordance...

## Example from gene instrumented with rare IVs (GMPR?)
```

**Example results**

Plot effect estimates across IV sets for a given exposure-outcome pair

```{r}
#| label: plot-mr-results
#| fig.width: 8
#| fig.height: 4.5

plot_results <- function(exposure, outcome){
  
  pair <- paste(exposure, outcome, sep = "_")
  # Individual SNP results
  dat_snps <- do.call("rbind",lapply(harmonised_snps_relaxed[[outcome]][[pair]][names(paper_IVsets)], function(x){
    x |> dplyr::filter(cis_trans == "cis", mr_keep == TRUE) |>
      dplyr::select(SNP, beta.exposure, beta.outcome, pval.exposure, pval.outcome, se.exposure, se.outcome)})) |>
    tibble::rownames_to_column("IV_set") |>
    dplyr::mutate(IV_set = factor(IVsets[sub("\\..*","",IV_set)], levels = IVsets)) |>
    dplyr::filter(IV_set %in% paper_IVsets) |>
    dplyr::arrange(desc(IV_set))
  
  # MR estimates
  dat_res <- res_MR_relaxed[[outcome]][[pair]] |> 
    dplyr::filter(cis_trans == "cis", method %in% c("Wald ratio", "Inverse variance weighted"), IV_set %in% names(paper_IVsets)) |>
    dplyr::mutate(IV_set = factor(IVsets[IV_set], levels = IVsets))
  
  ## IVW plot
  p_slope <- dat_snps |> ggplot(aes(x = beta.exposure, y = beta.outcome, colour = IV_set)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_errorbar(aes(xmin = beta.exposure - 1.96*se.exposure, xmax = beta.exposure + 1.96*se.exposure), height = 0, alpha = 0.3) +
    geom_errorbar(aes(ymin = beta.outcome - 1.96*se.outcome, ymax = beta.outcome + 1.96*se.outcome), height = 0, alpha = 0.3) +
    geom_abline(data = dat_res |> dplyr::filter(nsnp > 1), aes(slope = b, intercept = 0, colour = IV_set)) +
    geom_point(size = 2) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5)) +
    guides(colour = guide_legend(nrow = 2)) +
    labs(x = expression(beta["exposure"]), y = expression(beta["outcome"]), title = paste("Exposure:", exposure, "Outcome:", outcome))
    
  ## Forest plot
  p_forest <- dat_res |> ggplot(aes(x = b, y = IV_set, colour = IV_set)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(aes(size = nsnp)) +
    geom_errorbarh(aes(xmin = b-1.96*se, xmax = b+1.96*se, group = cis_trans), height = 0.4) +
    scale_colour_manual(values = cols_IVsets) +
    theme_minimal() + guides(colour = "none") +
    theme(plot.title = element_text(hjust = 0.5), 
          axis.text.x = element_text(size = 7),
          legend.position = "right") +
    labs(x = "IVW", y = "", size = "n(IVs)")
  
  cowplot::ggdraw() + cowplot::draw_plot(p_slope,0,0,0.5,1) + cowplot::draw_plot(p_forest,0.5,0.3,0.5,0.6)
}

plot_results(exposure = "PRSS53", outcome = "BMI")
plot_results(exposure = "GMPR", outcome = "RBCcount")
plot_results(exposure = "VNN1", outcome = "VitD")
plot_results(exposure = "PDCD6", outcome = "VitD")
```

**Comparison of IVW effect estimates across instrument sets**

Given the relationships shown above, particularly that reducing the stringency of significance thresholding for rare IVs does not reduce instrument specificity (pleiotropy profiles remain consistant), I have compared IVW effect estimates for protein --> outcome relationships instrumented using at least three common (at P<5e-8) and at least three rare (at P<1e-4) cis or trans instruments.

```{r}
#| label: compare estimates
#| cache: TRUE
#| message: FALSE

# Functions
# Generate a table of IVW estimates for a pair of instrument sets
compare_estimates <- function(results_MR, IVset1, IVset2, n_snps=3, cis_trans="all"){
  # Pull out IVW estimates and SE for instrument sets
  if(cis_trans == "all"){
    df <- lapply(results_MR, function(x){
      x |> dplyr::filter(method == "Inverse variance weighted", 
                         cis_trans == "cis_trans",
                         IV_set %in% c(IVset1,IVset2))
    })
  }
  if(cis_trans == "cis"){
    df <- lapply(results_MR, function(x){
      x |> dplyr::filter(method == "Inverse variance weighted", 
                         cis_trans == "cis",
                         IV_set %in% c(IVset1,IVset2))
    })
  }
  if(cis_trans == "trans"){
    df <- lapply(results_MR, function(x){
      x |> dplyr::filter(method == "Inverse variance weighted", 
                         cis_trans == "trans",
                         IV_set %in% c(IVset1,IVset2))
    })
  }
  df_out_b <- do.call("rbind",df) |> 
    dplyr::filter(nsnp >= n_snps) |>
    tidyr::pivot_wider(names_from = "IV_set", values_from = c("b"), id_cols = "exposure") |>
    na.exclude() |>
    dplyr::relocate(exposure,IVset1,IVset2)
  
  df_out_se <- do.call("rbind",df) |> 
    dplyr::filter(nsnp >= n_snps) |>
    tidyr::pivot_wider(names_from = "IV_set", values_from = "se", id_cols = "exposure") |>
    na.exclude() |>
    dplyr::relocate(exposure,IVset1,IVset2)
  
  df_out_p <- do.call("rbind",df) |> 
    dplyr::filter(nsnp >= n_snps) |>
    tidyr::pivot_wider(names_from = "IV_set", values_from = "pval", id_cols = "exposure") |>
    na.exclude() |>
    dplyr::relocate(exposure,IVset1,IVset2)
  colnames(df_out_p)[2:3] <- paste0(colnames(df_out_p)[2:3],".pval")
  
  ## Caluclate effect heterogeneity
  if (!(identical(df_out_b$exposure, df_out_se$exposure) &&
        identical(colnames(df_out_b), colnames(df_out_se)))) {
    message("Error - mismatch of beta and SE dataframes")
    return(NULL)
  }
  
  beta_vecs <- lapply(split(df_out_b[,c(2,3)], 1:nrow(df_out_b)), as.numeric)
  se_vecs <- lapply(split(df_out_se[,c(2,3)], 1:nrow(df_out_se)), as.numeric)
  
  het_res <- do.call("rbind", mapply(effect_heterogeneity, beta_vec = beta_vecs, se_vec = se_vecs, SIMPLIFY = F)) |>
    as.data.frame() |>
    dplyr::mutate(exposure = df_out_b$exposure, .before = "Q") |>
    dplyr::mutate(Qpval_bon = p.adjust(Qpval, method = "bonferroni"),
                  Qpval_fdr = p.adjust(Qpval, method = "fdr"))
  
  df_out <- dplyr::left_join(df_out_b, df_out_se, by = "exposure", suffix = c(".beta",".se")) |>
    dplyr::left_join(df_out_p, by = "exposure") |>
    dplyr::left_join(het_res, by = "exposure")
  
  return(df_out)
}

# Generate a table of cis vs trans IVW estimates for given instrument set
compare_cistrans <- function(results_MR, IVset, n_snps=3){
  # Pull out IVW estimates and SE for instrument sets
  df <- lapply(results_MR, function(x){
    x |> 
      dplyr::filter(method == "Inverse variance weighted",
                    IV_set %in% IVset,
                    nsnp >= n_snps,
                    cis_trans %in% c("cis","trans")) |>
      dplyr::mutate(IV_set = paste(IV_set, cis_trans, sep = "_")) |>
      dplyr::arrange(cis_trans)})
  df <- df[lapply(df, nrow) == 2]
  
  df_out_b <- do.call("rbind",df) |> 
    tidyr::pivot_wider(names_from = "IV_set", values_from = "b", id_cols = "exposure") |>
    na.exclude()
  
  df_out_se <- do.call("rbind",df) |> 
    tidyr::pivot_wider(names_from = "IV_set", values_from = "se", id_cols = "exposure") |>
    na.exclude()
  
  df_out_p <- do.call("rbind",df) |> 
    tidyr::pivot_wider(names_from = "IV_set", values_from = "pval", id_cols = "exposure") |>
    na.exclude()
  colnames(df_out_p)[2:3] <- paste0(colnames(df_out_p)[2:3],".pval")
  
  ## Caluclate effect heterogeneity
  if (!(identical(df_out_b$exposure, df_out_se$exposure) &&
        identical(colnames(df_out_b), colnames(df_out_se)))) {
    message("Error - mismatch of beta and SE dataframes")
    return(NULL)
  }
  
  beta_vecs <- lapply(split(df_out_b[,c(2,3)], 1:nrow(df_out_b)), as.numeric)
  se_vecs <- lapply(split(df_out_se[,c(2,3)], 1:nrow(df_out_se)), as.numeric)
  
  het_res <- do.call("rbind", mapply(effect_heterogeneity, beta_vec = beta_vecs, se_vec = se_vecs, SIMPLIFY = F)) |>
    as.data.frame() |>
    dplyr::mutate(exposure = df_out_b$exposure, .before = "Q") |>
    dplyr::mutate(Qpval_bon = p.adjust(Qpval, method = "bonferroni"),
                  Qpval_fdr = p.adjust(Qpval, method = "fdr"))
  
  df_out <- dplyr::left_join(df_out_b, df_out_se, by = "exposure", suffix = c(".beta",".se")) |>
    dplyr::left_join(df_out_p, by = "exposure") |>
    dplyr::left_join(het_res, by = "exposure")
  
  return(df_out)
}

# Proportion of concordant/discordant MR estimates based on heterogenity estimates and direction of effect
prop_concordant <- function(estimates){
  
  estimate_cols <- grep("\\.beta",colnames(estimates), value = T)
  pval_cols <- grep("\\.pval",colnames(estimates), value = T)
  sets <- sub(".beta", "", estimate_cols) # IV sets being compared
  
  df_counts <- data.frame(n_instrumented_genes = nrow(estimates),
                       n_nohet = sum(estimates$Qpval > 0.05),
                       n_het_concordant = sum(estimates$Qpval <= 0.05 & sign(estimates[estimate_cols[1]]) == sign(estimates[estimate_cols[2]])),
                       n_het_discordant = sum(estimates$Qpval <= 0.05 & sign(estimates[estimate_cols[1]]) != sign(estimates[estimate_cols[2]])))
  
  df_props <- df_counts[,2:4]/df_counts[,1]
  colnames(df_props) <- sub("n_","prop_",colnames(df_props))
  
  df_out <- cbind(df_counts, df_props)
  
  # Discordant results
  estimates_het <- estimates |> dplyr::filter(Qpval <= 0.05)
  
  df_counts_het <- data.frame(n_het_genes = nrow(estimates_het),
                              sig1 = sum(estimates_het[pval_cols[1]] < estimates_het[pval_cols[2]] & estimates_het[pval_cols[2]] > 0.05),
                              sig2 = sum(estimates_het[pval_cols[2]] < estimates_het[pval_cols[1]] & estimates_het[pval_cols[1]] > 0.05),
                              n_both_sig_concordant = sum(estimates_het[pval_cols[1]] < 0.05 & estimates_het[pval_cols[2]] < 0.05 & 
                                                    sign(estimates_het[estimate_cols[1]]) == sign(estimates_het[estimate_cols[2]])),
                              n_both_sig_discordant = sum(estimates_het[pval_cols[1]] < 0.05 & estimates_het[pval_cols[2]] < 0.05 & 
                                                    sign(estimates_het[estimate_cols[1]]) != sign(estimates_het[estimate_cols[2]])))
  
  colnames(df_counts_het)[2] <- paste0("n_",sets[1],"_sig")
  colnames(df_counts_het)[3] <- paste0("n_",sets[2],"_sig")
  
  df_het_props <- df_counts_het[,2:5]/df_counts_het[,1]
  colnames(df_het_props) <- sub("n_","prop_",colnames(df_het_props))
  
  df_out_het <- cbind(df_counts_het, df_het_props)
  
  return(list(all = df_out, sigQ = df_out_het))
}

plot_concordance <- function(concordance, title){
  
  plot_dat_1 <- concordance$all |> 
    tibble::rownames_to_column("outcome") |> 
    dplyr::mutate(n_het = n_instrumented_genes - n_nohet) |>
    dplyr::select(c(1,3,9)) |> tidyr::pivot_longer(cols = c(2:3))
  
  p1 <- plot_dat_1 |> ggplot(aes(x = outcome, y = value, fill = name)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = "Outcome", 
         y = " Proportion of MR results \n(protein --> outcome)") +
    scale_fill_manual(name = "",
      values = c("tomato","grey90"), 
      labels = c("Qpval<0.05","Qpval>0.05")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 10),
          legend.key.size = unit(0.6, "cm"),
          legend.position = "bottom") +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
  
  plot_dat_2 <- concordance$Qsig |> 
    tibble::rownames_to_column("outcome") |> 
    dplyr::select(c(1,3:6)) |> tidyr::pivot_longer(cols = c(2:5))
  
  sets = unique(plot_dat_2$name)[1:2]
  sets = sub("_sig$","",sub("^n_","",sets))
  
  if(all(grepl("cis|trans", sets))){
    fill_names = c(paste(IVsets[sub("_(cis|trans)$","",sets)],c("(cis)","(trans)")), "Sig. conc.", "Sig. dis.")
    names(fill_names) <- unique(plot_dat_2$name)
    fill_col = c("grey", "grey30","tomato","steelblue1")
  }else{
    fill_names = c(IVsets[sets], "Sig. conc.", "Sig. dis.")
    names(fill_names) <- unique(plot_dat_2$name)
    fill_col = c("grey", "grey30",as.character(cols_IVsets[fill_names[1:2]]))
  }
  
  p2 <- plot_dat_2 |> ggplot(aes(x = outcome, y = value, fill = name)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = "Outcome", 
         y = " Proportion of heterogenous MR results \n(protein --> outcome)") +
    scale_fill_manual(name = "",
      values = fill_col,
      labels = fill_names) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 10),
          legend.position = "bottom",
          legend.key.size = unit(0.6, "cm")) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE))
  
  gridExtra::grid.arrange(p1,p2,nrow = 1, top = title)
  
}

plot_estimates <- function(estimates, name){
  
  estimates <- as.data.frame(estimates)
  col_names <- names(estimates)
  
  if(any(grepl("cis|trans",col_names))){
    lab_x <- "cis-IVs"
    lab_y <- "trans-IVs"
  } else {
    lab_x <- IVsets[sub("\\.beta","",col_names[2])]
    lab_y <- IVsets[sub("\\.beta","",col_names[3])]
  }
  
  CI_x <- data.frame(LCI = estimates[,2] - 1.96*estimates[,4], UCI = estimates[,2] + 1.96*estimates[,4])
  CI_y <- data.frame(LCI = estimates[,3] - 1.96*estimates[,5], UCI = estimates[,3] + 1.96*estimates[,5])
  
  mod_lm <- lm(estimates[,3] ~ estimates[,2])
  slope <- round(coefficients(mod_lm)[2],2)
  
  flag_point <- estimates$Qpval <= 0.05
  
  range_min <- min(min(estimates[,2]), min(estimates[,3])) #min(na.exclude(errorbar_xmin, errorbar_ymin)))
  range_max <- max(max(estimates[,2]), max(estimates[,3])) #min(na.exclude(errorbar_xmax, errorbar_ymax)))
  
  estimates_plot <- estimates |>
    dplyr::mutate(exposure = factor(exposure, levels = estimates$exposure[order(estimates$Qpval, decreasing = TRUE)]),
                  errorbar_xmin = ifelse(flag_point, CI_x$LCI, NA),
                  errorbar_xmax = ifelse(flag_point, CI_x$UCI, NA),
                  errorbar_ymin = ifelse(flag_point, CI_y$LCI, NA),
                  errorbar_ymax = ifelse(flag_point, CI_y$UCI, NA)) |>
    dplyr::arrange(exposure)
  
  estimates_plot |>
    ggplot(aes_string(x = col_names[2], y = col_names[3])) +
    geom_abline(slope = 1, intercept = 0, colour = "steelblue", lty = 2, lwd = 0.5) +
    geom_point(aes(colour = estimates_plot$Qpval <= 0.05), size = 2, shape = 16) +  
    geom_errorbar(xmin = estimates_plot$errorbar_xmin,
                  xmax = estimates_plot$errorbar_xmax, colour = "tomato") +
    geom_errorbar(ymin = estimates_plot$errorbar_ymin,
                  ymax = estimates_plot$errorbar_ymax, colour = "tomato") +
    geom_smooth(method = "lm", se = FALSE, colour = "tomato", lwd = 0.5) +
    annotate(geom = "text", x = range_min/2, y = range_max, label = paste0("Slope=",slope), size = 3, colour = "tomato") +
    scale_colour_manual(values = c("black","tomato")) +
    scale_x_continuous(limits = c(range_min, range_max)) +
    scale_y_continuous(limits = c(range_min, range_max)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 10), legend.position = "none") +
    labs(x=lab_x, y=lab_y, title = name)
}

# Upset plot comparing signifiant causal estimates across variant sets
# Restrtcited to those proteins that could be instrumented with >= 3 variants across the comparison sets

plot_upset <- function(IVsets_list, cis_trans = "all", pval = 0.01){
  
  if(cis_trans == "all"){inlist = summary_sig}
  if(cis_trans == "cis"){inlist = summary_sig_cis}
  if(cis_trans == "trans"){inlist = summary_sig_trans}
  
  set_sig <- lapply(inlist, function(x){
    dat <- x$min_p[,IVsets_list]
    # Extract only those tests that could be conducted in all IV sets
    tests <- rownames(dat) # Exposure_outcome
    shared_tests <- tests[apply(dat, MARGIN = 1, function(x){sum(!(is.na(x)))}) == 2]
    dat <- dat[shared_tests,]
    dat_sig <- dat <= pval
    
    tests <- rownames(dat_sig)
    list_sig <- apply(dat_sig, MARGIN = 2, FUN = function(x){tests[x]}, simplify = F) # Names of significant exposures
    names(list_sig) <- IVsets[names(list_sig)]
    list_sig <- list_sig[lapply(list_sig, length) > 0] # Remove empty exposure sets
    return(list_sig)
  })
  
  # Upset plots
  plots <- mapply(FUN = function(list_sig, outcome){
    if(length(list_sig) == 1){return(NULL)}
    upset(fromList(list_sig), order.by = "freq", text.scale = 1.5, mainbar.y.label = paste0("Protein:",outcome,"\n (IVW p<", pval, ")"))
  }, list_sig = set_sig, outcome = names(set_sig), SIMPLIFY = F)
  
  plots <- plots[!sapply(plots, is.null)]
  
  # for (v in names(plots)) {
  #   print(plots[[v]])
  #   #grid.text(v, x = 0.65, y=0.97, gp = gpar(fontsize = 20))
  #   grid.edit('arrange', name = v)
  #   vp <- grid.grab()
  #   plots[[v]] <- vp
  # }
  return(plots)
}
```

1a) Common vs rare variant cis-IVs (ExWAS)\
1b) Common vs rare variant trans-IVs (ExWAS)

2a) Common vs ultra-rare variant cis-IVs (ExWAS)\
2b) Common vs ultra-rare variant trans-IVs (ExWAS)

```{r}
#| label: fig-compIVs-commonvrare
#| cache: TRUE
#| message: FALSE
#| fig.width: 14
#| fig.height: 6

comp_IVWs_1a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_1a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_1a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_1a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_1a, name = names(comp_IVWs_1a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and rare cis-pQTL instrumentation strategies")

comp_IVWs_1b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_1b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_1b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_1b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_1b, name = names(comp_IVWs_1b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and rare trans-pQTL instrumentation strategies")

comp_IVWs_2a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_2a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_2a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_2a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_2a, name = names(comp_IVWs_2a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and ultra-rare cis-pQTL instrumentation strategies")

comp_IVWs_2b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "dhindsa_exwas_variant_common",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_2b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_2b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_2b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_2b, name = names(comp_IVWs_2b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from WES common and ultra-rare trans-pQTL instrumentation strategies")
```

```{r}
#| label: fig-compIVs-commonvrare-concordance
#| fig.width: 8
#| fig.height: 5

plot_concordance(concordance = comp_IVWs_1a_conc, title = "IVW estimate concordance: WES common vs rare cis-IVs")
plot_concordance(concordance = comp_IVWs_1b_conc, title = "IVW estimate concordance: WES common vs rare trans-IVs") 

plot_concordance(concordance = comp_IVWs_2a_conc, title = "IVW estimate concordance: WES common vs ultrarare cis-IVs")
plot_concordance(concordance = comp_IVWs_2b_conc, title = "IVW estimate concordance: WES common vs ultrarare trans-IVs") 

```

3a) Common (GWAS) vs rare variant cis-IVs (ExWAS)\
3b) Common (GWAS) vs rare variant trans-IVs (ExWAS)

4a) Common (GWAS) vs ultra-rare variant cis-IVs (ExWAS)\
4b) Common (GWAS) vs ultra-rare variant trans-IVs (ExWAS)

```{r}
#| label: fig-compIVs-commonGWASvrare
#| cache: TRUE
#| message: FALSE
#| fig.width: 14
#| fig.height: 6

comp_IVWs_3a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_3a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_3a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_3a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_3a, name = names(comp_IVWs_3a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and WES rare cis-pQTL instrumentation strategies")

comp_IVWs_3b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_rare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_3b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_3b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_3b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_3b, name = names(comp_IVWs_3b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS  common and WES rare trans-pQTL instrumentation strategies")

comp_IVWs_4a <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "cis")

# Proportion concordant
comp_IVWs_4a_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_4a, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_4a, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_4a, name = names(comp_IVWs_4a), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and WES ultra-rare cis-pQTL instrumentation strategies")

comp_IVWs_4b <- lapply(res_MR_relaxed, FUN = compare_estimates, 
                       IVset1 = "sun_gwas_variant_common_clumped",
                       IVset2 = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3,
                       cis_trans = "trans")

# Proportion concordant
comp_IVWs_4b_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_4b, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_4b, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_4b, name = names(comp_IVWs_4b), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and WES ultra-rare trans-pQTL instrumentation strategies")

# Extract list of genes with significant heterogenity between cis common and rare/ultrarare variant estimates
het_genes <- lapply(
  Map(c, lapply(comp_IVWs_1a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_2a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_3a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_4a, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)})), function(x){unique(x)})

het_genes <- mapply(FUN = function(x,y){paste(x,y,sep = "_")}, x = het_genes, y = names(het_genes), SIMPLIFY = F)
```

```{r}
#| label: fig-compIVs-GWAScommonvrare-concordance
#| fig.width: 8
#| fig.height: 5

plot_concordance(concordance = comp_IVWs_3a_conc, title = "IVW estimate concordance: GWAS common vs rare cis-IVs")
plot_concordance(concordance = comp_IVWs_3b_conc, title = "IVW estimate concordance: GWAS common vs rare trans-IVs") 

plot_concordance(concordance = comp_IVWs_4a_conc, title = "IVW estimate concordance: GWAS common vs ultrarare cis-IVs")
plot_concordance(concordance = comp_IVWs_4b_conc, title = "IVW estimate concordance: GWAS common vs ultrarare trans-IVs") 

```

**Overlap between protein:outcome pairs with significant causal effectes estimated across cis-IV sets**

```{r}
#| label: upset-compIVs-common
#| fig.width: 8
#| fig.height: 3
#| eval: FALSE

plot_upset(IVsets_list = c("sun_gwas_variant_common_clumped","dhindsa_exwas_variant_rare","dhindsa_exwas_variant_ultrarare"), cis_trans = "cis", pval = 0.05)

```

7)  Common cis-IVs vs trans-IVs (GWAS)\
8)  Common cis-IVs vs trans-IVs (ExWAS)\
9)  Rare cis-IVs vs trans-IVs (ExWAS)\
10) Ultra-rare cis-IVs vs trans-IVs\

```{r}
#| label: fig-compIVs-cisvtrans
#| cache: TRUE
#| message: FALSE
#| fig.width: 14
#| fig.height: 6

comp_IVWs_5 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                      IVset = "sun_gwas_variant_common_clumped",
                      n_snps = 3)

# Proportion concordant
comp_IVWs_5_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_5, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_5, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_5, name = names(comp_IVWs_5), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from GWAS common and cis- and trans-pQTL instrumentation strategies")

comp_IVWs_6 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                      IVset = "dhindsa_exwas_variant_common",
                      n_snps = 3)

# Proportion concordant
comp_IVWs_6_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_6, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_6, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_6, name = names(comp_IVWs_6), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from ExWAS common and cis- and trans-pQTL instrumentation strategies")

comp_IVWs_7 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                      IVset = "dhindsa_exwas_variant_rare_pruned",
                      n_snps = 3)

# Proportion concordant
comp_IVWs_7_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_7, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_7, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_7, name = names(comp_IVWs_7), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from ExWAS rare and cis- and trans-pQTL instrumentation strategies")

comp_IVWs_8 <- lapply(res_MR_relaxed, FUN = compare_cistrans, 
                       IVset = "dhindsa_exwas_variant_ultrarare_pruned",
                       n_snps = 3)

# Proportion concordant
comp_IVWs_8_conc <- list("all" = do.call("rbind", lapply(lapply(comp_IVWs_8, FUN = prop_concordant), function(x){x$all})),
                          "Qsig" = do.call("rbind", lapply(lapply(comp_IVWs_8, FUN = prop_concordant), function(x){x$sigQ})))

gridExtra::grid.arrange(grobs = mapply(plot_estimates, estimates = comp_IVWs_8, name = names(comp_IVWs_8), SIMPLIFY = FALSE), nrow = 2,
                        top = "Comparison of IVW estimates from ExWAS ultra-rare and cis- and trans-pQTL instrumentation strategies")

# Extract list of genes with significant heterogenity between cis and trans estimates
het_genes_cistrans <- lapply(
  Map(c, lapply(comp_IVWs_5, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_6, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_7, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_8, function(x){x |> dplyr::filter(Qpval <= 0.01) |> dplyr::pull(exposure)})), function(x){unique(x)})

het_genes_cistrans <- mapply(FUN = function(x,y){paste(x,y,sep = "_")}, x = het_genes_cistrans , y = names(het_genes_cistrans), SIMPLIFY = F)

```

```{r}
#| label: fig-compIVs-cisvtans-concordance
#| fig.width: 9
#| fig.height: 5

plot_concordance(concordance = comp_IVWs_5_conc, title = "IVW estimate concordance: GWAS common cis vs trans-IVs")
plot_concordance(concordance = comp_IVWs_6_conc, title = "IVW estimate concordance: WES common cis vs trans-IVs") 
plot_concordance(concordance = comp_IVWs_7_conc, title = "IVW estimate concordance: WES rare cis vs trans-IVs")
plot_concordance(concordance = comp_IVWs_8_conc, title = "IVW estimate concordance: WES common ultrarare cis vs trans-IVs") 
```

### Heterogenity in causal effects

Restricting to those protein exposures that show signficiant causal effects on a given outcome (at p \<= 0.05 with \>= 3 IVs), and significant heterogenity between rare and common cis variant IV sets (at Qstat P(FDR) \< 0.05):

```{r}
#| label: heterogenous effects
#| cache: TRUE

# Retain exposure-outcome pairs with significant IVW estimate for at least one cis instrumentation strategy (with n instruments >= n_snps)
res_MR_top <- lapply(res_MR_relaxed, filter_results, p_thresh = 0.05, n_snps = 3, cistrans = "cis")
res_MR_top_cistrans <- lapply(res_MR_relaxed, filter_results, p_thresh = 0.05, n_snps = 3, cistrans = c("cis","trans"))

# Filter to those with evidence of estimate heterogenity between variant sets
res_MR_top_het <- mapply(function(x,y){x[names(x) %in% y]}, x = res_MR_top, y = het_genes, SIMPLIFY = F)

# Filter to those with evidence of estimate heterogeneity between cis and trans IVs
res_MR_top_cistrans_het <- mapply(function(x,y){x[names(x) %in% y]}, x = res_MR_top, y = het_genes_cistrans, SIMPLIFY = F)

```

**Heterogeneity in rare v common cis IV causal estimates**

```{r}
#| label: fig-forestplots1
#| fig.width: 8.5
#| fig.height: 3.5
#| results: 'hide'

# Plot estimates
res_IVW_cis <- lapply(res_MR_top_het, function(x){
  df <- do.call("rbind", x) |> 
    filter(method == "Inverse variance weighted",
           grepl("variant",IV_set),
           cis_trans == "cis") |>
    dplyr::mutate(instrument_class = factor(variant_IVsets[IV_set], levels = rev(variant_IVsets)))
  return(df)})

plot_forest <- function(results_IVW){
  pd <- ggplot2::position_dodge2(width = 0.8, preserve = "single")
  exposure <- unique(sub(".*_", "", results_IVW$pair))
  
  n_row <- ceiling(length(unique(results_IVW$exposure))/4)
  
  p <- ggplot(results_IVW, aes(x = b, y = instrument_class, colour = instrument_class)) +
    facet_wrap(.~pair, ncol = 5, scale = "free_x") +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(aes(size = nsnp), position = pd) +
    geom_errorbarh(aes(xmin = b-1.96*se, xmax = b+1.96*se, group = cis_trans), position = pd, height = 0.4) +
    scale_colour_manual(values = cols_IVsets) +
    theme_bw() + guides(colour = "none") +
    theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(size = 7)) +
    labs(x = "IVW", y = "", size = "n (instruments)", title = paste("Outcome:", exposure))
  
  return(p)
}

lapply(res_IVW_cis, plot_forest)[c(1,4,6,7,8,9,10)]
```

```{r}
#| label: fig-forestplots2
#| fig.width: 8.5
#| fig.height: 2
#| results: 'hide'

lapply(res_IVW_cis, plot_forest)[c(3,5,11)]

```

```{r}
#| label: fig-forestplots3
#| fig.width: 8.5
#| fig.height: 5
#| results: 'hide'

lapply(res_IVW_cis, plot_forest)[c(2)]

```

**Heterogeneity in cis v trans IV causal estimates**

```{r}
#| label: fig-forestplots-cistrans-1
#| fig.width: 8.5
#| fig.height: 4
#| results: 'hide'

# Plot estimates
res_IVW_cistrans <- lapply(res_MR_top_cistrans_het, function(x){
  df <- do.call("rbind", x) |> 
    filter(method == "Inverse variance weighted",
           grepl("variant",IV_set),
           cis_trans %in% c("cis","trans")) |>
    dplyr::mutate(instrument_class = factor(variant_IVsets[IV_set], levels = rev(variant_IVsets)))
  return(df)})

plot_forest_cistrans <- function(results_IVW){
  pd <- ggplot2::position_dodge2(width = 0.8, preserve = "single")
  exposure <- unique(sub(".*_", "", results_IVW$pair))
  
  n_row <- ceiling(length(unique(results_IVW$exposure))/4)
  
  p <- ggplot(results_IVW, aes(x = b, y = instrument_class, colour = instrument_class)) +
    facet_wrap(.~pair, ncol = 5, scale = "free_x") +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(aes(pch = cis_trans, size = nsnp), position = pd) +
    geom_errorbarh(aes(xmin = b-1.96*se, xmax = b+1.96*se, group = cis_trans), position = pd) +
    scale_colour_manual(values = cols_IVsets) +
    scale_shape_manual(name = "", values = c(16,1)) +
    theme_bw() + guides(colour = "none") +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(x = "IVW", y = "", size = "n (instruments)", title = paste("Outcome:", exposure))
  
  return(p)
}

lapply(res_IVW_cistrans, plot_forest_cistrans)[c(1,2,3,4,7,9)]
```

```{r}
#| label: fig-forestplots-cistrans-2
#| fig.width: 8.5
#| fig.height: 2.5
#| results: 'hide'

lapply(res_IVW_cistrans, plot_forest_cistrans)[c(6,8,10,11)]

```

```{r}
#| label: fig-forestplots-cistrans-3
#| fig.width: 8.5
#| fig.height: 5.5
#| results: 'hide'

lapply(res_IVW_cistrans, plot_forest_cistrans)[c(5)]

```

**Concordance vs discordance in cis and trans effects**

How often do causal estimates for a protein:outcome effect derived from cis- and trans-pQTLs agree (in terms of significance and direction of effect)? 

### Consistency in causal effects

Restricting to those protein exposures that show signficiant causal effects on a given outcome (at p \<= 0.05 with \>= 3 IVs) supported by rare and common cis variant IV sets (Qsat P(FDR) \< 0.05):

```{r}
#| label: consistent effects
#| fig.width: 10.5
#| fig.height: 5

# Extract list of genes with consistent (significant) effects between cis common and rare/ultrarare variant estimates
consistant_genes <- lapply(
  Map(c, 
      lapply(comp_IVWs_1a, function(x){x |> dplyr::filter(Qpval > 0.1 & x[,6] < 0.05 & x[,7] < 0.05 & sign(x[,2]) == sign(x[,3])) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_2a, function(x){x |> dplyr::filter(Qpval > 0.1 & x[,6] < 0.05 & x[,7] < 0.05 & sign(x[,2]) == sign(x[,3])) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_3a, function(x){x |> dplyr::filter(Qpval > 0.1 & x[,6] < 0.05 & x[,7] < 0.05 & sign(x[,2]) == sign(x[,3])) |> dplyr::pull(exposure)}),
      lapply(comp_IVWs_4a, function(x){x |> dplyr::filter(Qpval > 0.1 & x[,6] < 0.05 & x[,7] < 0.05 & sign(x[,2]) == sign(x[,3])) |> dplyr::pull(exposure)})), function(x){unique(x)})

consistant_genes <- mapply(FUN = function(x,y){paste(x,y,sep = "_")}, x = consistant_genes, y = names(consistant_genes), SIMPLIFY = F)

# Filter to significant MR results supported across IV sets
res_MR_top_consistant <- mapply(function(x,y){x[names(x) %in% y]}, x = res_MR_top, y = consistant_genes, SIMPLIFY = F)

res_IVW_consistant <- lapply(res_MR_top_consistant, function(x){
  df <- do.call("rbind", x) |> 
    filter(method == "Inverse variance weighted",
           grepl("variant",IV_set),
           cis_trans == "cis") |>
    dplyr::mutate(instrument_class = factor(variant_IVsets[IV_set], levels = rev(variant_IVsets)))
  return(df)})
```

```{r}
#| label: fig-forestplots-consistant-1
#| fig.width: 8.5
#| fig.height: 2
#| results: 'hide'

lapply(res_IVW_consistant, plot_forest)[c(1,2,5,6,7,8,9,10,11)]

```

```{r}
#| label: fig-forestplots-consistant-2
#| fig.width: 8.5
#| fig.height: 3.5
#| results: 'hide'

lapply(res_IVW_consistant, plot_forest)[c(3,4)]

```

## Allelic series

Assessing for monotonically increasing causal effect estimates (Wald Ratios) across rare-variant series of alleles:\
- Rare variants (\<1% MAF): benign missense variants (BMV) - damaging missense variants (DMVs) - protein truncating variants (PTVs)\

Using the [COAST-SS](https://github.com/insitro/AllelicSeries) approach (developed for allelic series association tests), which has been developed to handle GWAS summary statistics, but here we are passing Wald Ratios (variant outcome effect/ variant protein effect)

What is needed here?: \
- Summary statistics for all rare variants (capped at p<0.01 by the AzPheWAS portal for pQTLs) falling in the coding region of genes encoding the O'link plasma proteome assayed proteins\
- The corresponding variant associations for outcomes of interest (with which to calculate Wald ratios)\ 
- Variant predicted functional effects (with which to categorise SNPs)

**Magnitude of variant class effects**

Do cis-variants predicted to be more deleterious in nature (loss of function) actually have larger effects on protein expression?

```{r}
#| label: vep-classes
#| fig.width: 7
#| fig.height: 3

# Read in variant effect estimates, wald ratios and VEP annotations
load(file.path(res_dir,"results_molecular_waldratios_ciscodingvariants_VEPannotated.rda"))

wald_ratios <- lapply(wald_ratios, function(x){
  x$annots_relaxed <- as.factor(x$annots_relaxed)
  x})

vep_class <- c("1" = "syn./noncoding", 
               "2" = "benign missense",
               "3" = "del. missense",
               "4" = "putative LoF")

cols_vep <- c("1" = "steelblue", 
              "2" = "olivedrab3",
              "3" = "orange",
              "4" = "tomato")

# Plot variant consequnce call on cis pQTL effect
wald_ratios$LDL |> 
  dplyr::filter(!is.na(annots_relaxed)) |> 
  ggplot(aes(x = beta.exposure, 
             y = annots_relaxed)) +
  ggridges::geom_density_ridges(aes(fill = annots_relaxed), trim = TRUE) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  scale_y_discrete(labels = vep_class) +
  scale_fill_manual(values = cols_vep) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "cis-pQTL effect (beta)", y = "Variant annotation")

# Plot wald ratios split by variant consequence
plot_wald_vep <- function(gene, outcome){
  dat <- wald_ratios[[outcome]] |> 
    dplyr::filter(exposure == gene) |>
    dplyr::arrange(annots_relaxed, wald_ratio_b) |>
    dplyr::mutate(SNP = factor(SNP, levels = SNP))
  
  res <- coastss_annots_relaxed[[outcome]] |>
    dplyr::filter(exposure == gene)
  
  p_theme <- theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          legend.position = "bottom",
          legend.key.height = unit(0.2, 'cm'))
  
  # Exposure
  p_exp <- ggplot(dat, aes(x = beta.exposure, y = SNP, colour = annots_relaxed)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point() +
    scale_colour_manual(values = cols_vep, labels = vep_class, name = "") +
    geom_errorbarh(aes(xmin = beta.exposure-1.96*se.exposure, xmax = beta.exposure+1.96*se.exposure)) +
    labs(x = "Bx", y = "", 
         title = paste("Exposure:", gene),
         subtitle = "") +
    guides(colour = guide_legend(nrow=2, byrow=TRUE)) +
    p_theme
  
  # Outcome
  p_out <- ggplot(dat, aes(x = beta.outcome, y = SNP, colour = annots_relaxed)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point() +
    scale_colour_manual(values = cols_vep, labels = vep_class, name = "") +
    geom_errorbarh(aes(xmin = beta.outcome-1.96*se.outcome, xmax = beta.outcome+1.96*se.outcome)) +
    labs(x = "By", y = "", 
         title = paste("Outcome:", outcome),
         subtitle = "") +
    guides(colour = guide_legend(nrow=2, byrow=TRUE)) +
    p_theme
  
  # Wald Ratio
  p_wald <- ggplot(dat, aes(x = wald_ratio_b, y = SNP, colour = annots_relaxed)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point() +
    scale_colour_manual(values = cols_vep, labels = vep_class, name = "") +
    geom_errorbarh(aes(xmin = wald_ratio_b-1.96*wald_ratio_se, xmax = wald_ratio_b+1.96*wald_ratio_se)) +
    labs(x = "Wald Ratio (By/Bx)", y = "", 
         title = paste("Exposure:", gene, "Outcome:", outcome),
         subtitle = paste("COAST-SS p-value:", signif(res$alleleic_skat_p, 2))) +
    guides(colour = guide_legend(nrow=2, byrow=TRUE)) +
    p_theme
  
  return(gridExtra::grid.arrange(grobs = list(p_wald, p_exp, p_out), nrow = 1))
}

```

### COAST-SS analysis

```{r}
#| label: COAST-SS 
#| fig.width: 12
#| fig.height: 20

# Load results from COAST-SS allelic series anlysis 
# 1 = synonymous/noncoding, 2 = benign missense, 3 = deleterious missense, 4 = putative loss of funcion

load(file.path(res_dir,"results_molecular_COASTSS_relaxed.rda")) # coastss_annots_relaxed

## Filter out exposure genes for which there were insufficient variants for alleleic series analysis and calculate adjusted p-vals
coastss_annots_relaxed <- lapply(coastss_annots_relaxed, function(x){
  x |> 
    dplyr::filter(!is.na(alleleic_skat_p)) |>
    dplyr::mutate(adj_p = p.adjust(alleleic_skat_p, method = "BH"))
})

outcome_names <- c("LDL direct" = "LDL",
                    "Body mass index (BMI)" = "BMI",
                    "Vitamin D" = "Vitamin D",
                    "Triglycerides" = "Triglycerides",
                    "Glycated haemoglobin (HbA1c)" = "HbA1c",
                    "Mean platelet (thrombocyte) volume" = "PlatVol",
                    "IGF-1" = "IGF-1",
                    "Waist-to-hip ratio adjusted for body mass index (WHRadjBMI)" = "WHRadjBMI",
                    "Red blood cell (erythrocyte) count" = "RBCcount",
                    "Mean corpuscular volume" = "CorpVol",
                    "Systolic blood pressure, automated reading" = "SBP")
                    

do.call("rbind", coastss_annots_relaxed) |>
  dplyr::mutate(label = ifelse(adj_p < 0.05, exposure, NA),
                outcome = outcome_names[outcome]) |>
  ggplot(aes(x = exposure, y = -log10(alleleic_skat_p))) +
  facet_grid(outcome ~ ., scale = "free_y") +
  geom_point(aes(colour = adj_p < 0.05)) +
  scale_colour_manual(name = "P(FDR) < 0.05", values = c("grey","tomato")) +
  ggrepel::geom_text_repel(aes(label = label), size = 4, max.overlaps = Inf, force = 10) +
  theme_minimal() +
  theme(strip.text = element_text(size = 10),
        panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.title = element_text(size = 14),
        legend.position = "bottom") +
  labs(x = "Protein exposure", y = "-log10(P) Allelic series")
```

**Dopamine Beta Hydrozylase (DBH) on systolic blood pressure**

```{r}
#| label: DBH-SBP 
#| fig.width: 12
#| fig.height: 4

plot_wald_vep(gene = "DBH", outcome = "SBP")
```

## Gene based analysis

In gene based analysis, variant aggregate approaches are used to test for the effect of carrying any single or combination of rare genetic variants in a gene on a phenotype. Variant masks are often generated based on variant properties (e.g putative loss of function or missense variants), allowing the consequence of varying degrees of gene perturbation to be assessed. Here gene-based masks have been testing for cis and trans associations with plasma protein levels and complex traits, allowing them to be used in an MR framework. It is perhaps most straight forward to handle gene-level associations as single instuments and interpret Wald ratios, although in some case multiple gene-level instruments can exist for a given exposure.

```{r}
#| label: extract-masks

# Extract gene-based burden mask and calculate Wald ratios
res_masks <- lapply(harmonised_snps, function(x){
  masks <- lapply(x, function(y){
    dat <- y[IV_sets[1:4]] # gene mask results
    # Remove empty results dfs
    dat <- dat[unlist(lapply(dat, function(x){all(!(is.na(x$SNP)))}))]
    
    dat <- do.call("rbind", lapply(dat, function(z){
      z |> dplyr::select(
        mask = test.exposure, 
        mask_gene = SNP, 
        chr = chr.outcome, 
        pos = pos.outcome, 
        exposure, outcome, 
        beta.exposure, beta.outcome, 
        se.exposure, se.outcome, 
        pval.exposure, pval.outcome, 
        cis_trans)
    }))
  })
})

res_masks <- lapply(res_masks, function(x){
  out <- do.call("rbind",x)
  rownames(out) <- NULL
  return(out)})

# Wald ratios
res_masks <- lapply(res_masks, function(x){
  wrs <- apply(x, MARGIN = 1, FUN = function(x){
    TwoSampleMR::mr_wald_ratio(
    b_exp = as.numeric(x["beta.exposure"]),
    b_out = as.numeric(x["beta.outcome"]),
    se_exp = as.numeric(x["se.exposure"]),
    se_out = as.numeric(x["se.outcome"]))
  })

  wrs_df <- do.call("rbind",lapply(wrs, unlist))
  colnames(wrs_df) <- paste("wald_ratio", colnames(wrs_df), sep = "_")

  # Bind to harmonised data
  out <- cbind(x,wrs_df)
  return(out)
})
```

**Gene-level cis-mask associations**

Focusing on aggregate tests combining putative loss of function (protein truncating) and damaging rare varaints, which genes show significant causal effects on outcomes of interest:

```{r}
#| label: fig-maskcis_IVs
#| fig.width: 22
#| fig.height: 8
  
res_masks_cis <- lapply(res_masks, function(x){
  out <- x |> 
    dplyr::filter(cis_trans == "cis",
                            mask == "ptvraredmg") |> 
    dplyr::mutate(wald_ratio_pval_adj = p.adjust(wald_ratio_pval, method = "BH")) |>
    dplyr::arrange(wald_ratio_pval_adj)
  return(out)
})

# Plot Wald ratios for all gene maks associated with an outcome
plot_wald_masks <- function(results_masks, outcome, pthresh){
  dat <- results_masks[[outcome]] |> 
    dplyr::filter(wald_ratio_pval_adj < pthresh) |>
    dplyr::arrange(wald_ratio_b) |>
    dplyr::mutate(
      mask_gene = toupper(mask_gene),
      mask_gene = factor(mask_gene, levels = mask_gene))

  p_wald <- ggplot(dat, aes(x = wald_ratio_b, y = mask_gene)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(colour = "steelblue4", size = 2) +
    geom_errorbarh(aes(xmin = wald_ratio_b-1.96*wald_ratio_se, xmax = wald_ratio_b+1.96*wald_ratio_se),
                   colour = "steelblue4",
                   height = 0.4) +
    labs(x = "Wald Ratio (By/Bx)", y = "Burden test gene \n (instrumenting cis-protein expression)", 
         title = paste("Outcome:", outcome),
         subtitle = "cis-IVs (pLOF and rare damaging)") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
  
  return(p_wald)}

gridExtra::grid.arrange(grobs = lapply(names(res_masks_cis)[-1], plot_wald_masks, results_masks = res_masks_cis, pthresh = 0.1),
                        nrow = 2)
```

**Cis vs trans gene-level mask associations**

```{r}

res_masks_trans <- lapply(res_masks, function(x){
  out <- x |> 
    dplyr::filter(cis_trans == "trans",
                            mask == "ptvraredmg") |> 
    dplyr::mutate(wald_ratio_pval_adj = p.adjust(wald_ratio_pval, method = "BH")) |>
    dplyr::arrange(wald_ratio_pval_adj)
  return(out)
})

# Extract trans IVs for gene showing significant cis-instrumented causal effects
comp_masks_cistrans <- mapply(x = res_masks_cis, y = res_masks_trans, function(x,y){
  x <- x[,c("mask", "mask_gene", "exposure", "outcome", "wald_ratio_b", "wald_ratio_se", "wald_ratio_pval", "wald_ratio_pval_adj")]
  y <- y[,c("mask", "mask_gene", "exposure", "outcome", "wald_ratio_b", "wald_ratio_se", "wald_ratio_pval", "wald_ratio_pval_adj")]
  join_res <- x |> left_join(y, by = c("exposure", "outcome", "mask"), suffix = c("_cis","_trans"))
  return(join_res)
}, SIMPLIFY = F)

# Plot comparison of wald ratios for significant cis or trans IV associations
# lapply(comp_masks_cistrans, function(x){
#   dat <- x |>
#     dplyr::filter(wald_ratio_pval_adj_cis < 0.1 | wald_ratio_pval_adj_trans < 0.1) |>
#     na.exclude()
# 
#   dat |> ggplot(aes(x = wald_ratio_b_cis, y = wald_ratio_b_trans)) +
#     geom_point()
# })
```

### Rare copy number variants (CNVs)

Rare copy number variants (CNVs), including gene deletion (or complete ablation of function through loss of function mutations), and duplication (or higher order) events have been identified. As anticipated, these have been shown to have a monotonic effect on protein expression levels; that is, gene deletions reduce measured gene expression, and duplication events increase gene expression, often substantially [(Milind and Pritchard et al.)](https://pubmed.ncbi.nlm.nih.gov/40832387/). Rare genetic variation with disruptive effects on protein expression, in both directional extremes, present opportunities for protein instrumentation for molecular MR. However, it is of interest whether instrumenting gene expression using deletion and duplication events returns consistent causal estimates for exposure --> outcome effects, or non-monotonicity in genetic effects exist. The latter would suggest divergent pathways through which cis-perturbatuions of a gene may act to influence a phentoype, or potentially indicate that gene expression is not an appropriate measure of the true functional consequences of such perturbations.

Wald ratios have been derived using published gene-level summary statistics (Milind et al.) for the effect of gene deletions, duplications and pLOF variants (burdan masks) on O'link plasma protein levels and a series of 56 complex traits measured in UKB individuals. Note, Wald ratios were derived for all available cis and trans gene masks and CNV events irrespective of whether the event showed a significant protein expression effect. P-value filtered results are imported below.

Data: https://zenodo.org/records/16800547

```{r}
#| label: import-cnv-results

# Relaxed exposure p-value threshold: <0.05
#load(file = file.path(res_dir, "results_molecular_waldratios_CNVs_alloutcomes_p0.05.rda"))
#res_CNVs <- CNV_relaxed
```

**Magnitude of CNV effects**

```{r}
#| label: cis-cnvs
#| fig.width: 13
#| fig.height: 3

# Merge all results for each outcome into a single dataframe and retain cis variants
# res_CNVs_cis <- lapply(lapply(res_CNVs, function(x){
#   lapply(x, function(x){
#     do.call("rbind",x)})}), function(y){
#       out <- do.call("rbind", y) |>
#         dplyr::filter(cis_trans == "cis")
#       rownames(out) <- NULL
#       return(out)})
# 
# save(res_CNVs_cis, file = file.path(res_dir, "results_molecular_waldratios_CNVs_alloutcomes_p0.05_cis.rda"))

# Check direction of effect of cis-acting duplications, deletions and pLOFs on protein expression

load(file.path(res_dir, "results_molecular_waldratios_CNVs_alloutcomes_p0.05_cis.rda")) #res_CNVs_cis

# Single representitive outcome data to plot exposure associations from w no p-value filtering
load(file.path(res_dir, "results_molecular_waldratios_CNVs_Standing-height.rda")) #wald_ratios
dat_cnvs <- wald_ratios
dat_cnvs_cis <- do.call("rbind",lapply(dat_cnvs, function(x){
  do.call("rbind", x) |> dplyr::filter(cis_trans == "cis")}))
  
cnv_class <- c("milind_plofs" = "pLoF gene burden",
               "milind_lofs" = "LoF gene burden",
               "milind_duplications" = "gene duplication",
               "milind_deletions" = "gene deletion")

cols_cnv <- c("milind_deletions" = "steelblue4",
              "milind_lofs" = "#74ADD1",
              "milind_plofs" = "#ABD9E9",
              "milind_duplications" = "orange")

# Plot variant CNV consequnce call on cis pQTL effect
p7 <- dat_cnvs_cis |> 
  dplyr::mutate(test = factor(test, 
                              levels = c("milind_deletions","milind_lofs","milind_plofs","milind_duplications"))) |>
  ggplot(aes(x = beta.exposure, 
             y = test)) +
  ggridges::geom_density_ridges(aes(fill = test), trim = TRUE) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  scale_y_discrete(labels = cnv_class) +
  scale_fill_manual(values = cols_cnv) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "cis-pQTL expression effect (beta)", y = "CNV annotation")

# pLOF v duplication (effect significant at p<0.05 only)

p8 <- dat_cnvs_cis |>
  dplyr::select(exposure, test, beta.exposure, se.exposure) |>
  tidyr::pivot_wider(
    names_from  = test,
    values_from = c(beta.exposure, se.exposure),
    names_glue  = "{test}_{.value}") |>
  ggplot(aes(x = milind_lofs_beta.exposure, y = milind_duplications_beta.exposure)) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_hline(yintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_abline(slope = -1, intercept = 0, colour ="grey30", linewidth = 0.5) +
  geom_point(size = 2, colour = "steelblue4") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(-3,3)) +
  theme_minimal() +
  labs(x = "cis-LoF burden effect \non protein expression", y = "cis-duplication burden effect \non protein expression")

p9 <- dat_cnvs_cis |>
  dplyr::select(exposure, test, beta.exposure, se.exposure) |>
  tidyr::pivot_wider(
    names_from  = test,
    values_from = c(beta.exposure, se.exposure),
    names_glue  = "{test}_{.value}") |>
  ggplot(aes(x = milind_deletions_beta.exposure, y = milind_duplications_beta.exposure)) +
  geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_hline(yintercept = 0, colour ="grey30", linewidth = 1) + 
  geom_abline(slope = -1, intercept = 0, colour ="grey30", linewidth = 0.5) +
  geom_point(size = 2, colour = "steelblue4") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(-3,3)) +
  theme_minimal() +
  labs(x = "cis-deletion burden effect \non protein expression", y = "cis-duplication burden effect \non protein expression")

gridExtra::grid.arrange(grobs = list(p7, p8, p9), nrow = 1, widths = c(0.5,0.25,0.25))

```

**Comparison of Wald Ratios**

Does instrumenting a gene using a rare loss of function mutation or a rare deletion give consistant causal estiamtes?

```{r}
#| label: compare-walds
#| fig.width: 14
#| fig.height: 52

comp_WRs_lofdups <- lapply(res_CNVs_cis, function(x){
  
  if(!(all(c("milind_lofs","milind_duplications") %in% x$test))){
    return(NULL)
    next
  }
  # Data frame of paired wald ratios for each gene
  dat <- x |> dplyr::filter(test == c("milind_lofs","milind_duplications")) |>
    dplyr::select(exposure, test, wald_ratio_b, wald_ratio_se, wald_ratio_pval) |>
    tidyr::pivot_wider(
      names_from  = test,
      values_from = c(wald_ratio_b, wald_ratio_se, wald_ratio_pval),
      names_glue  = "{test}_{.value}") |>
    as.data.frame() |>
    na.exclude()
  
  hettest <- apply(dat, MARGIN = 1, function(y){
    effect_heterogeneity(
      beta_vec = as.numeric(y[2:3]), 
      se_vec = as.numeric(y[4:5]))}) |> 
    t() |> as.data.frame() |>
    dplyr::mutate(Qpval_adj = p.adjust(Qpval, method = "BH"))
  
  dat <- cbind(dat, hettest) |>
    dplyr::mutate(outcome = unique(x$outcome)) |>
    dplyr::arrange(Qpval_adj)
  
  return(dat)
})
    
p10 <- lapply(comp_WRs_lofdups[sapply(comp_WRs_lofdups, is.null) == FALSE], function(x){
  outcome <- unique(x$outcome)
  
  x$col_point <- factor(ifelse(x$Qpval_adj < 0.05, "Het. Q pval <0.05 (FDR)",
                      ifelse((p.adjust(x$milind_lofs_wald_ratio_pval) < 0.05 | p.adjust(x$milind_duplications_wald_ratio_pval) < 0.05), "WR pval <0.05 (FDR)",
                      "NS")), levels = c("Het. Q pval <0.05 (FDR)", "WR pval <0.05 (FDR)", "NS"))
  
  x |> ggplot(aes(x = milind_lofs_wald_ratio_b, y = milind_duplications_wald_ratio_b)) +
    geom_abline(slope = 1, intercept = 0, colour ="grey30", linewidth = 0.5) +
    geom_vline(xintercept = 0, colour ="grey30", linewidth = 1) + 
    geom_hline(yintercept = 0, colour ="grey30", linewidth = 1) + 
    geom_point(aes(colour = col_point, size = col_point)) +
    scale_colour_manual(values = c("tomato","black","grey"), drop = FALSE) +
    scale_size_manual(values = c(3,3,2), drop = FALSE) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "none") +
    labs(x = "Wald Ratio \n(cis-LoF instrumantaion)", 
         y = "Wald Ratio \n(cis-duplication instrumentation)",
         title = paste("Protein ->\n", outcome))
})

gridExtra::grid.arrange(grobs = p10, nrow = 14)

## Pull out instances of effect heterogeneity between LoF and duplication instruments
het_cnv_wrs <- do.call("rbind", lapply(comp_WRs_lofdups[sapply(comp_WRs_lofdups, is.null) == FALSE], function(x){x |> dplyr::filter(Qpval_adj < 0.05)}))
```

**Heterogeneity in Wald Ratios**

```{r}
#| label: compare-walds-hets
#| fig.width: 15
#| fig.height: 20
#| 
# Plot wald ratios of differing cnv instrumentation strategies
plot_wald_cnvs <- function(gene, outcome){
  
  dat <- res_CNVs_cis[[outcome]] |> 
    dplyr::filter(exposure == gene)
  
  p_theme <- theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
  
  labs <- c("milind_lofs" = "cis-LoF", "milind_duplications" = "cis-Dup.", "milind_deletions" = "cis-Del.")
  
  exp_range <- max(abs(dat$beta.exposure) + 1.96*dat$se.exposure) + 0.1
  out_range <- max(abs(dat$beta.outcome) + 1.96*dat$se.outcome) + 0.1
  wr_range <- max(abs(dat$wald_ratio_b) + 1.96*dat$wald_ratio_se) + 0.1
  
  # Exposure
  p_exp <- ggplot(dat, aes(x = beta.exposure, y = labs[test], colour = test)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    scale_colour_manual(values = cols_cnv, labels = cnv_class, name = "") +
    scale_x_continuous(limits = c(-exp_range, exp_range)) +
    geom_errorbarh(aes(xmin = beta.exposure-1.96*se.exposure, xmax = beta.exposure+1.96*se.exposure), height = 0.4) +
    labs(x = "Bx", y = "") +
    p_theme +
    theme(legend.position = "none")
  
  # Outcome
  p_out <- ggplot(dat, aes(x = beta.outcome, y = labs[test], colour = test)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    scale_colour_manual(values = cols_cnv, labels = cnv_class, name = "") +
    scale_x_continuous(limits = c(-out_range, out_range)) +
    geom_errorbarh(aes(xmin = beta.outcome-1.96*se.outcome, xmax = beta.outcome+1.96*se.outcome), height = 0.4) +
    labs(x = "By", y = "") +
    p_theme +
    theme(legend.position = "none")
  
  # Wald Ratio
  p_wald <- ggplot(dat, aes(x = wald_ratio_b, y = labs[test], colour = test)) +
    geom_vline(xintercept = 0, colour = "grey20", lty = 2) +
    geom_point(size = 3) +
    scale_colour_manual(values = cols_cnv, labels = cnv_class, name = "") +
    scale_x_continuous(limits = c(-wr_range, wr_range)) +
    geom_errorbarh(aes(xmin = wald_ratio_b-1.96*wald_ratio_se, xmax = wald_ratio_b+1.96*wald_ratio_se), height = 0.4) +
    labs(x = "Wald Ratio (By/Bx)", y = "", 
         title = paste("Exposure:", gene, "\nOutcome:", outcome)) +
    p_theme +
    theme(legend.position = "none", plot.title = element_text(size = 8))

  return(gridExtra::arrangeGrob(grobs = list(p_wald, p_exp, p_out), nrow = 3, heights = c(0.38,0.31,0.31)))
}

gridExtra::grid.arrange(grobs = mapply(plot_wald_cnvs, gene = het_cnv_wrs$exposure, outcome = het_cnv_wrs$outcome, SIMPLIFY = F), nrow = 5)
```
